---
title: "Hands-on Exercise 4"
subtitle: "Spatial Weights and Applications"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "September 17, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

::: {.pagefacts}
[88 counties of Hunan, one year of indicators]{.fact}
[Hunan shapefile, Hunan_2012.csv]{.fact .data}
[`sf` `spdep` `tmap` `tidyverse` `knitr` `ggrepel`]{.fact .pkg}
:::

# Overview

Exercise 1 was about choropleth maps and the last two were about points. This one is about areas, and about the one idea everything after it depends on. Before you can ask whether rich counties sit next to rich counties, you have to decide what "next to" means. That decision is the spatial weights matrix.

The chapter builds several of them for the 88 counties of Hunan and then uses them to compute a spatially lagged GDP per capita. Nothing here is a test yet. Those come in the next chapter.

# Loading the Packages

Only one package is new for the chapter. **spdep** builds neighbour lists and weights and does the lagging. **sf** reads the county boundaries and handles geometry, **tmap** draws the maps, **tidyverse** reads the csv, joins the tables and draws the scatter plot with **ggplot2**, and **knitr** prints tables with `kable()`. I added **ggrepel** myself so the point labels on the scatter plot do not sit on top of the points or the fitted line.

```{r}
#| output: false
pacman::p_load(sf, spdep, tmap, tidyverse, knitr, ggrepel)
```

# Preparing the Data

Three steps, the boundaries, the indicators, and the join that brings them together.

::: panel-tabset
## Importing the Hunan shapefile

`st_read()` of **sf** reads the Hunan county boundaries from the shapefile.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

Eighty-eight polygons in WGS 84. The `County` column is the join key for the next step.

## Importing the Hunan_2012 table

`read_csv()` of **readr** reads the 2012 indicators, and `glimpse()` of **dplyr** shows every column with its type.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
glimpse(hunan2012)
```

## Performing the relational join

`left_join()` of **dplyr** attaches the indicators to the polygons, and `select()` keeps the name columns and GDPPC.

```{r}
hunan <- left_join(hunan, hunan2012) %>%
  select(NAME_2, ID_3, NAME_3, ENGTYPE_3, County, GDPPC)

nrow(hunan)
```

:::

::: {.callout-note}
## Learning diary
The zip came with a `Dictionary.xlsx` the chapter never mentions. It says GDPPC is GDP per capita in RMB for 2005 to 2012, so the 2012 in the file name is the year, not a version. It also flags that three counties, Shuangqing, Beita and Daxiang, share the same disposable income value. That is not a column this exercise uses, but it is the kind of thing worth knowing before the take-home.
:::

::: {.callout-caution}
## Where I differ from the book
The book does `select(1:4, 7, 15)`. That works, but only because column 15 happens to be GDPPC after the join. Add one column to the csv and it silently picks the wrong variable. I select by name instead, which is the same result and cannot drift.

`left_join()` found the key on its own because both tables have a `County` column. I checked that all 88 names match on both sides before trusting that. They do.
:::

# Visualising Regional Development Indicator

`tm_shape()` and `tm_polygons()` of **tmap** draw a reference map with the county names next to a quantile map of GDP per capita, and `tmap_arrange()` puts the two side by side. `remove_overlap = TRUE` drops any name that would collide with another, so a few of the small counties packed into the middle of the province lose their label.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
basemap <- tm_shape(hunan) +
  tm_polygons(fill = "#F7F4EC", col = "grey60") +
  tm_text("NAME_3", size = 0.6,
          options = opt_tm_text(remove_overlap = TRUE)) +
  tm_title("Hunan, 88 counties", size = 1.3, fontface = "bold")

gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.oranges"),
              fill.legend = tm_legend(title = "GDP per capita (RMB)",
                                      position = tm_pos_out("right", "center")),
              col = "grey60") +
  tm_title("GDP per capita, 2012", size = 1.3, fontface = "bold")

tmap_arrange(basemap, gdppc, ncol = 2)
```

## The Code

```{r}
#| eval: false
basemap <- tm_shape(hunan) +
  tm_polygons(fill = "#F7F4EC", col = "grey60") +
  tm_text("NAME_3", size = 0.6,
          options = opt_tm_text(remove_overlap = TRUE)) +
  tm_title("Hunan, 88 counties", size = 1.3, fontface = "bold")

gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.oranges"),
              fill.legend = tm_legend(title = "GDP per capita (RMB)",
                                      position = tm_pos_out("right", "center")),
              col = "grey60") +
  tm_title("GDP per capita, 2012", size = 1.3, fontface = "bold")

tmap_arrange(basemap, gdppc, ncol = 2)
```
:::

The high values sit in the middle and the north east, around Changsha. The west and south are poorer. Whether that is a real pattern or the eye finding one is exactly what the next chapter tests, and what this chapter builds the tools for.

# Computing Contiguity Spatial Weights

Two polygons are contiguous if they touch. Queen contiguity counts a shared corner as touching. Rook contiguity needs a shared edge.

## Computing Queen and Rook contiguity based neighbours

::: panel-tabset
## Queen contiguity

`poly2nb()` of **spdep** builds the neighbour list from shared boundaries, and `queen = TRUE` lets a single shared corner count. `summary()` prints the link counts.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

Every county has at least one neighbour, so there are no islands to worry about. The busiest county has eleven.

The summary names region 85 as the busiest, so I look up its name, the row numbers of its neighbours, and their names.

```{r}
hunan$County[85]
wm_q[[85]]
hunan$County[wm_q[[85]]]
```

::: {.callout-note}
## Learning diary
`wm_q` is a list of 88 integer vectors. Element 85 holds the row numbers of `r hunan$County[85]`'s neighbours, and looking them up in `hunan$County` turns numbers into names. That lookup is the whole trick to reading any neighbour object. The numbers are positions, not IDs, so if the rows of `hunan` were ever reordered every weight matrix would have to be rebuilt.
:::

`nb2mat()` of **spdep** turns the list into a full matrix, with `style = "B"` giving a 1 for each neighbour and 0 otherwise. It is 88 by 88, so I only show a corner of it.

```{r}
wm_q_mat <- nb2mat(wm_q, style = "B")
wm_q_mat[1:8, 1:8]
```

## Rook contiguity

The same `poly2nb()` call with `queen = FALSE` only counts counties that share an edge.

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

Same 88 regions, 440 links instead of 448. Every link is counted from both ends, so that is four county pairs that touch only at a corner.

:::

## Visualising contiguity weights

The links are drawn between polygon centroids, so I need those first. `map_dbl()` of **purrr** runs `st_centroid()` of **sf** on each polygon and keeps the longitude, then the latitude, and `cbind()` puts them into a two column matrix.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
coords <- cbind(longitude, latitude)
head(coords)
```

`plot()` draws the county outlines first, then the `nb` method of **spdep** draws a line between every pair of neighbouring centroids on top with `add = TRUE`.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0), cex.main = 1.3)
plot(hunan$geometry, border = "grey70", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.5, add = TRUE, col = "#D55E00")
plot(hunan$geometry, border = "grey70", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.5, add = TRUE, col = "#0072B2")
par(op)
```

## The Code

```{r}
#| eval: false
op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0), cex.main = 1.3)
plot(hunan$geometry, border = "grey70", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.5, add = TRUE, col = "#D55E00")
plot(hunan$geometry, border = "grey70", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.5, add = TRUE, col = "#0072B2")
par(op)
```
:::

The two look identical at this scale. The four missing pairs are corner touches, and a corner touch is a single point, so there is nothing to see.

# Computing Distance Based Neighbours

Contiguity does not care how big a county is. A distance rule does. The book uses two, a fixed radius and a fixed number of nearest neighbours.

## Determine the cut-off distance

The radius has to be big enough that nobody is left without a neighbour. The book finds it by asking how far each county's single nearest neighbour is, and taking the largest of those. `knearneigh()` of **spdep** finds each county's nearest centroid, `knn2nb()` turns that into a neighbour list, and `nbdists()` returns the great circle distances in km because of `longlat = TRUE`.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The largest nearest-neighbour distance is just under 62 km, so 62 km guarantees every county at least one neighbour. `which.max()` tells me which county that longest distance belongs to.

```{r}
which.max(k1dists)
hunan$County[which.max(k1dists)]
```

::: {.callout-important}
## Check that caught something
One county sets the cut-off for all 88. I wanted to know which. It is `r hunan$County[which.max(k1dists)]`, the same county that had the most neighbours under Queen. That is not a contradiction. It is one of the largest counties on the map, fourth by area after Yuanling, Liuyang and Anhua, and it has a long irregular shape, so it touches many others while its centroid sits far from all of theirs. Contiguity says it is the best connected county in Hunan. Distance says it is the most isolated. Everything about the fixed distance matrix below is shaped by that one row.
:::

## Computing fixed and adaptive distance weight matrices

::: panel-tabset
## Fixed distance weight matrix

`dnearneigh()` of **spdep** links every pair of centroids between 0 and 62 km apart, measured along the great circle.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

`n.comp.nb()` of **spdep** counts the connected components of the graph, which tells me whether any group of counties is cut off from the rest.

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

One connected component, so the graph does not fall apart into islands at this radius.

## Adaptive distance weight matrix

`knearneigh()` with `k = 6` fixes the number of neighbours instead of the distance, and `knn2nb()` turns the result into a neighbour list.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

Every county gets exactly six, whether it is a small county packed around Changsha or a large one in the mountains.

:::

## Plotting distance based neighbours

The same outline plus `nb` plot as before, now for the 62 km list and the six nearest list.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0), cex.main = 1.3)
plot(hunan$geometry, border = "grey70", main = "Fixed distance, 62 km")
plot(wm_d62, coords, pch = 19, cex = 0.5, add = TRUE, col = "#D55E00")
plot(hunan$geometry, border = "grey70", main = "Adaptive, k = 6")
plot(knn6, coords, pch = 19, cex = 0.5, add = TRUE, col = "#0072B2")
par(op)
```

## The Code

```{r}
#| eval: false
op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0), cex.main = 1.3)
plot(hunan$geometry, border = "grey70", main = "Fixed distance, 62 km")
plot(wm_d62, coords, pch = 19, cex = 0.5, add = TRUE, col = "#D55E00")
plot(hunan$geometry, border = "grey70", main = "Adaptive, k = 6")
plot(knn6, coords, pch = 19, cex = 0.5, add = TRUE, col = "#0072B2")
par(op)
```
:::

::: {.callout-note}
## Learning diary
The fixed radius is dense where counties are small and sparse where they are large, because 62 km reaches many small centroids and few large ones. The adaptive rule is the opposite. Everyone has six, so the links are long in the mountains and short around the capital.

Which is right depends on what the weights are for. If the question is about physical spillover, distance matters and the fixed radius is honest about it. If the question is about comparing each county to its peers, the adaptive rule keeps the comparison fair.
:::

# Weights Based on IDW

A neighbour list says who counts. A weight says how much. Two schemes here, inverse distance and row standardised.

`nbdists()` returns the great circle distance from each county to each of its Queen neighbours, and `lapply()` turns every distance into its inverse.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1 / x)
ids[[1]]
```

Closer neighbours weigh more. The unit is one over km, which is why the numbers are small.

# Row-standardised Weights Matrix

`nb2listw()` of **spdep** turns a neighbour list into a weights list. With `style = "W"` each county's weights are divided by its number of neighbours so they sum to one.

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q$weights[85]
```

`r hunan$County[85]` has `r length(wm_q[[85]])` neighbours, so each gets a weight of 1/`r length(wm_q[[85]])`. The row sums to one.

The book then passes the inverse distances to `nb2listw()` through `glist`, with `style = "B"`.

```{r}
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids$weights[1]
```

::: {.callout-note}
## Learning diary
`style = "W"` divides each neighbour's weight by the number of neighbours, so a lag becomes an average. `style = "B"` leaves the weights as given, so a lag becomes a sum. The book calls the inverse distance version "row standardised" but passes `style = "B"`, so it is not. Its lags will be sums of inverse distances, not averages. That is fine, it just is not what the book's heading says.

`zero.policy = TRUE` tells spdep to carry on if a region has no neighbours. Nobody in Hunan is in that position, so it does nothing here, but leaving it on costs nothing and stops a hard error on a dataset that does have islands.
:::

# Application of Spatial Weight Matrix

A spatial lag replaces each county's value with something computed from its neighbours' values. Four versions, and the difference between them is only which weights go in.

## Spatial lag with row-standardised weights

`lag.listw()` of **spdep** multiplies the weights by GDPPC. With the style W weights that gives the average GDPPC of each county's neighbours.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
head(GDPPC.lag)
```

To map it, I pair the lag with the county names in a data frame and `left_join()` it back onto `hunan` by `NAME_3`.

```{r}
lag.list <- list(hunan$NAME_3, GDPPC.lag)
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res)
```

## Spatial lag as a sum of neighbouring values

Giving every neighbour a weight of one and using `style = "B"` turns the same `lag.listw()` call into a sum of the neighbours' GDPPC, which is joined back the same way.

```{r}
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
hunan <- left_join(hunan, lag.res)
```

::: {.callout-note}
## Learning diary
`0 * x + 1` is a small trick. Multiplying the neighbour list by zero and adding one gives a vector of ones the same length as each county's neighbour list. With `style = "B"` those ones are used as is, so the lag is a plain sum.
:::

## Spatial window average

`include.self()` of **spdep** adds each county to its own neighbour list, so the row-standardised lag becomes the average of the county and its neighbours. `nb2listw()` uses style W by default.

```{r}
wm_qs <- include.self(wm_q)
wm_qs <- nb2listw(wm_qs)
lag_w_avg_gpdpc <- lag.listw(wm_qs, hunan$GDPPC)
lag_wm_qs.res <- as.data.frame(list(hunan$NAME_3, lag_w_avg_gpdpc))
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
hunan <- left_join(hunan, lag_wm_qs.res)
```

## Spatial window sum

The window sum is the lag sum with each county included in its own list, so it is the county's GDPPC plus all of its neighbours'.

```{r}
wm_qs <- include.self(wm_q)
b_weights <- lapply(wm_qs, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_qs, glist = b_weights, style = "B")
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
hunan <- left_join(hunan, w_sum_gdppc.res)
```

`kable()` of **knitr** lines up the original value and the four lags for the first eight counties.

```{r}
hunan %>%
  st_drop_geometry() %>%
  select(NAME_3, GDPPC, `lag GDPPC`, `lag_window_avg GDPPC`,
         `lag_sum GDPPC`, `w_sum GDPPC`) %>%
  head(8) %>%
  kable(digits = 0)
```

## Visualising the spatial lags

A small helper function builds one quantile map per variable, and `tmap_arrange()` puts GDPPC and three of the lags in a 2x2 grid. Each panel has its own quantile breaks, so the colours compare ranks within a map, not RMB across maps.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
lag_map <- function(var, title, legend_title) {
  tm_shape(hunan) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.oranges"),
                fill.legend = tm_legend(title = legend_title,
                                        position = tm_pos_out("right", "center")),
                col = "grey60") +
    tm_title(title, size = 1.3, fontface = "bold")
}

tmap_arrange(lag_map("GDPPC", "GDPPC",
                     "GDP per capita (RMB)"),
             lag_map("lag GDPPC", "Lag, neighbour average",
                     "Neighbour average (RMB)"),
             lag_map("lag_window_avg GDPPC", "Window average",
                     "Window average (RMB)"),
             lag_map("lag_sum GDPPC", "Lag sum",
                     "Sum of neighbours (RMB)"),
             ncol = 2)
```

## The Code

```{r}
#| eval: false
lag_map <- function(var, title, legend_title) {
  tm_shape(hunan) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.oranges"),
                fill.legend = tm_legend(title = legend_title,
                                        position = tm_pos_out("right", "center")),
                col = "grey60") +
    tm_title(title, size = 1.3, fontface = "bold")
}

tmap_arrange(lag_map("GDPPC", "GDPPC",
                     "GDP per capita (RMB)"),
             lag_map("lag GDPPC", "Lag, neighbour average",
                     "Neighbour average (RMB)"),
             lag_map("lag_window_avg GDPPC", "Window average",
                     "Window average (RMB)"),
             lag_map("lag_sum GDPPC", "Lag sum",
                     "Sum of neighbours (RMB)"),
             ncol = 2)
```
:::

::: {.callout-note}
## Learning diary
The two averages smooth the map. The north east block around Changsha firms up on both, spreading over most of the counties around the capital. The south east block around Chenzhou holds on the window average, but on the neighbour average the darkest class moves out to the ring of counties around it, because each county's own value is left out and Zixing's neighbours are poorer than Zixing. The two lone highs in the west, Jishou and Zhongfang, and the small high spot of Lengshuijiang in the middle all fade, because their neighbours are poor.

The sum does something different. Look at the big county in the north. It sits in the middle class on the GDPPC map and on the neighbour average, but it is in the darkest class on the sum, and that is Taoyuan with its eleven neighbours. The sum rewards having many neighbours regardless of how rich they are, so a lag sum map is partly a map of neighbour counts. It carries the shape of the weights matrix as much as the values.
:::

# Beyond the Exercise

Three things I wanted to check that the chapter does not.

## Four rules, one table

The chapter builds four neighbour definitions and never lines them up. `card()` of **spdep** returns the number of neighbours of each county, and `tibble()` collects the totals, minimum, median and maximum for each rule.

```{r}
nb_summary <- tibble(
  rule = c("Queen", "Rook", "Fixed 62 km", "k = 6"),
  links = c(sum(card(wm_q)), sum(card(wm_r)),
            sum(card(wm_d62)), sum(card(knn6))),
  min_nb = c(min(card(wm_q)), min(card(wm_r)),
             min(card(wm_d62)), min(card(knn6))),
  median_nb = c(median(card(wm_q)), median(card(wm_r)),
                median(card(wm_d62)), median(card(knn6))),
  max_nb = c(max(card(wm_q)), max(card(wm_r)),
             max(card(wm_d62)), max(card(knn6)))
)
kable(nb_summary)
```

To see where each rule is generous, I add the Queen and 62 km neighbour counts to a copy of `hunan` and map them side by side with the same five classes, so the two panels share one scale.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
hunan_nb <- hunan %>%
  mutate(queen_nb = card(wm_q),
         d62_nb = card(wm_d62))

nb_map <- function(var, title) {
  tm_shape(hunan_nb) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(breaks = c(1, 3, 5, 7, 9, 12),
                                                labels = c("1 to 2", "3 to 4", "5 to 6",
                                                           "7 to 8", "9 to 11"),
                                                values = c("#BCBDDC", "#9E9AC8", "#807DBA",
                                                           "#6A51A3", "#4A1486")),
                fill.legend = tm_legend(title = "Number of neighbours",
                                        position = tm_pos_out("right", "center")),
                col = "grey60") +
    tm_title(title, size = 1.3, fontface = "bold")
}

tmap_arrange(nb_map("queen_nb", "Neighbours under Queen"),
             nb_map("d62_nb", "Neighbours within 62 km"),
             ncol = 2)
```

## The Code

```{r}
#| eval: false
hunan_nb <- hunan %>%
  mutate(queen_nb = card(wm_q),
         d62_nb = card(wm_d62))

nb_map <- function(var, title) {
  tm_shape(hunan_nb) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(breaks = c(1, 3, 5, 7, 9, 12),
                                                labels = c("1 to 2", "3 to 4", "5 to 6",
                                                           "7 to 8", "9 to 11"),
                                                values = c("#BCBDDC", "#9E9AC8", "#807DBA",
                                                           "#6A51A3", "#4A1486")),
                fill.legend = tm_legend(title = "Number of neighbours",
                                        position = tm_pos_out("right", "center")),
                col = "grey60") +
    tm_title(title, size = 1.3, fontface = "bold")
}

tmap_arrange(nb_map("queen_nb", "Neighbours under Queen"),
             nb_map("d62_nb", "Neighbours within 62 km"),
             ncol = 2)
```
:::

::: {.callout-tip}
## Alternative explored
The table surprised me. I expected the 62 km rule to be the densest, because a radius reaches past the counties you touch. It is the sparsest, 324 links against Queen's 448, and nobody gets more than six. The reason is scale. Hunan's counties are mostly 40 to 60 km across, so a 62 km circle around a centroid rarely reaches past the immediate ring, and for the big counties in the west it does not even reach that.

The map shows where each rule is generous. Queen hands the most neighbours to the largest counties, because a big polygon touches many small ones. The 62 km rule hands the most to the middle band of small counties, where centroids are packed tightly, and leaves the north west and the far south with one or two. Same province, opposite answers about who is well connected.

That matters for the next chapter. Whichever rule I pick decides which counties carry weight in every statistic after it. Queen leans on the big western counties. Fixed distance leans on the small central ones.
:::

## Does the projection matter

The book keeps the data in WGS 84 and asks `nbdists()` for great circle distances with `longlat = TRUE`. That is correct for distances. But `st_centroid()` on a lon/lat polygon is not quite right, and sf says so in a warning that `warning: false` hides. I wanted to know how far off it is. `st_transform()` of **sf** projects the counties to UTM zone 49N, `st_centroid()` then works in metres, and the same nearest neighbour steps run on the projected centroids.

```{r}
hunan_utm <- st_transform(hunan, crs = 32649)
coords_utm <- st_coordinates(st_centroid(hunan_utm))

k1_utm <- knn2nb(knearneigh(coords_utm))
k1dists_utm <- unlist(nbdists(k1_utm, coords_utm)) / 1000
summary(k1dists_utm)
```

A small table puts the two cut-offs and the county that sets each one next to each other.

```{r}
tibble(
  method = c("WGS 84 centroids, great circle", "UTM 49N centroids, planar"),
  max_nn_km = c(max(k1dists), max(k1dists_utm)),
  cutoff_county = c(hunan$County[which.max(k1dists)],
                    hunan$County[which.max(k1dists_utm)])
) %>% kable(digits = 2)
```

::: {.callout-tip}
## Alternative explored
Hunan sits at about 112 east and 28 north, inside UTM zone 49 north, which is EPSG 32649. I expected the two methods to agree within a few hundred metres. They do not. Projecting first moves the cut-off from 61.8 km to 58.3 km, and a different county sets it, Yuanling instead of Taoyuan.

The distance formula is not the cause. Great circle and planar UTM agree to well under a kilometre over 60 km. The cause is the centroids. `st_centroid()` on lon/lat treats degrees as if they were flat, and for a big irregular county that puts the centroid a few kilometres from where it belongs. Taoyuan is big and irregular, so its centroid moved the most, and its nearest neighbour got closer.

The book's 62 km radius still works, because 62 is above both numbers. But the warning it hides is not about tiny errors. It is about which county is the loneliest, and that changed. I am projecting first from now on.
:::

## A preview of Moran's I

The next chapter asks whether GDPPC is spatially autocorrelated. The lag I just computed is half of that answer already. If a county's value predicts its neighbours' average, the points on this plot line up. `lm()` fits the line so its slope can go in the subtitle, and `geom_text_repel()` of **ggrepel** labels the richest counties and the ones with the richest neighbours without covering the points or the line.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
lag_df <- hunan %>%
  st_drop_geometry() %>%
  select(NAME_3, GDPPC, lag = `lag GDPPC`)

fit <- lm(lag ~ GDPPC, data = lag_df)

ggplot(lag_df, aes(GDPPC, lag)) +
  scale_x_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0.05, 0.12))) +
  scale_y_continuous(labels = scales::comma) +
  geom_hline(yintercept = mean(lag_df$lag), colour = "grey70", linetype = 2) +
  geom_vline(xintercept = mean(lag_df$GDPPC), colour = "grey70", linetype = 2) +
  geom_point(colour = "#D55E00", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#0072B2", linewidth = 0.7) +
  geom_text_repel(data = lag_df %>% filter(GDPPC > 60000 | lag > 45000),
                  aes(label = NAME_3), size = 3.5, seed = 42,
                  box.padding = 0.5, min.segment.length = 0) +
  labs(x = "GDPPC (RMB)", y = "Neighbour average GDPPC (RMB)",
       title = "Each county against the average of its Queen neighbours",
       subtitle = paste0("Slope of the fitted line, ",
                         round(coef(fit)[2], 3))) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13))
```

## The Code

```{r}
#| eval: false
lag_df <- hunan %>%
  st_drop_geometry() %>%
  select(NAME_3, GDPPC, lag = `lag GDPPC`)

fit <- lm(lag ~ GDPPC, data = lag_df)

ggplot(lag_df, aes(GDPPC, lag)) +
  scale_x_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0.05, 0.12))) +
  scale_y_continuous(labels = scales::comma) +
  geom_hline(yintercept = mean(lag_df$lag), colour = "grey70", linetype = 2) +
  geom_vline(xintercept = mean(lag_df$GDPPC), colour = "grey70", linetype = 2) +
  geom_point(colour = "#D55E00", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#0072B2", linewidth = 0.7) +
  geom_text_repel(data = lag_df %>% filter(GDPPC > 60000 | lag > 45000),
                  aes(label = NAME_3), size = 3.5, seed = 42,
                  box.padding = 0.5, min.segment.length = 0) +
  labs(x = "GDPPC (RMB)", y = "Neighbour average GDPPC (RMB)",
       title = "Each county against the average of its Queen neighbours",
       subtitle = paste0("Slope of the fitted line, ",
                         round(coef(fit)[2], 3))) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13))
```
:::

::: {.callout-tip}
## Alternative explored
The slope of that line is Moran's I itself, 0.30 here. I first thought both variables had to be standardised for that to hold, but they do not. With row-standardised weights each row sums to one, so the neighbour average of a centred variable is just the neighbour average minus the overall mean, and the least squares slope of the lag on GDPPC works out to exactly the Moran's I formula. Rescaling both axes would not change it. It is positive, so rich counties do tend to have rich neighbours, and `moran.test()` in the next chapter should return the same 0.30 with a p-value attached.

The labelled points are the ones off the line. Changsha, Wangcheng, Liuyang and Ningxiang sit top right, rich with rich neighbours, which is the capital and its ring. Zixing and Lengshuijiang sit bottom right, rich counties surrounded by poor ones. Both are mining and industrial towns, so the wealth is local to the pit and does not spill over. Pingjiang is the reverse, a poor county whose neighbours average over 55,000 because it borders the Changsha ring.

Those off-line counties are what the local statistics in the next chapter pick out one by one. Whether 0.30 is more than chance is also a question for that chapter, because only the test can say.
:::

# Limitations

- The centroids used for the link plots and the distance rules were computed on lon/lat, which moves them by a few kilometres for the big counties. The projection check in Beyond the Exercise shows that is enough to change which county sets the distance cut-off.
- The 62 km cut-off is set by a single county. A different boundary file with a different centroid for that county would change the radius for everyone.
- Row standardising treats every neighbour as equal within a county. Inverse distance does not, but the book then uses it with binary style, so neither version here is a distance weighted average.
- All of this is 2012 GDPPC. One year, one variable, no sense yet of whether the pattern is stable.

# Reflections

Four things I am taking from this one.

**The weights matrix is a decision, not a fact.** Four reasonable rules gave four different graphs of the same 88 counties. Nothing downstream is more objective than the choice made here, and that choice needs a sentence of justification every time.

**Read the count, not just the map.** Queen and Rook looked identical when plotted. The summary said eight links were missing. The picture would never have told me.

**Style W and style B are the difference between an average and a sum.** Same neighbours, same values, and the maps come out telling different stories. One is about levels, the other is partly about how many neighbours you have.

**The projection shortcut is not as safe as it looks.** The book's degrees plus `longlat = TRUE` gets the distances right but the centroids wrong, and that was enough to move the cut-off by about 3.5 km and hand it to a different county. The radius survived. The reasoning under it did not. Better to project once at the top and stop thinking about it.
