---
title: "In-class Exercise 6"
subtitle: "Emerging Hot Spot Analysis"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "October 3, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

# Overview

This in-class exercise uses the 88 counties of Hunan and their GDP per capita from 2005 to 2021. Emerging hot spot analysis (EHSA) asks how hot and cold spots change over time. It has four steps.

1. Build a space-time cube, with one row per county per year.
2. Compute local Gi* for every county in every year.
3. Test each county's Gi* series for a trend with the Mann-Kendall test.
4. Classify each county, for example as a new, persistent or sporadic hot spot.

# Loading the Packages

**sfdep** builds the space-time cube and runs the hot spot analysis. **Kendall** does the Mann-Kendall test, and **plotly** makes the trend chart interactive. **sf**, **tmap** and **tidyverse** do the same jobs as before, and **knitr** prints tables.

```{r}
#| output: false
pacman::p_load(sf, sfdep, tmap, plotly, tidyverse, Kendall, knitr)
```

# The Data

::: panel-tabset
## Importing the Hunan shapefile

`st_read()` of **sf** reads the county boundaries, and `st_transform()` projects them to UTM zone 49 north.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan") %>%
  st_transform(crs = 32649)
```

## Importing the GDPPC table

`read_csv()` of **readr** reads GDP per capita for every county and year.

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
glimpse(GDPPC)
```
:::

The table has `r nrow(GDPPC)` rows, `r n_distinct(GDPPC$County)` counties over `r n_distinct(GDPPC$Year)` years from `r min(GDPPC$Year)` to `r max(GDPPC$Year)`.

::: {.callout-caution}
## Where I differ from the book
The book projects Hunan to EPSG 32650, UTM zone 50 north. Hunan runs from about 108.8 to 114.3 degrees east, and zone 50 only starts at 114. So I use EPSG 32649, as in Hands-on Exercises 4 and 5. The distances between counties change a little with the projection, so my Gi* values may differ slightly from the book's.
:::

# Creating a Time Series Cube

`spacetime()` of **sfdep** links the yearly table to the county boundaries. `.loc_col` names the column that matches the two, and `.time_col` names the time column.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

`is_spacetime_cube()` checks that every county has a row for every year.

```{r}
is_spacetime_cube(GDPPC_st)
```

::: {.callout-note}
## Learning diary
A space-time cube needs a complete grid. Each of the `r n_distinct(GDPPC$County)` counties must appear once in each of the `r n_distinct(GDPPC$Year)` years, so there should be `r n_distinct(GDPPC$County) * n_distinct(GDPPC$Year)` rows. The table has `r nrow(GDPPC)`, so the cube is complete. A missing year for one county would make this return `FALSE`.
:::

# Computing Gi*

## Deriving the Spatial Weights

`activate("geometry")` switches to the geometry part of the cube. `st_contiguity()` finds the Queen neighbours, `include_self()` adds each county to its own list, and `st_inverse_distance()` weights closer neighbours more. `set_nbs()` and `set_wts()` then copy the neighbours and weights to every year.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb,
                             geometry,
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

::: {.callout-note}
## Learning diary
Gi* counts the county itself as part of its neighbourhood, which is the star in Gi*. That is why `include_self()` is needed here, as in In-class Exercise 5.
:::

## Computing the Local Gi*

`group_by(Year)` makes `local_gstar_perm()` run once per year. `unnest()` spreads the results into columns.

```{r}
set.seed(1234)
gi_stars <- GDPPC_nb %>%
  group_by(Year) %>%
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>%
  tidyr::unnest(gi_star)
```

::: {.callout-caution}
## Where I differ from the book
`local_gstar_perm()` uses random permutations for its p-values, so I added `set.seed()`. The Gi* values do not depend on the seed, but the simulated p-values do.
:::

# Mann-Kendall Test

The Mann-Kendall test checks whether a series has a steady upward or downward trend. It does not assume the trend is a straight line. The null hypothesis is no trend.

## Mann-Kendall Test on Gi* for Changsha

`filter()` keeps the Gi* series for Changsha, the provincial capital.

```{r}
cbg <- gi_stars %>%
  ungroup() %>%
  filter(County == "Changsha") |>
  select(County, Year, gi_star)
```

## Interactive Mann-Kendall Plot

`ggplot()` draws the series and `ggplotly()` of **plotly** makes it interactive. Hovering shows the value for each year.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
p <- ggplot(data = cbg,
            aes(x = Year, y = gi_star)) +
  geom_line(colour = "#0072B2", linewidth = 1) +
  geom_point(colour = "#0072B2", size = 2) +
  labs(x = "Year", y = "Gi*",
       title = "Changsha's Gi* by year, 2005 to 2021") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"))

ggplotly(p)
```

## The Code

```{r}
#| eval: false
p <- ggplot(data = cbg,
            aes(x = Year, y = gi_star)) +
  geom_line(colour = "#0072B2", linewidth = 1) +
  geom_point(colour = "#0072B2", size = 2) +
  labs(x = "Year", y = "Gi*",
       title = "Changsha's Gi* by year, 2005 to 2021") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"))

ggplotly(p)
```
:::

## Printing the Mann-Kendall Test Report

`MannKendall()` of **Kendall** runs the test. `unclass()` and `unnest_wider()` turn the result into a one-row table.

```{r}
cbg_mk <- cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
kable(cbg_mk, digits = 4)
```

For Changsha, tau is `r round(cbg_mk$tau, 3)` and the p-value (`sl`) is `r signif(cbg_mk$sl, 3)`. `r if (cbg_mk$sl < 0.05) "This is below 0.05, so I reject the null hypothesis of no trend." else "This is above 0.05, so I cannot reject the null hypothesis of no trend."` `r if (cbg_mk$tau > 0) "A positive tau means the trend is upward. Changsha's Gi* has risen over the period." else "A negative tau means the trend is downward. Changsha's Gi* has fallen over the period."`

::: {.callout-note}
## Learning diary
The report has five columns. `tau` is Kendall's rank correlation between Gi* and time, from -1 to 1. `sl` is the two-sided p-value. `S` is the score. It counts the pairs of years where the later year is higher, minus the pairs where it is lower. `D` is the largest possible value of S, and `varS` is the variance of S. So tau is S divided by D.
:::

## Mann-Kendall Test for Every County

`group_by(County)` repeats the test for all 88 counties.

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
```

`arrange()` sorts by p-value and `slice()` keeps the ten strongest trends.

```{r}
emerging <- ehsa %>%
  arrange(sl, abs(tau)) %>%
  slice(1:10)
emerging %>%
  mutate(sl = format.pval(sl, digits = 3)) %>%
  kable(digits = 4)
```

`r sum(ehsa$sl < 0.05)` of the 88 counties have a significant trend at 0.05. `r sum(ehsa$sl < 0.05 & ehsa$tau > 0)` of them are trending up and `r sum(ehsa$sl < 0.05 & ehsa$tau < 0)` are trending down.

::: {.callout-caution}
## Where I differ from the book
The book sorts with `arrange(sl, abs(tau))`, so ties in p-value put the weaker tau first. I kept it to match the book, but `arrange(sl, desc(abs(tau)))` would list the stronger trend first.
:::

# Performing Emerging Hot Spot Analysis

`emerging_hotspot_analysis()` of **sfdep** runs all the steps at once. `k = 1` uses one time lag for the neighbours, and `nsim = 99` runs 99 permutations.

```{r}
set.seed(1234)
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st,
  .var = "GDPPC",
  k = 1,
  nsim = 99)
```

## Visualising the Distribution of EHSA Classes

I colour hot spot classes in reds, oranges and pink, cold spot classes in blues, and "no pattern detected" in grey. The same colours are used on the map below. The names must match sfdep's output exactly, and sfdep spells oscillating with one l.

```{r}
class_cols <- c(
  "persistent hotspot"    = "#67001F",
  "intensifying hotspot"  = "#B2182B",
  "consecutive hotspot"   = "#D6604D",
  "new hotspot"           = "#F4A582",
  "sporadic hotspot"      = "#E69F00",
  "oscilating hotspot"    = "#CC79A7",
  "diminishing hotspot"   = "#FDDBC7",
  "historical hotspot"    = "#8C510A",
  "persistent coldspot"   = "#053061",
  "intensifying coldspot" = "#2166AC",
  "consecutive coldspot"  = "#4393C3",
  "new coldspot"          = "#92C5DE",
  "sporadic coldspot"     = "#56B4E9",
  "oscilating coldspot"   = "#0072B2",
  "diminishing coldspot"  = "#D1E5F0",
  "historical coldspot"   = "#5E4FA2",
  "no pattern detected"   = "grey80")
```

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
class_count <- count(ehsa, classification, sort = TRUE)

ggplot(class_count,
       aes(x = n, y = reorder(classification, n), fill = classification)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.3, size = 4.5) +
  scale_fill_manual(values = class_cols) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Number of counties", y = NULL,
       title = paste0(str_to_sentence(class_count$classification[1]),
                      " is the most common class")) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"))
```

## The Code

```{r}
#| eval: false
class_count <- count(ehsa, classification, sort = TRUE)

ggplot(class_count,
       aes(x = n, y = reorder(classification, n), fill = classification)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.3, size = 4.5) +
  scale_fill_manual(values = class_cols) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Number of counties", y = NULL,
       title = paste0(str_to_sentence(class_count$classification[1]),
                      " is the most common class")) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"))
```
:::

::: {.callout-caution}
## Where I differ from the book
The book uses `geom_bar()` with the classes on the x axis, where the long names overlap. I count the classes first and draw horizontal bars sorted by size, with the count printed on each bar.
:::

## Visualising EHSA

`left_join()` attaches the EHSA results to the county boundaries. In the EHSA output the county name is in `location`.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

`filter()` keeps the counties whose class is significant at 0.05. The map shows them on top of a grey base map of all counties.

::: panel-tabset
## The Plot

```{r}
#| echo: false
ehsa_sig <- hunan_ehsa %>%
  filter(p_value < 0.05) %>%
  mutate(classification = factor(classification,
                                 levels = intersect(names(class_cols),
                                                    classification)))

tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons(fill = "grey95", col = "grey60") +
tm_shape(ehsa_sig) +
  tm_polygons(fill = "classification",
              fill.scale = tm_scale_categorical(
                values = unname(class_cols[levels(ehsa_sig$classification)])),
              fill.legend = tm_legend(title = "EHSA class",
                                      position = tm_pos_out("right", "center")),
              col = "grey40") +
  tm_title("Significant emerging hot and cold spots (p < 0.05)",
           size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
ehsa_sig <- hunan_ehsa %>%
  filter(p_value < 0.05) %>%
  mutate(classification = factor(classification,
                                 levels = intersect(names(class_cols),
                                                    classification)))

tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons(fill = "grey95", col = "grey60") +
tm_shape(ehsa_sig) +
  tm_polygons(fill = "classification",
              fill.scale = tm_scale_categorical(
                values = unname(class_cols[levels(ehsa_sig$classification)])),
              fill.legend = tm_legend(title = "EHSA class",
                                      position = tm_pos_out("right", "center")),
              col = "grey40") +
  tm_title("Significant emerging hot and cold spots (p < 0.05)",
           size = 1.5, fontface = "bold")
```
:::

`r nrow(ehsa_sig)` of the 88 counties have a significant class.

```{r}
ehsa_sig %>%
  st_drop_geometry() %>%
  count(classification, name = "counties") %>%
  kable()
```

`r sum(ehsa_sig$classification %in% c("sporadic hotspot", "sporadic coldspot"))` of the `r nrow(ehsa_sig)` significant counties are sporadic hot or cold spots. A sporadic hot spot switches on and off. It is a significant hot spot in some years but not in others, and is never a cold spot. A sporadic cold spot is the same the other way round. So for most of Hunan the pattern of GDP per capita has not settled into lasting hot or cold spots over these 17 years. Only `r sum(ehsa_sig$classification == "consecutive hotspot")` county is a consecutive hot spot, and there are no persistent or intensifying ones.

::: {.callout-important}
## Check that caught something
My first render showed `r sum(grepl("oscilating", ehsa_sig$classification))` significant counties as "Missing" on the map. My colour list used "oscillating hotspot", but sfdep spells it "oscilating", with one l. The names did not match, so those classes got no colour. I changed the names in `class_cols` to sfdep's spelling. When colouring categories by name, it is worth checking `unique()` of the column first.
:::

::: {.callout-caution}
## Where I differ from the book
The book draws the significant counties with `tm_fill("classification")` and tmap's default colours, which include green. I use the colour list above so hot spots are always warm and cold spots always blue. The legend sits outside the map, and the factor keeps only the classes that appear, so the legend has no empty entries.
:::
