---
title: "Hands-on Exercise 1"
subtitle: "Geospatial Data Wrangling and Thematic Mapping"
author: "ZeXin Chen"
date: "2026-08-26"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

::: {.pagefacts}
[Six datasets, all downloaded fresh]{.fact}
[data.gov.sg, LTA DataMall, SingStat, Inside Airbnb]{.fact .data}
[`sf` `tmap` `tidyverse`]{.fact .pkg}
:::

# Getting Started

## Installing and Loading R Packages

Three packages are used in this exercise. **sf** handles the geospatial data, **tmap** draws the thematic maps, and **tidyverse** covers importing and wrangling.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

A quick version check, since the tmap syntax used later only exists in version 4.

```{r}
packageVersion("tmap")
```

# Geospatial Data Wrangling with sf

## Importing Geospatial Data

### Polygon feature data

data.gov.sg has retired the shapefile and KML versions of this layer, so the geojson is what is available. For geojson `st_read()` takes a single path rather than the `dsn` and `layer` pair the shapefile version needs.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

323 features, 13 fields, multipolygon. Two things are different from the book. The layer comes in as **WGS84**, not SVY21. And the area and length fields are called `SHAPE_1.AREA` and `SHAPE_1.LEN`, not `SHAPE_Area` and `SHAPE_Leng`. Both of these bite me later.

::: {.callout-note}
## Learning diary
Small thing, but worth writing down. The 2019 file downloads as `MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson`, with GEOJSON in the middle of the name rather than just the extension. Nobody would guess that. Now I copy filenames out of `ls` instead of typing what I think they are called.
:::

### Polyline feature data

```{r}
cyclingpath <- st_read(dsn = "data/geospatial",
                       layer = "CyclingPathGazette")
```

LTA gives this one as a shapefile, not geojson. So `st_read()` needs the `dsn` and `layer` pair instead of one path. Look at the CRS it reports too, this one is already in SVY21 while the data.gov.sg layers are not.

::: {.callout-warning}
## Issue I hit
The zip unpacked into a folder called `CyclingPath_Mar2026`, so my files sat one level deeper than the code expected and `st_read()` could not find the layer. I moved the nine `CyclingPathGazette.*` files up into `data/geospatial` and it worked. Nothing clever, I just had not looked at where the unzip actually put them.
:::

### Point feature data

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

2290 point features, in WGS84.

```{r}
names(preschool)
```

Worth pausing on this. The preschool layer has no real attribute columns. Everything the file knows about each centre is packed into a single `Description` string as an HTML table, a quirk of how data.gov.sg exports these layers. Oddly the two Master Plan boundary layers do not have this problem, they come through with proper columns, so it is not consistent across data.gov.sg exports. Here it does not matter either way, because every use of `preschool` below relies only on the geometry.

## Checking the Content of a Simple Feature Data Frame

Before wrangling anything it is worth looking at what we actually have.

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

This prints only the geometry column, so we see the feature type, extent and coordinate system without the attribute table getting in the way.

### Working with glimpse()

```{r}
glimpse(mpsz)
```

`glimpse()` shows the data type of every field. Two things differ from the book here. The shape fields are `SHAPE_1.AREA` and `SHAPE_1.LEN` here, not `SHAPE_Area` and `SHAPE_Leng`. And the `X_ADDR` and `Y_ADDR` centroid fields are gone.

### Working with head()

```{r}
head(mpsz, n = 5)
```

The `n` argument controls how many records are shown, which keeps the output readable.

## Plotting the Geospatial Data

The default plot of an sf object draws every attribute as its own small map.

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(mpsz)
```

## The Code

```{r}
#| eval: false
plot(mpsz)
```
:::

Passing only the geometry gives the outline on its own.

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz),
     col = "#EFE9DB", border = "grey65", lwd = 0.4,
     main = "Singapore planning subzones, Master Plan 2014")
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz),
     col = "#EFE9DB", border = "grey65", lwd = 0.4,
     main = "Singapore planning subzones, Master Plan 2014")
```
:::

Or a single named attribute.

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(mpsz["PLN_AREA_N"],
     key.pos = NULL, border = "grey80", lwd = 0.3,
     main = "Subzones coloured by planning area")
```

## The Code

```{r}
#| eval: false
plot(mpsz["PLN_AREA_N"],
     key.pos = NULL, border = "grey80", lwd = 0.3,
     main = "Subzones coloured by planning area")
```
:::

`plot()` is fine for a quick look. Anything meant for presentation should go through tmap instead, which is the whole of the second half of this exercise.

## Working with Projection

Two layers can only be used together if they share a coordinate system, so this step is not optional.

### Checking what we actually have

```{r}
st_crs(mpsz)
```

Here my file and the book part ways, so I had to stop and think about it.

The book reads a shapefile that really is in SVY21 but has the wrong EPSG code attached. The coordinates are fine, only the label is wrong, so it fixes the label with `st_set_crs(3414)`.

Mine is a different problem. It says WGS84 and it means it. The bounding box runs 103.6 to 104.0 across and 1.15 to 1.47 up, and those are degrees. SVY21 would give me metres in the tens of thousands.

So if I copied `st_set_crs(3414)` I would be sticking an SVY21 label on degree values without changing the numbers underneath. `st_transform()` is the one that actually converts them.

::: {.callout-note}
## Learning diary
This is the one that would have got me. Nothing errors either way. The map still draws, the code still runs, and every area and buffer after it is quietly wrong.

What saved me was reading the bounding box instead of skipping past it. Degrees look like 103.8. Metres look like 28,000. Once I noticed that, the rest was obvious.

The rule I am taking from this. `st_set_crs` fixes a wrong label. `st_transform` changes the actual coordinates. Ask which one is broken before picking.
:::

```{r}
mpsz3414 <- st_transform(mpsz, crs = 3414)
```

```{r}
st_crs(mpsz3414)$epsg
```

```{r}
st_bbox(mpsz3414)
```

The bounding box is now in metres, in the tens of thousands. That is my quick check that the transform actually did something rather than just changing the label.

### Transforming preschool to svy21

Same situation, same tool.

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

```{r}
st_geometry(preschool3414)
```

Values well outside the 0 to 360 range of decimal degrees, so this one worked too.

### Overlaying the two layers

Before the transform, the preschool points would not have landed on the subzone map at all. Now they do.

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "2,290 pre-schools over the subzone boundaries")
plot(st_geometry(preschool3414), add = TRUE,
     pch = 21, cex = 0.42, bg = "#0072B2AA", col = "#00548A", lwd = 0.3)
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "2,290 pre-schools over the subzone boundaries")
plot(st_geometry(preschool3414), add = TRUE,
     pch = 21, cex = 0.42, bg = "#0072B2AA", col = "#00548A", lwd = 0.3)
```
:::

## Importing and Converting Aspatial Data

The Airbnb listings file is not geospatial data, but it carries latitude and longitude columns, which is enough to turn it into one.

### Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
glimpse(listings)
```

The two fields we need are `latitude` and `longitude`, both in decimal degrees. The safe assumption is WGS84.

### Creating a simple feature data frame

```{r}
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)
```

Three things about the arguments. `coords` wants x first then y, so longitude comes before latitude. `crs = 4326` is WGS84, the system the raw coordinates are in. The piped `st_transform()` then moves it to 3414 to match everything else.

```{r}
glimpse(listings_sf)
```

A `geometry` column has appeared and the separate longitude and latitude columns are gone.

### Plotting the listings over the subzones

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "Airbnb listings cluster in the centre and south")
plot(st_geometry(listings_sf), add = TRUE,
     pch = 20, cex = 0.3, col = "#D55E0099")
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "Airbnb listings cluster in the centre and south")
plot(st_geometry(listings_sf), add = TRUE,
     pch = 20, cex = 0.3, col = "#D55E0099")
```
:::

Most listings sit in the central and southern subzones. They thin out quickly towards the west and the north coast, with a few clusters in the east near the airport. Big parts of the island have almost none.

## Geoprocessing with sf

### Use case 1, land acquisition analysis

The scenario is that the authority wants to widen the existing cycling paths and needs 5 metres of reserved land on either side. The task is to work out how much land that comes to.

`dist = 5` in `st_buffer()` means five units of whatever coordinate system the layer is in. On a WGS84 layer that would be five degrees, a buffer some 550 km wide, and it would run without error and give a meaningless answer.

The cycling path is the one layer here that already arrived in SVY21, so this transform changes nothing. I have kept it anyway rather than relying on LTA continuing to ship it that way. It costs nothing and it makes the assumption explicit.

```{r}
cyclingpath3414 <- st_transform(cyclingpath, crs = 3414)
st_crs(cyclingpath3414)$epsg
```

Now `st_buffer()` builds buffers that really are 5 metres.

```{r}
buffer_cycling <- st_buffer(cyclingpath3414, dist = 5, nQuadSegs = 30)
```

Then the area of each buffer.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

And the total.

```{r}
sum(buffer_cycling$AREA)
```

To see what this looks like on the ground, we can clip the buffers to a single subzone. Tampines West here.

```{r}
mpsz_selected <- mpsz3414 %>%
  filter(SUBZONE_N == "TAMPINES WEST")

buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected)
```

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz_selected),
     col = "#F7F4EC", border = "grey55", lwd = 0.8,
     main = "Five metre reserve along cycling paths, Tampines West")
plot(st_geometry(buffer_cycling_selected), add = TRUE,
     col = "#D55E00", border = NA)
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz_selected),
     col = "#F7F4EC", border = "grey55", lwd = 0.8,
     main = "Five metre reserve along cycling paths, Tampines West")
plot(st_geometry(buffer_cycling_selected), add = TRUE,
     col = "#D55E00", border = NA)
```
:::

### Use case 2, counting pre-schools by planning subzone

The scenario is that the authority needs a pre-school count for every planning subzone to feed into forward planning.

`st_intersects()` returns, for each subzone, the list of preschools falling inside it. Wrapping that in `lengths()` turns the lists into counts in one step.

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

This is `st_intersects()`, not `st_intersection()`.

::: {.callout-note}
## Learning diary
These two are one letter apart and do different jobs, so this is worth saying out loud before typing either. `st_intersects()` answers which things overlap and gives you a list. `st_intersection()` cuts the shapes and gives you the overlapping pieces back. Only the first one is any use for counting.
:::

```{r}
summary(mpsz3414$`PreSch Count`)
```

The median subzone has only a handful of pre-schools while the maximum is far higher, so the distribution is heavily skewed.

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

Counts on their own are misleading because subzones differ a lot in size, so we convert to a density per square kilometre.

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
```

### Exploratory data analysis

The base R histogram is quick but hard to customise.

::: panel-tabset
## The Plot

```{r}
#| echo: false
dens <- as.numeric(mpsz3414$`PreSch Density`)

hist(dens,
     breaks = 20, col = "#CFE3F0", border = "white",
     main = "Pre-school density is heavily skewed",
     xlab = "Pre-schools per square kilometre",
     ylab = "Number of subzones")
abline(v = median(dens, na.rm = TRUE), col = "#D55E00", lwd = 2, lty = 2)
legend("topright", bty = "n", col = "#D55E00", lwd = 2, lty = 2,
       legend = sprintf("median %.1f", median(dens, na.rm = TRUE)))
```

## The Code

```{r}
#| eval: false
dens <- as.numeric(mpsz3414$`PreSch Density`)

hist(dens,
     breaks = 20, col = "#CFE3F0", border = "white",
     main = "Pre-school density is heavily skewed",
     xlab = "Pre-schools per square kilometre",
     ylab = "Number of subzones")
abline(v = median(dens, na.rm = TRUE), col = "#D55E00", lwd = 2, lty = 2)
legend("topright", bty = "n", col = "#D55E00", lwd = 2, lty = 2,
       legend = sprintf("median %.1f", median(dens, na.rm = TRUE)))
```
:::

ggplot2 gives a far more presentable version, with room for a proper title and axis labels.

::: panel-tabset
## The Plot

```{r}
#| echo: false
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```
:::

Same story as the summary. Over a hundred subzones are in the lowest bin and the count drops off fast after that, with a few stragglers past 30 per square kilometre.

A scatterplot of count against density shows how the two relate.

::: panel-tabset
## The Plot

```{r}
#| echo: false
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```
:::

Count and density do rise together, but the spread is wide. Two subzones with the same density can be twenty pre-schools apart. The points on the far right, around 35 per square kilometre, only have a few pre-schools each because those subzones are small.

::: {.column-margin}
**Watch the axis limits.** `xlim` and `ylim` stop at 40 here. Anything past that is dropped from the plot without a warning, so an outlier just vanishes instead of showing up.
:::

One thing to watch. `xlim` and `ylim` cut the axes at 40, so any point past that is dropped from the plot instead of showing up as an outlier. The chart looks tidier than the data actually is.

# Thematic Mapping and Geovisualisation with tmap

## Importing the Data

This half uses two different datasets, the Master Plan 2019 subzone boundaries and the 2024 resident population figures.

### Importing the geospatial data

```{r}
mpsz19 <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

### Checking the attributes

```{r}
names(mpsz19)
```

The reference book at this point runs a chunk of `rvest` code to parse attributes out of an HTML table buried in a `Description` field. That was necessary for the KML version of this layer, which data.gov.sg has since unpublished. The geojson that replaced it carries `SUBZONE_N`, `PLN_AREA_N` and `REGION_N` as proper columns, so there is nothing to parse and the join below can use them directly.

The geojson is also in WGS84, which is why `st_transform()` is piped onto the import above.

```{r}
mpsz19
```

332 features and 13 fields, with the three name columns the join needs.

### Importing the attribute data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

## Data Preparation

The target is one row per subzone with six fields. `YOUNG` is ages 0 to 24, `ECONOMY ACTIVE` is 25 to 64, and `AGED` is 65 and over. `DEPENDENCY` divides the young plus the aged by the economically active.

First the aggregate and pivot.

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP)
```

Now a step the reference book does not take, and which turns out to matter a great deal.

```{r}
names(popdata2024)
```

The book builds the three age bands by column number, like `rowSums(.[3:6]) + rowSums(.[12])` for `YOUNG`. That only works if the age groups come out in a particular order. Look at the list above. They do not.

`pivot_wider()` names the new columns after the values it found, then those names get sorted as text. As text, `5_to_9` comes after `55_to_59`, so it sits at position 14 instead of 12.

Run the book's numbers against my columns and `YOUNG` grabs the 50 to 54 band instead of 5 to 9. `ECONOMY ACTIVE` ends up counting primary school children.

::: {.callout-warning}
## Issue I hit
Like the projection one, this does not error. It just gives you a wrong dependency ratio, and every map after it looks completely normal.

I nearly skipped `names(popdata2024)`. It was one line and it caught the whole thing.
:::

Selecting by name instead removes the dependency on ordering entirely.

```{r}
young_ag <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19", "20_to_24")

active_ag <- c("25_to_29", "30_to_34", "35_to_39", "40_to_44",
               "45_to_49", "50_to_54", "55_to_59", "60_to_64")

aged_ag <- c("65_to_69", "70_to_74", "75_to_79",
             "80_to_84", "85_to_89", "90_and_over")
```

A guard against the same class of problem returning. If a future release renames or adds a band, this stops the render rather than letting a wrong number through.

```{r}
stopifnot(all(c(young_ag, active_ag, aged_ag) %in% names(popdata2024)))
```

```{r}
popdata2024 <- popdata2024 %>%
  mutate(
    YOUNG = rowSums(across(all_of(young_ag))),
    `ECONOMY ACTIVE` = rowSums(across(all_of(active_ag))),
    AGED = rowSums(across(all_of(aged_ag))),
    TOTAL = rowSums(across(all_of(c(young_ag, active_ag, aged_ag)))),
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

Now a check. The three bands should add up to the total, and the ratio for Singapore overall should be somewhere near 0.7.

```{r}
popdata2024 %>%
  summarise(
    young = sum(YOUNG),
    active = sum(`ECONOMY ACTIVE`),
    aged = sum(AGED),
    total = sum(TOTAL),
    bands_match_total = sum(YOUNG) + sum(`ECONOMY ACTIVE`) + sum(AGED) == sum(TOTAL),
    overall_dependency = (sum(YOUNG) + sum(AGED)) / sum(`ECONOMY ACTIVE`)
  )
```

::: {.callout-important}
## Check that caught it
The bands add to 4,189,460 and Singapore's published resident population is about 4.19 million, so that matches. The overall ratio is 0.732, which is in the right range.

Both had to be right before I trusted any map on this page. If I had used the book's column numbers, this check is where it would have shown up.
:::

### Joining the attribute and geospatial data

The `PA` and `SZ` values are mixed case while `PLN_AREA_N` and `SUBZONE_N` are uppercase, so they will not match until we fix that.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
mpsz_pop2024 <- left_join(mpsz19, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

The sf object goes on the left of the join deliberately. That is what keeps the result an sf data frame rather than a plain tibble.

```{r}
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## Choropleth Mapping with tmap

### The quick way, qtm()

::: panel-tabset
## The Plot

```{r}
#| echo: false
tmap_mode("plot")
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY",
    title = "The default bins, left exactly as qtm gives them")
```

## The Code

```{r}
#| eval: false
tmap_mode("plot")
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY",
    title = "The default bins, left exactly as qtm gives them")
```
:::

`qtm()` is quick to write, but look at what the default bins did. The legend runs 0 to 10 in steps of two. Almost every subzone is under 1, so the whole island lands in the palest class and one dark polygon sits near Changi.

The map is not wrong. It just does not show anything.

::: {.callout-note}
## Learning diary
One subzone with a ratio of 10 is enough to wreck the default bins for all 332 of them. That is why the custom breaks section below matters.

It is also why I stopped using `qtm()` after this point. It is fine for a first look but it gives you no way to change the bins.
:::

### Drawing a base map

`tm_shape()` sets the data, then layer functions draw on top of it.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons()
```
:::

### Drawing a choropleth with tm_polygons()

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```
:::

The default binning is "pretty" and the default palette is blues3. Missing values come out grey.

### tm_fill() and tm_borders() separately

`tm_polygons()` is really a wrapper around these two.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```
:::

Adding the borders back in.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```
:::

`tm_borders()` takes `col`, `lwd` and `lty` for colour, line width and line type.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60", lwd = 0.1, lty = "dashed")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60", lwd = 0.1, lty = "dashed")
```
:::

## Data Classification Methods

tmap offers ten classification styles. The choice matters more than it looks, because the same data can tell quite different stories depending on where the breaks fall.

### Quantile

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile puts an equal number of subzones in each class")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile puts an equal number of subzones in each class")
```
:::

### Equal interval

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval hands almost every subzone to one class")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval hands almost every subzone to one class")
```
:::

Same column, same data, two maps that look nothing alike.

Quantile puts the same number of subzones in each class, so the breaks come out at 0.64, 0.70, 0.74 and 0.82. Those all sit inside the range where the data actually is, and the map separates the central and eastern subzones clearly.

Equal interval cuts 0 to 10 into five equal parts. Nearly everything falls in the first one, so the island goes pale with a single dark polygon. If you only saw that map you would think dependency hardly varies in Singapore, which is not true.

Neither map is lying. They are the same numbers. But one subzone at 10 is enough to make the equal interval version useless, and I would not have known that without putting them side by side.

### Custom breaks

Always look at the numbers before choosing break points.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Most values sit between 0.6 and 1.0, so breaks at 0.60, 0.70, 0.80 and 0.90 will separate the bulk of the data. Remember tmap needs n+1 values including a minimum and maximum.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Breaks chosen from the summary rather than by the default")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Breaks chosen from the summary rather than by the default")
```
:::

## Colour Scheme

### ColorBrewer palettes

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.greens"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Same breaks, a different ramp")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.greens"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Same breaks, a different ramp")
```
:::

A minus prefix reverses the ramp.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```
:::

### Cartographic furniture

A compass, scale bar, grid and credits turn a plot into something that reads as a map.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```
:::

## Map Layout

### Legend

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```
:::

### Map style

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```
:::

Resetting to the default so it does not carry over into later chunks.

```{r}
tmap_style("white")
```

## Small Multiple Choropleth Maps

Facet maps put several maps side by side, which makes it easy to see how a spatial pattern shifts with another variable.

### By passing multiple variables to an aesthetic

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```
:::

### By arranging separate maps in a grid

::: panel-tabset
## The Plot

```{r}
#| echo: false
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Where the young live, as a raw count")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Where the aged live, as a raw count")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

## The Code

```{r}
#| eval: false
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Where the young live, as a raw count")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Where the aged live, as a raw count")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```
:::

The two maps look almost the same. I expected the aged map to light up the older estates and the young map to light up the newer towns, but the same subzones are dark on both.

That is because both are raw counts. A subzone with more people in it has more of everyone. So these maps are mostly showing where people live, not where the population is young or old.

::: {.callout-tip}
## Alternative explored
The book stops at the count maps. Since they were not answering my question, I redrew both as shares of each subzone's own population instead.
:::

Dividing each band by the subzone total takes population size out of it.

```{r}
mpsz_pop2024 <- mpsz_pop2024 %>%
  mutate(`YOUNG PCT` = YOUNG / TOTAL * 100,
         `AGED PCT` = AGED / TOTAL * 100)
```

::: panel-tabset
## The Plot

```{r}
#| echo: false
youngpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG PCT",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Young as a share of the subzone")

agedpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED PCT",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Aged as a share of the subzone")

tmap_arrange(youngpct, agedpct, asp = 1, ncol = 2)
```

## The Code

```{r}
#| eval: false
youngpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG PCT",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Young as a share of the subzone")

agedpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED PCT",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Aged as a share of the subzone")

tmap_arrange(youngpct, agedpct, asp = 1, ncol = 2)
```
:::

Now the two maps look different, which is what I expected the first pair to show. Read them against the count maps above to see how much of that first pattern was just population size.

### By a group-by variable in tm_facets()

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile",
                                          values = "brewer.blues"),
          fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_facets(by = "REGION_N", nrow = 2, ncols = 3,
            free.coords = TRUE, drop.units = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Dependency by region, each panel on its own extent")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile",
                                          values = "brewer.blues"),
          fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_facets(by = "REGION_N", nrow = 2, ncols = 3,
            free.coords = TRUE, drop.units = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Dependency by region, each panel on its own extent")
```
:::

## Mapping a Selection Criterion

Rather than facetting, we can just filter down to the region of interest.

::: panel-tabset
## The Plot

```{r}
#| echo: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```
:::

## Complementing the Map with a Statistical Chart

A map shows where, a chart shows how much. `fill.chart` puts both in one figure.

::: panel-tabset
## The Plot

```{r}
#| echo: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

## The Code

```{r}
#| eval: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```
:::

The boxplot shows something the map does not. All the outliers are below the box, near zero, not above it. Those are subzones with working age residents and hardly any children or elderly.

Once labelled, they turn out to be bunched in the downtown core. The labels overlap a lot at this size and would need fixing before I used this anywhere, but you can still see where they are.

::: panel-tabset
## The Plot

```{r}
#| echo: false
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## The Code

```{r}
#| eval: false
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```
:::

## Creating an Interactive Map

Switching `tmap_mode()` to view turns the same code into a zoomable, clickable map.

::: panel-tabset
## The Plot

```{r}
#| echo: false
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

## The Code

```{r}
#| eval: false
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```
:::

`set_zoom_limits` is doing real work here. Total freedom to zoom sounds generous but in practice users end up lost somewhere over the ocean, so bounding the range keeps them inside the study area.

```{r}
tmap_mode("plot")
```

# Beyond the Exercise

The two chapters finish at the choropleth. Three things bothered me once I got there, so I went looking.

First, whether any of this is actually spatial. Second, whether the pattern I mapped survives a change in the size of the units. Third, whether the map itself is misleading me about who lives where.

## Is this actually spatial analysis?

Anselin's lecture gives a test for this. If you move things around and nothing changes, it was not spatial analysis. If it does change, it was.

Easy enough to run on my own data. `sample()` only reshuffles which subzone holds which number, so every value is preserved.

```{r}
set.seed(1234)
mpsz_shuffled <- mpsz_pop2024
mpsz_shuffled$DEPENDENCY <- sample(mpsz_pop2024$DEPENDENCY)
```

If the values are the same, any summary that ignores location should be identical.

```{r}
tibble(
  version  = c("Real Singapore", "Same numbers, shuffled"),
  subzones = c(sum(!is.na(mpsz_pop2024$DEPENDENCY)),
               sum(!is.na(mpsz_shuffled$DEPENDENCY))),
  mean     = c(mean(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               mean(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  sd       = c(sd(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               sd(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  median   = c(median(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               median(mpsz_shuffled$DEPENDENCY, na.rm = TRUE))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

Same for the histogram.

::: panel-tabset
## The Plot

```{r}
#| echo: false
bind_rows(
  tibble(version = "Real Singapore",         value = mpsz_pop2024$DEPENDENCY),
  tibble(version = "Same numbers, shuffled", value = mpsz_shuffled$DEPENDENCY)
) %>%
  filter(!is.na(value), value < 2) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white", linewidth = 0.2) +
  facet_wrap(~version) +
  labs(title = "The same histogram twice",
       subtitle = "Values under 2 only, so the single subzone at 10 does not flatten the axis",
       x = "Dependency ratio", y = "Subzones") +
  theme_minimal()
```

## The Code

```{r}
#| eval: false
bind_rows(
  tibble(version = "Real Singapore",         value = mpsz_pop2024$DEPENDENCY),
  tibble(version = "Same numbers, shuffled", value = mpsz_shuffled$DEPENDENCY)
) %>%
  filter(!is.na(value), value < 2) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white", linewidth = 0.2) +
  facet_wrap(~version) +
  labs(title = "The same histogram twice",
       subtitle = "Values under 2 only, so the single subzone at 10 does not flatten the axis",
       x = "Dependency ratio", y = "Subzones") +
  theme_minimal()
```
:::

Now the maps.

::: panel-tabset
## The Plot

```{r}
#| echo: false
shuffle_map <- function(dat, ttl) {
  tm_shape(dat) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(show = FALSE)) +
    tm_borders(fill_alpha = 0.4) +
    tm_title(ttl, size = 0.9)
}

tmap_arrange(shuffle_map(mpsz_pop2024,  "Real Singapore"),
             shuffle_map(mpsz_shuffled, "Same numbers, shuffled"),
             ncol = 2)
```

## The Code

```{r}
#| eval: false
shuffle_map <- function(dat, ttl) {
  tm_shape(dat) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(show = FALSE)) +
    tm_borders(fill_alpha = 0.4) +
    tm_title(ttl, size = 0.9)
}

tmap_arrange(shuffle_map(mpsz_pop2024,  "Real Singapore"),
             shuffle_map(mpsz_shuffled, "Same numbers, shuffled"),
             ncol = 2)
```
:::

::: {.callout-note}
## Learning diary
Every number is the same. The mean, the spread, the median, the histogram. And the two maps look nothing alike.

So whatever the difference is between those two pictures, none of the statistics I know how to compute can see it. That gap is what the rest of this course is for. Lesson 4 defines what neighbouring means and Lesson 5 turns this difference into a single number I can test.

It also answers something I had wondered at the start, which was why not just do this in pandas. A non-spatial toolkit would tell me these two datasets are identical, because on everything it can measure, they are.
:::

## Does the pattern survive a change of unit?

This is the Modifiable Areal Unit Problem. Results shift when you change the size or the shape of the units you are measuring on.

Singapore's subzones sit inside planning areas, so I can test the size half of it directly.

```{r}
mpsz_pa <- mpsz_pop2024 %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL)) %>%
  group_by(PLN_AREA_N) %>%
  summarise(YOUNG  = sum(YOUNG),
            ACTIVE = sum(`ECONOMY ACTIVE`),
            AGED   = sum(AGED),
            n_subzones = n(),
            .groups = "drop") %>%
  mutate(DEPENDENCY_PA = (YOUNG + AGED) / ACTIVE) %>%
  filter(is.finite(DEPENDENCY_PA))
```

```{r}
tibble(
  unit   = c("Subzone", "Planning area"),
  n      = c(sum(!is.na(mpsz_pop2024$DEPENDENCY)), nrow(mpsz_pa)),
  min    = c(min(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),    min(mpsz_pa$DEPENDENCY_PA)),
  median = c(median(mpsz_pop2024$DEPENDENCY, na.rm = TRUE), median(mpsz_pa$DEPENDENCY_PA)),
  max    = c(max(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),    max(mpsz_pa$DEPENDENCY_PA))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))
```

::: panel-tabset
## The Plot

```{r}
#| echo: false
bind_rows(
  tibble(unit = "Subzone",
         value = mpsz_pop2024$DEPENDENCY[!is.na(mpsz_pop2024$DEPENDENCY)]),
  tibble(unit = "Planning area", value = mpsz_pa$DEPENDENCY_PA)
) %>%
  ggplot(aes(x = value, fill = unit)) +
  geom_density(alpha = 0.45) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Aggregating flattens the spread",
       subtitle = "Zoomed to 0 to 1.5. The density still uses every value, including the outlier",
       x = "Dependency ratio", y = "Density", fill = NULL) +
  theme_minimal()
```

## The Code

```{r}
#| eval: false
bind_rows(
  tibble(unit = "Subzone",
         value = mpsz_pop2024$DEPENDENCY[!is.na(mpsz_pop2024$DEPENDENCY)]),
  tibble(unit = "Planning area", value = mpsz_pa$DEPENDENCY_PA)
) %>%
  ggplot(aes(x = value, fill = unit)) +
  geom_density(alpha = 0.45) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Aggregating flattens the spread",
       subtitle = "Zoomed to 0 to 1.5. The density still uses every value, including the outlier",
       x = "Dependency ratio", y = "Density", fill = NULL) +
  theme_minimal()
```
:::

::: {.callout-note}
## Learning diary
The planning area curve is taller and narrower. Same people, same formula, and the extremes have gone.

A map drawn at planning area level would make Singapore look far more even than it is. Neither unit is the right one. All I can do is say which one I used and admit the answer depends on it.

I could only test the size half of MAUP. The other half is that results also change when you redraw the boundaries at the same size, and I have no alternative zoning to try that with.
:::

## Who does the map actually show?

A choropleth colours areas, and the eye reads area as amount. So a big empty subzone shouts while a small crowded one whispers.

That is measurable.

```{r}
mpsz_pop2024 %>%
  filter(!is.na(TOTAL)) %>%
  mutate(area_km2 = as.numeric(st_area(.)) / 1e6) %>%
  st_drop_geometry() %>%
  summarise(
    `Map area held by the least populated half of subzones (%)` =
      round(sum(area_km2[TOTAL <= median(TOTAL)]) / sum(area_km2) * 100, 1),
    `Population those same subzones hold (%)` =
      round(sum(TOTAL[TOTAL <= median(TOTAL)]) / sum(TOTAL) * 100, 1)
  ) %>%
  pivot_longer(everything(), names_to = "measure", values_to = "value")
```

::: {.callout-note}
## Learning diary
Put those two numbers next to each other and the map stops looking neutral. Anyone reading it is weighting land, not people.

This is the same problem as the young and aged count maps further up, arriving from a different direction. There I was mapping counts when I wanted rates. Here I am mapping rates but the canvas itself is weighted by area.
:::

## Limitations

Worth stating what this does not cover.

- Part 1 uses the 2014 Master Plan boundaries and Part 2 uses 2019, because that is what the chapters specify. Subzone definitions changed between them, so the pre-school counts and the dependency ratios are not measured on quite the same geography.
- 94 subzones have no population data and are grey on every map. They are dropped rather than shown as zero.
- The dependency ratio treats 25 to 64 as working age. That is a convention, not a fact, and it is drifting.
- Point in polygon counting ignores catchments. A pre-school just over a subzone boundary serves those children anyway.
- Only the scale half of MAUP is tested, as noted above.

::: {.callout-warning}
## Issue I hit
`na.rm = TRUE` does not remove `Inf`. I assumed it did.

If a subzone has residents but nobody of working age, the dependency ratio divides by zero and gives `Inf`, and then `mean(x, na.rm = TRUE)` returns `Inf` for the whole column. `is.finite()` is the filter that actually catches it.

My data avoids this because I filtered on `ECONOMY ACTIVE > 0` before the join, but that was luck rather than intent.
:::

# Reflections

Four things I am taking away from this one.

**Most of what went wrong was small.** The cycling path zip unpacked into a subfolder and my path was one level off. The 2019 file had GEOJSON stuck in the middle of its name. And `st_intersects` and `st_intersection` are one letter apart with completely different jobs, which is a trap waiting to happen. None of these were hard, they just cost time, and all of them came from not looking at what was actually on disk before writing the code.

**Two things went wrong quietly, and those were the dangerous ones.** The projection and the age band ordering. Neither threw an error. The code ran, the maps drew, and the numbers would have been wrong. What caught both was checking output I could have skipped, the bounding box in one case and `names()` in the other.

**I got one of my own readings wrong.** I looked at the young and aged maps and expected them to show different parts of the island. They look almost identical. Both are counts, and counts mostly follow how many people live somewhere. I had to redraw them as rates to answer the question I thought I was already asking.

**The classification method decides a lot.** Quantile and equal interval on the same column gave me two maps that say opposite things about whether dependency varies across Singapore. One subzone at 10 flattens the equal interval version completely. Picking a classification is part of the analysis, not a formatting choice at the end.
