---
title: "Hands-on Exercise 1"
subtitle: "Geospatial Data Wrangling and Thematic Mapping"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "August 26, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

::: {.pagefacts}
[Six datasets, all downloaded fresh]{.fact}
[data.gov.sg, LTA DataMall, SingStat, Inside Airbnb]{.fact .data}
[`sf` `tmap` `tidyverse`]{.fact .pkg}
:::

# Overview

This exercise covers the first two chapters of the course book. The first half uses **sf** to import, inspect, reproject and process geospatial data (subzone boundaries, cycling paths, pre-schools and Airbnb listings). The second half uses **tmap** to turn the 2024 resident population into choropleth maps of the dependency ratio.

The main question is where in Singapore the young and the aged make up a large share of residents compared with the working-age population. I also want to see how much the answer depends on how the map is drawn.

# Loading the Packages

I use three packages. **sf** handles the geospatial data, **tmap** draws the thematic maps, and **tidyverse** covers importing and wrangling. `p_load()` of **pacman** installs any that are missing and loads them all.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

`tmap_options()` makes the legend title and text a little bigger on every tmap map on this page, so the legends stay readable at page width.

```{r}
tmap_options(legend.title.size = 1.1, legend.text.size = 0.8)
```

`packageVersion()` checks the tmap version, because the syntax used later only exists in version 4.

```{r}
packageVersion("tmap")
```

# Preparing the Data

The first half needs three geospatial layers. I import them one at a time and check each one before doing anything else.

## Importing Geospatial Data

::: panel-tabset
## Importing the polygon feature data

data.gov.sg has retired the shapefile and KML versions of this layer, so only the geojson is available. For geojson, `st_read()` of **sf** takes a single path instead of the `dsn` and `layer` pair that a shapefile needs.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

It has 323 multipolygon features and 13 fields. Two things differ from the book. The layer is in **WGS84**, not SVY21. The area and length fields are called `SHAPE_1.AREA` and `SHAPE_1.LEN`, not `SHAPE_Area` and `SHAPE_Leng`. Both caused problems later.

## Importing the polyline feature data

LTA gives this one as a shapefile, so `st_read()` needs the `dsn` and `layer` pair instead of one path.

```{r}
cyclingpath <- st_read(dsn = "data/geospatial",
                       layer = "CyclingPathGazette")
```

The CRS it reports is already SVY21, while the data.gov.sg layers are not.

::: {.callout-warning}
## Issue I hit
The zip unpacked into a folder called `CyclingPath_Mar2026`, so my files sat one level deeper than the code expected. `st_read()` could not find the layer. I moved the nine `CyclingPathGazette.*` files up into `data/geospatial` and it worked. I simply had not checked where the unzip put them.
:::

## Importing the point feature data

The pre-school locations come from data.gov.sg as geojson, so it is the single path form of `st_read()` again.

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

It has 2290 point features in WGS84. `names()` lists its columns.

```{r}
names(preschool)
```

The preschool layer has no real attribute columns. Each centre's details are packed into one `Description` string as an HTML table, because of how data.gov.sg exports these layers. Oddly, the two Master Plan boundary layers come with proper columns, so data.gov.sg exports are not consistent. This does not matter here, because I only use the geometry of `preschool`.
:::

## Checking the Content of a Simple Feature Data Frame

Before wrangling anything, I check what the data looks like.

::: panel-tabset
## Working with st_geometry()

`st_geometry()` prints only the geometry column, so I can see the feature type, extent and coordinate system without the attribute table.

```{r}
st_geometry(mpsz)
```

## Working with glimpse()

`glimpse()` of **dplyr** shows the data type of every field.

```{r}
glimpse(mpsz)
```

Two things differ from the book. The shape fields are `SHAPE_1.AREA` and `SHAPE_1.LEN`, not `SHAPE_Area` and `SHAPE_Leng`. The `X_ADDR` and `Y_ADDR` centroid fields are also gone.

## Working with head()

`head()` shows the first few complete records. The `n` argument sets how many, which keeps the output short.

```{r}
head(mpsz, n = 5)
```
:::

# Plotting the Geospatial Data

By default `plot()` draws each attribute of an sf object as a small map, up to the first nine. This is the book's demo. I only changed the palette, because the default `sf.colors` ramp runs through green. I pass `hcl.colors()` with "Plasma" instead.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(oma = c(0, 0, 3, 0))
plot(mpsz, pal = function(n) hcl.colors(n, "Plasma"))
mtext("The first nine attributes of the 2014 subzone layer",
      outer = TRUE, line = 1, cex = 1.3, font = 2)
```

## The Code

```{r}
#| eval: false
par(oma = c(0, 0, 3, 0))
plot(mpsz, pal = function(n) hcl.colors(n, "Plasma"))
mtext("The first nine attributes of the 2014 subzone layer",
      outer = TRUE, line = 1, cex = 1.3, font = 2)
```
:::

Plotting only `st_geometry()` draws just the outline.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz),
     col = "#EFE9DB", border = "grey65", lwd = 0.4,
     main = "Singapore planning subzones, Master Plan 2014")
```

## The Code

```{r}
#| eval: false
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz),
     col = "#EFE9DB", border = "grey65", lwd = 0.4,
     main = "Singapore planning subzones, Master Plan 2014")
```
:::

I can also plot one named attribute. With 55 planning areas, no palette can keep every one distinct. So this map is only good for a quick look at how the subzones group.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5.5
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(mpsz["PLN_AREA_N"],
     key.pos = NULL, border = "grey80", lwd = 0.3,
     pal = function(n) hcl.colors(n, "Plasma"),
     main = "Subzones coloured by planning area")
```

## The Code

```{r}
#| eval: false
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(mpsz["PLN_AREA_N"],
     key.pos = NULL, border = "grey80", lwd = 0.3,
     pal = function(n) hcl.colors(n, "Plasma"),
     main = "Subzones coloured by planning area")
```
:::

`plot()` is fine for a quick look. For maps meant for presentation I use tmap, which is the second half of this exercise.

# Working with Projection

Two layers can only be used together if they share a coordinate system, so this step is needed.

## Transforming mpsz to SVY21

`st_crs()` shows the coordinate reference system the layer actually carries.

```{r}
st_crs(mpsz)
```

::: {.callout-caution}
## Where I differ from the book
The book reads a shapefile that really is in SVY21 but has the wrong EPSG code attached. The coordinates are fine and only the label is wrong, so the book fixes the label with `st_set_crs(3414)`.

My file has a different problem. It really is WGS84. The bounding box runs 103.6 to 104.1 across and 1.16 to 1.47 up, and those are degrees. In SVY21 the values would be metres in the tens of thousands.

If I copied `st_set_crs(3414)`, I would put an SVY21 label on degree values without changing the numbers. `st_transform()` actually converts them, so I use that.
:::

::: {.callout-note}
## Learning diary
This mistake is easy to make, because nothing errors either way. The map still draws, but every area and buffer after it is wrong.

What helped was reading the bounding box instead of skipping past it. Degrees look like 103.8 and metres look like 28,000.

What I learned is that `st_set_crs` fixes a wrong label, while `st_transform` changes the actual coordinates. I should check which one is broken before picking.
:::

::: panel-tabset
## Transforming the layer

`st_transform()` reprojects the coordinates into SVY21, EPSG 3414.

```{r}
mpsz3414 <- st_transform(mpsz, crs = 3414)
```

## Checking the EPSG code

`st_crs()` again, this time returning only the EPSG code.

```{r}
st_crs(mpsz3414)$epsg
```

## Checking the bounding box

`st_bbox()` returns the extent in the new units.

```{r}
st_bbox(mpsz3414)
```
:::

The bounding box is now in metres, in the tens of thousands. This shows the transform changed the coordinates and not just the label.

## Transforming Preschool from WGS84 to SVY21

The pre-school layer is also in WGS84, so the same fix applies.

::: panel-tabset
## Transforming the layer

`st_transform()` moves the pre-school points into EPSG 3414.

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

## Checking the result

`st_geometry()` prints the new extent.

```{r}
st_geometry(preschool3414)
```
:::

The bounding box now runs from about 11,810 to 45,404 across and 25,596 to 49,301 up. Longitude only goes from minus 180 to 180 and latitude from minus 90 to 90, so these values must be metres. The transform worked.

## Overlaying the Two Layers

Before the transform, the preschool points would not have landed on the subzone map. Now they line up. `plot()` with `add = TRUE` draws the points over the boundaries.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "2,290 pre-schools over the subzone boundaries")
plot(st_geometry(preschool3414), add = TRUE,
     pch = 21, cex = 0.42, bg = "#0072B2AA", col = "#00548A", lwd = 0.3)
```

## The Code

```{r}
#| eval: false
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "2,290 pre-schools over the subzone boundaries")
plot(st_geometry(preschool3414), add = TRUE,
     pch = 21, cex = 0.42, bg = "#0072B2AA", col = "#00548A", lwd = 0.3)
```
:::

# Importing and Converting Aspatial Data

The Airbnb listings file is not geospatial data, but its latitude and longitude columns are enough to turn it into one.

## Importing the Aspatial Data

::: panel-tabset
## Importing the listings

`read_csv()` of **readr** reads the listings into a tibble.

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

## Checking the fields

`glimpse()` shows which columns hold the coordinates.

```{r}
glimpse(listings)
```

The two fields I need are `latitude` and `longitude`, both in decimal degrees. It is safe to assume WGS84.
:::

## Creating a Simple Feature Data Frame from an Aspatial Data Frame

::: panel-tabset
## Converting to sf

`st_as_sf()` builds point geometry from the two coordinate columns, and the piped `st_transform()` reprojects the result.

```{r}
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)
```

`coords` wants x first and y second, so longitude comes before latitude. `crs = 4326` is WGS84, the system of the raw coordinates. The piped `st_transform()` then moves it to 3414 to match the other layers.

## Checking the result

`glimpse()` checks the new sf object.

```{r}
glimpse(listings_sf)
```

A `geometry` column has appeared and the separate longitude and latitude columns are gone.
:::

## Plotting the Listings over the Subzones

I use the same `plot()` overlay as before, with the listings as small orange dots.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "Airbnb listings cluster in the centre and south")
plot(st_geometry(listings_sf), add = TRUE,
     pch = 20, cex = 0.3, col = "#D55E0099")
```

## The Code

```{r}
#| eval: false
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz3414),
     col = "#F7F4EC", border = "grey72", lwd = 0.4,
     main = "Airbnb listings cluster in the centre and south")
plot(st_geometry(listings_sf), add = TRUE,
     pch = 20, cex = 0.3, col = "#D55E0099")
```
:::

Most listings are in the central and southern subzones. They thin out quickly towards the west and the north coast, with a few clusters in the east near the airport. Large parts of the island have almost none.

# Geoprocessing with sf

## Land Acquisition Analysis

In this scenario the authority wants to widen the existing cycling paths and needs 5 metres of reserved land on either side. The task is to work out how much land that is.

`dist = 5` in `st_buffer()` uses the units of the layer's coordinate system. In a projected system like SVY21 that is metres. In degrees it is less clear. With the s2 engine on, which is the sf default, a buffer on geographic coordinates is computed on the sphere and `dist` is taken as metres. With s2 off, the same call reads 5 as five degrees. That is a buffer of some 550 km on every side, and it still runs without error. Working in SVY21 means the answer does not depend on that setting.

The cycling path already came in SVY21, so this transform changes nothing. I kept it so the code does not rely on LTA always shipping it that way. It costs nothing and makes the assumption clear.

::: panel-tabset
## Confirming the projection

`st_transform()` to 3414, then `st_crs()` to confirm.

```{r}
cyclingpath3414 <- st_transform(cyclingpath, crs = 3414)
st_crs(cyclingpath3414)$epsg
```

## Buffering the paths

`st_buffer()` builds the 5 metre buffers. `nQuadSegs = 30` keeps the rounded ends smooth.

```{r}
buffer_cycling <- st_buffer(cyclingpath3414, dist = 5, nQuadSegs = 30)
```

## Computing the area

`st_area()` inside `mutate()` gives the area of each buffer.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

## Adding up the total

`sum()` adds the buffer areas together.

```{r}
sum(buffer_cycling$AREA)
```
:::

To see what this looks like on the ground, I clip the buffers to one subzone, Tampines West. `filter()` picks the subzone and `st_intersection()` cuts the buffers to its boundary.

```{r}
mpsz_selected <- mpsz3414 %>%
  filter(SUBZONE_N == "TAMPINES WEST")

buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected)
```

`plot()` draws Tampines West and lays the clipped buffers over it in orange.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz_selected),
     col = "#F7F4EC", border = "grey55", lwd = 0.8,
     main = "Five metre reserve along cycling paths, Tampines West")
plot(st_geometry(buffer_cycling_selected), add = TRUE,
     col = "#D55E00", border = NA)
```

## The Code

```{r}
#| eval: false
par(mar = c(0.5, 0.5, 2, 0.5), cex.main = 1.3)
plot(st_geometry(mpsz_selected),
     col = "#F7F4EC", border = "grey55", lwd = 0.8,
     main = "Five metre reserve along cycling paths, Tampines West")
plot(st_geometry(buffer_cycling_selected), add = TRUE,
     col = "#D55E00", border = NA)
```
:::

## Counting Pre-schools by Planning Subzone

In this scenario the authority needs a pre-school count for every planning subzone for forward planning.

::: panel-tabset
## Counting the points

`st_intersects()` returns, for each subzone, the list of preschools inside it. Wrapping that in `lengths()` turns the lists into counts in one step.

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

Note that this uses `st_intersects()`, not `st_intersection()`.

::: {.callout-note}
## Learning diary
These two names are one letter apart but do different jobs. `st_intersects()` tells me which things overlap and returns a list. `st_intersection()` cuts the shapes and returns the overlapping pieces. Only the first one works for counting.
:::

## Summarising the counts

`summary()` gives the spread of the counts.

```{r}
summary(mpsz3414$`PreSch Count`)
```

The median subzone has only 4 pre-schools while the maximum is 72, so the distribution is heavily skewed.

## Finding the top subzone

`top_n()` of **dplyr** returns the subzone with the highest count.

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

## Converting to density

Raw counts are misleading because subzones differ a lot in size. So I convert them to a density per square kilometre. `st_area()` gives each subzone's area in square metres and `mutate()` divides.

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
```
:::

## Exploratory Data Analysis

The base R `hist()` is quick but hard to customise.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
dens <- as.numeric(mpsz3414$`PreSch Density`)

hist(dens,
     breaks = 20, col = "#CFE3F0", border = "white",
     main = "Pre-school density is heavily skewed",
     xlab = "Pre-schools per square kilometre",
     ylab = "Number of subzones")
abline(v = median(dens, na.rm = TRUE), col = "#D55E00", lwd = 2, lty = 2)
legend("topright", bty = "n", col = "#D55E00", lwd = 2, lty = 2,
       legend = sprintf("median %.1f", median(dens, na.rm = TRUE)))
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
dens <- as.numeric(mpsz3414$`PreSch Density`)

hist(dens,
     breaks = 20, col = "#CFE3F0", border = "white",
     main = "Pre-school density is heavily skewed",
     xlab = "Pre-schools per square kilometre",
     ylab = "Number of subzones")
abline(v = median(dens, na.rm = TRUE), col = "#D55E00", lwd = 2, lty = 2)
legend("topright", bty = "n", col = "#D55E00", lwd = 2, lty = 2,
       legend = sprintf("median %.1f", median(dens, na.rm = TRUE)))
```
:::

`geom_histogram()` of **ggplot2** gives a more presentable version, with a proper title and axis labels.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 13),
        axis.text = element_text(size = 11))
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 13),
        axis.text = element_text(size = 11))
```
:::

This matches the summary. Over a hundred subzones are in the lowest bin and the count drops quickly after that. Two subzones sit past 30 per square kilometre.

A scatterplot of count against density with `geom_point()` shows how the two relate.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  labs(title = "Pre-school count against density by subzone",
       x = "Pre-school density (per km sq)",
       y = "Pre-school count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 13),
        axis.text = element_text(size = 11))
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  labs(title = "Pre-school count against density by subzone",
       x = "Pre-school density (per km sq)",
       y = "Pre-school count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 13),
        axis.text = element_text(size = 11))
```
:::

Count and density do rise together, but the spread is wide. Two subzones with the same density can be twenty pre-schools apart. The points on the far right, around 35 per square kilometre, have only a few pre-schools each because those subzones are small.

::: {.callout-caution}
## Where I differ from the book
The book adds `xlim(0, 40)` and `ylim(0, 40)` to this plot. These drop every point past 40, and the render hides the warning. Tampines East, with 72 pre-schools, would disappear and the chart would look tidier than the data is. I left the limits off so every subzone is shown.
:::

# Preparing the Data for Thematic Mapping

The tmap half uses the Master Plan 2019 subzone boundaries and the 2024 resident population figures by planning area, subzone, age group and sex.

## Importing Data into R

::: panel-tabset
## Importing the 2019 subzone boundary

`st_read()` reads the geojson and the piped `st_transform()` moves it straight to SVY21, because this layer is also in WGS84.

```{r}
mpsz19 <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

::: {.callout-note}
## Learning diary
The 2019 file downloads as `MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson`, with GEOJSON in the middle of the name as well as in the extension. I did not expect that. Now I copy filenames from `ls` instead of typing what I think they are called.
:::

## Checking the attributes

`names()` lists the columns the layer came with.

```{r}
names(mpsz19)
```

::: {.callout-caution}
## Where I differ from the book
At this point the book uses `rvest` code to parse attributes out of an HTML table in a `Description` field. That was needed for the KML version of this layer, which data.gov.sg has since removed. The geojson that replaced it has `SUBZONE_N`, `PLN_AREA_N` and `REGION_N` as proper columns, so the join below can use them directly.
:::

Printing the object shows the first ten features and confirms the new CRS.

```{r}
mpsz19
```

It has 332 features and 13 fields, including the three name columns the join needs.

## Importing the population table

`read_csv()` reads the SingStat table of residents by planning area, subzone, age group and sex.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```
:::

## Data Wrangling

The goal is one row per subzone with six fields. `YOUNG` is ages 0 to 24, `ECONOMY ACTIVE` is 25 to 64, and `AGED` is 65 and over. `DEPENDENCY` divides the young plus the aged by the economically active.

::: panel-tabset
## Aggregating and pivoting

`group_by()` and `summarise()` add up the residents in each subzone and age group, then `pivot_wider()` of **tidyr** spreads the age groups into columns.

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP)
```

The book skips the next step, but it turned out to matter a lot. `names()` shows the order of the age columns.

```{r}
names(popdata2024)
```

The book builds the three age bands by column number. But in the list above, the age groups are not in age order.

This comes from the summarise step. `summarise()` sorts its rows by the grouping columns. `AG` is text, so the age groups are sorted as text. `pivot_wider()` then creates the columns in that sorted order. As text, `5_to_9` comes after `55_to_59`, so it sits at position 14 instead of 12.

::: {.callout-caution}
## Where I differ from the book
The book uses `rowSums(.[3:6]) + rowSums(.[12])` for `YOUNG` and similar column ranges for the other bands. With my columns, those numbers make `YOUNG` pick up the 50 to 54 band instead of 5 to 9. `ECONOMY ACTIVE` ends up counting primary school children. I select the bands by name instead.
:::

::: {.callout-warning}
## Issue I hit
Like the projection problem, this does not throw an error. It just gives a wrong dependency ratio, and every map after it looks normal.

I nearly skipped `names(popdata2024)`. That one line caught the problem.
:::

## Selecting the age bands by name

Selecting by name removes the dependency on column order. Three character vectors hold the band names.

```{r}
young_ag <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19", "20_to_24")

active_ag <- c("25_to_29", "30_to_34", "35_to_39", "40_to_44",
               "45_to_49", "50_to_54", "55_to_59", "60_to_64")

aged_ag <- c("65_to_69", "70_to_74", "75_to_79",
             "80_to_84", "85_to_89", "90_and_over")
```

I also add a guard against this problem. `stopifnot()` stops the render if a future release renames or drops a band, instead of letting a wrong number through.

```{r}
stopifnot(all(c(young_ag, active_ag, aged_ag) %in% names(popdata2024)))
```

## Deriving the dependency ratio

`rowSums()` over `across(all_of(...))` adds up each band by name, then `select()` keeps the six fields plus the keys.

```{r}
popdata2024 <- popdata2024 %>%
  mutate(
    YOUNG = rowSums(across(all_of(young_ag))),
    `ECONOMY ACTIVE` = rowSums(across(all_of(active_ag))),
    AGED = rowSums(across(all_of(aged_ag))),
    TOTAL = rowSums(across(all_of(c(young_ag, active_ag, aged_ag)))),
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

## Checking the totals

Next I check the result. The three bands should add up to the total, and the ratio for Singapore overall should be near 0.7. `summarise()` computes both.

```{r}
popdata2024 %>%
  summarise(
    young = sum(YOUNG),
    active = sum(`ECONOMY ACTIVE`),
    aged = sum(AGED),
    total = sum(TOTAL),
    bands_match_total = sum(YOUNG) + sum(`ECONOMY ACTIVE`) + sum(AGED) == sum(TOTAL),
    overall_dependency = (sum(YOUNG) + sum(AGED)) / sum(`ECONOMY ACTIVE`)
  )
```

::: {.callout-important}
## Check that caught something
The bands add to 4,189,460, which fits Singapore's resident population of a little over 4 million, and they match the total exactly. The overall ratio is 0.732, which is in the expected range.

Both had to be right before I trusted any map on this page. If I had used the book's column numbers, the problem would have shown up in this check.
:::
:::

## Joining the Attribute Data and Geospatial Data

::: panel-tabset
## Matching the case of the keys

The `PA` and `SZ` values are mixed case while `PLN_AREA_N` and `SUBZONE_N` are uppercase, so they will not match until I fix that. `mutate_at()` applies `toupper()` to both, and `filter()` drops rows with nobody of working age.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

## Joining the two

`left_join()` matches each subzone to its population row on the subzone name.

```{r}
mpsz_pop2024 <- left_join(mpsz19, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

The sf object goes on the left so the result stays an sf data frame, not a plain tibble.

## Saving the result

`write_rds()` saves the joined object so later exercises can read it back without redoing the wrangling.

```{r}
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```
:::

# Choropleth Mapping Geospatial Data Using tmap

## Plotting a Choropleth Map Quickly with qtm()

`tmap_mode("plot")` sets static output, and `qtm()` of **tmap** draws a choropleth in one line. I only added a legend title, a legend position and a figure title. The bins are the ones `qtm()` picks.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tmap_mode("plot")
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY",
    fill.legend = tm_legend(title = "Dependency ratio",
                            position = tm_pos_out("right", "center"))) +
  tm_title("qtm() with its default bins", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tmap_mode("plot")
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY",
    fill.legend = tm_legend(title = "Dependency ratio",
                            position = tm_pos_out("right", "center"))) +
  tm_title("qtm() with its default bins", size = 1.5, fontface = "bold")
```
:::

`qtm()` is quick, but the default bins are a problem. The legend runs 0 to 10 in steps of two. Almost every subzone is under 1, so the island lands in the palest class with one dark polygon near Changi. The map is not wrong, but it shows very little.

::: {.callout-note}
## Learning diary
One subzone with a ratio of 10 is enough to spoil the default bins for the other 237 subzones that have a ratio. The custom breaks section below deals with this.

It is also why I stop using `qtm()` here. To change the bins, the palette or the legend I have to pass `tm_polygons()` arguments through it anyway, as I did with the legend above. So from here on I write `tm_shape()` and `tm_polygons()` in full.
:::

## Drawing a Base Map

`tm_shape()` sets the data, then layer functions draw on top of it. `tm_polygons()` with no variable just draws the subzones.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons() +
  tm_title("The 2019 subzones as a plain base map", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons() +
  tm_title("The 2019 subzones as a plain base map", size = 1.5, fontface = "bold")
```
:::

## Drawing a Choropleth Map Using tm_polygons()

Naming a variable in `tm_polygons()` turns the base map into a choropleth.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_title("tm_polygons() with its default classes", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_title("tm_polygons() with its default classes", size = 1.5, fontface = "bold")
```
:::

The default binning is "pretty" and the default palette is blues3. Missing values come out grey.

## Drawing a Choropleth Map Using tm_fill() and tm_borders()

`tm_polygons()` is really a wrapper around these two. `tm_fill()` alone colours the subzones with no outlines.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_title("tm_fill() alone, no borders", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_title("tm_fill() alone, no borders", size = 1.5, fontface = "bold")
```
:::

Adding `tm_borders()` back gives the same map as the `tm_polygons()` one above. The two routes draw the same thing.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_borders() +
  tm_title("tm_fill() plus tm_borders()", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_borders() +
  tm_title("tm_fill() plus tm_borders()", size = 1.5, fontface = "bold")
```
:::

`tm_borders()` takes `col`, `lwd` and `lty` for colour, line width and line type. Only the borders change from the map above, now grey and dashed. I set `lwd = 0.8` because at 0.1 the dashes were too thin to see.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_borders(col = "grey60", lwd = 0.8, lty = "dashed") +
  tm_title("Dashed grey borders from tm_borders()", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY",
          fill.legend = tm_legend(title = "Dependency ratio",
                  position = tm_pos_out("right", "center"))) +
  tm_borders(col = "grey60", lwd = 0.8, lty = "dashed") +
  tm_title("Dashed grey borders from tm_borders()", size = 1.5, fontface = "bold")
```
:::

# Data Classification Methods of tmap

tmap offers ten classification styles. The choice matters, because the same data can look very different depending on where the breaks fall.

## Quantile Classification

`tm_scale_intervals()` with `style = "quantile"` puts the same number of subzones in each of the five classes.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Quantile, equal counts per class", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Quantile, equal counts per class", size = 1.5, fontface = "bold")
```
:::

## Equal Interval Classification

`style = "equal"` cuts the range of values into five classes of the same width instead.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Equal interval, equal widths per class", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Equal interval, equal widths per class", size = 1.5, fontface = "bold")
```
:::

The two maps use the same column but look very different.

Quantile puts the same number of subzones in each class, so the breaks come out at 0.64, 0.70, 0.74 and 0.82. These all fall where most of the data is, and the map clearly separates the central and eastern subzones.

Equal interval cuts 0 to 10 into five equal parts. Nearly everything falls in the first one, so the island goes pale with a single dark polygon. From that map alone I would think dependency hardly varies in Singapore, which is not true.

Both maps show the same numbers correctly. But one subzone at 10 makes the equal interval version almost useless. I only saw this by putting the two side by side.

## Plotting a Choropleth Map with Custom Breaks

I look at the numbers before choosing break points. `summary()` gives the quartiles and the extremes.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

The middle half of the subzones sits between 0.675 and 0.814, so breaks at 0.60, 0.70, 0.80 and 0.90 separate most of the data. But the maximum is 10, so I also need a top class above 1.00 for those subzones. `breaks` in `tm_scale_intervals()` takes n+1 values, including a minimum and maximum, so six classes need seven values.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00, 10)),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Custom breaks read from the summary", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00, 10)),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Custom breaks read from the summary", size = 1.5, fontface = "bold")
```
:::

::: {.callout-caution}
## Where I differ from the book
The book's breaks are `c(0, 0.60, 0.70, 0.80, 0.90, 1.00)` and stop at 1.00, but the summary shows a maximum of 10. Subzones above 1.00 fall outside the classes, but in my first render tmap still painted them in the darkest class. So the legend said the top class ended at 1.00 while it held a subzone at 10. I added a seventh break at 10 so the top class, 1.00 to 10, shows what it holds.
:::

# Colour Scheme

## Using ColorBrewer Palettes

`values` in `tm_scale_intervals()` picks the palette. This map reuses the quantile classes from above and only swaps the default blues for the ColorBrewer oranges.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.oranges"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("The quantile classes in brewer.oranges", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.oranges"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("The quantile classes in brewer.oranges", size = 1.5, fontface = "bold")
```
:::

A minus prefix reverses the ramp. Dark now means a low ratio, which feels backwards. It is only useful when the low values are the focus.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.oranges"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("The same ramp reversed, dark now means low", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.oranges"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("The same ramp reversed, dark now means low", size = 1.5, fontface = "bold")
```
:::

::: {.callout-caution}
## Where I differ from the book
The book shows palettes with `"brewer.greens"` and `"-brewer.greens"`, and uses the reversed greens again for the map style. I avoid green as a data colour because red and green are the pair most colour-blind readers struggle with. I use the blues default and `"brewer.oranges"` for the same two ideas, and `"brewer.or_rd"` for the style map.
:::

## Cartographic Furniture

`tm_compass()`, `tm_scalebar()`, `tm_grid()` and `tm_credits()` turn a plot into a proper map. I moved the credits to the top left, where there is only sea, because at the bottom left they covered the southern islands.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = tm_pos_in("left", "top")) +
  tm_title("Compass, scale bar, grid and credits", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = tm_pos_in("left", "top")) +
  tm_title("Compass, scale bar, grid and credits", size = 1.5, fontface = "bold")
```
:::

# Map Layout

## Map Legend

`tm_legend()` inside `fill.legend` sets the legend's title, position and text size. The book puts it inside the frame with `tm_pos_auto_in()`. I keep it outside on the right with `tm_pos_out()`, as on every map here, so it never covers a subzone.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"),
                                      title.size = 1.1,
                                      text.size = 0.8)) +
  tm_borders(col_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = tm_pos_in("left", "top")) +
  tm_title("Legend moved outside the frame", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"),
                                      title.size = 1.1,
                                      text.size = 0.8)) +
  tm_borders(col_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = tm_pos_in("left", "top")) +
  tm_title("Legend moved outside the frame", size = 1.5, fontface = "bold")
```
:::

## Map Style

Adding `tm_style()` to the chain applies a preset style to this one map. "natural" gives the sea a blue background.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.or_rd"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_style("natural") +
  tm_title("The natural style", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.or_rd"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_style("natural") +
  tm_title("The natural style", size = 1.5, fontface = "bold")
```
:::

`tm_style()` only styles the map it is added to, so nothing carries over. The book calls `tmap_style()` instead, which changes the style for every later map, and then resets it. I keep the reset with `tmap_style("white")` as a safeguard, and set the legend sizes again in case the style change reset them.

```{r}
tmap_style("white")
tmap_options(legend.title.size = 1.1, legend.text.size = 0.8)
```

# Drawing Small Multiple Choropleth Maps

Facet maps put several maps in one figure. This makes it easy to see how a spatial pattern changes with another variable.

## By Assigning Multiple Values to an Aesthetic

Giving `fill` two column names makes tmap draw one map per variable. `fill.free = FALSE` puts both on one shared equal interval scale, so the same shade means the same number of residents in both panels. `tm_facets_vstack()` stacks the two maps one above the other, so each one gets the full width of the page.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 8.5
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.free = FALSE,
              fill.legend = tm_legend(title = "Residents (count)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_facets_vstack() +
  tm_borders(col_alpha = 0.5) +
  tm_title("Young and aged residents on one shared scale", size = 1.3, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.free = FALSE,
              fill.legend = tm_legend(title = "Residents (count)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_facets_vstack() +
  tm_borders(col_alpha = 0.5) +
  tm_title("Young and aged residents on one shared scale", size = 1.3, fontface = "bold")
```
:::

## By Arranging Multiple Maps in a Grid

`tmap_arrange()` puts separately built maps in one figure. With `ncol = 1` they stack, so each map gets the full width. Each map here has its own quantile classes, so I compare the patterns and not the shades.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(title = "Residents 0 to 24",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Where the young live, as a raw count", size = 1.3, fontface = "bold")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(title = "Residents 65 and over",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Where the aged live, as a raw count", size = 1.3, fontface = "bold")

tmap_arrange(youngmap, agedmap, ncol = 1)
```

## The Code

```{r}
#| eval: false
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(title = "Residents 0 to 24",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Where the young live, as a raw count", size = 1.3, fontface = "bold")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(title = "Residents 65 and over",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Where the aged live, as a raw count", size = 1.3, fontface = "bold")

tmap_arrange(youngmap, agedmap, ncol = 1)
```
:::

The two maps look almost the same. I expected the aged map to show the older estates and the young map to show the newer towns, but the same subzones are dark on both.

This is because both are raw counts. A subzone with more people has more of every age group. So these maps mostly show where people live, not where the population is young or old.

::: {.callout-tip}
## Alternative explored
The book stops at the count maps. They did not answer my question, so I redrew both as shares of each subzone's own population.
:::

Dividing each band by the subzone total with `mutate()` removes the effect of population size.

```{r}
mpsz_pop2024 <- mpsz_pop2024 %>%
  mutate(`YOUNG PCT` = YOUNG / TOTAL * 100,
         `AGED PCT` = AGED / TOTAL * 100)
```

I use the same `tmap_arrange()` layout for the two shares, again with each map on its own quantile classes.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9
youngpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG PCT",
              fill.legend = tm_legend(title = "Share 0 to 24 (%)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Young as a share of the subzone", size = 1.3, fontface = "bold")

agedpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED PCT",
              fill.legend = tm_legend(title = "Share 65 and over (%)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Aged as a share of the subzone", size = 1.3, fontface = "bold")

tmap_arrange(youngpct, agedpct, ncol = 1)
```

## The Code

```{r}
#| eval: false
youngpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG PCT",
              fill.legend = tm_legend(title = "Share 0 to 24 (%)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Young as a share of the subzone", size = 1.3, fontface = "bold")

agedpct <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED PCT",
              fill.legend = tm_legend(title = "Share 65 and over (%)",
                                      position = tm_pos_out("right", "center")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Aged as a share of the subzone", size = 1.3, fontface = "bold")

tmap_arrange(youngpct, agedpct, ncol = 1)
```
:::

Now the two maps look different, which is what I expected the first pair to show. Comparing them with the count maps above shows how much of the first pattern was just population size.

## By Defining a Group-by Variable in tm_facets()

`tm_facets()` with `by = "REGION_N"` draws one panel per region. `free.coords = TRUE` zooms each panel to its own region, while all five share one set of quantile classes and one legend.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile",
                                          values = "brewer.blues"),
          fill.legend = tm_legend(title = "Dependency ratio",
                                  position = tm_pos_out("right", "center"))) +
  tm_facets(by = "REGION_N", nrow = 2, ncol = 3,
            free.coords = TRUE, drop.units = TRUE) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Dependency by region, each panel on its own extent", size = 1.3, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile",
                                          values = "brewer.blues"),
          fill.legend = tm_legend(title = "Dependency ratio",
                                  position = tm_pos_out("right", "center"))) +
  tm_facets(by = "REGION_N", nrow = 2, ncol = 3,
            free.coords = TRUE, drop.units = TRUE) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Dependency by region, each panel on its own extent", size = 1.3, fontface = "bold")
```
:::

# Mapping Spatial Object Meeting a Selection Criterion

Instead of facetting, I can `filter()` down to the region of interest before `tm_shape()`.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Dependency ratio in the Central Region", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Dependency ratio in the Central Region", size = 1.5, fontface = "bold")
```
:::

# Complementing Thematic Map with Statistical Chart

`fill.chart = tm_chart_box()` adds a boxplot of the mapped values next to the legend, so the map and the chart are in one figure.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7.5
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center")),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_title("Central Region with a boxplot of the same values", size = 1.5, fontface = "bold") +
  tm_layout(asp = 0.8)
```

## The Code

```{r}
#| eval: false
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center")),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_title("Central Region with a boxplot of the same values", size = 1.5, fontface = "bold") +
  tm_layout(asp = 0.8)
```
:::

The boxplot shows something the map does not. All the outliers are below the box, between 0 and about 0.3, not above it. These are subzones with working age residents and hardly any children or elderly.

`boxplot.stats()` returns the same outliers the boxplot draws. So I can outline those subzones in red and label them with `tm_text()`. The labels show they are bunched in the downtown core. `remove_overlap = TRUE` drops labels that would collide, so a few outlined subzones may have no name.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7.5
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center")),
              fill.chart = tm_chart_box()) +
  tm_borders(col_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "black", size = 0.6,
          options = opt_tm_text(remove_overlap = TRUE)) +
  tm_title("Central Region outliers outlined in red", size = 1.5, fontface = "bold") +
  tm_layout(asp = 0.8)
```

## The Code

```{r}
#| eval: false
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio",
                                      position = tm_pos_out("right", "center")),
              fill.chart = tm_chart_box()) +
  tm_borders(col_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "black", size = 0.6,
          options = opt_tm_text(remove_overlap = TRUE)) +
  tm_title("Central Region outliers outlined in red", size = 1.5, fontface = "bold") +
  tm_layout(asp = 0.8)
```
:::

# Creating Interactive Map

Switching `tmap_mode()` to view turns the same code into a map I can zoom and click. `tm_view()` sets its zoom limits.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          fill.legend = tm_legend(title = "Dependency ratio",
                                  position = tm_pos_in("left", "bottom")),
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

## The Code

```{r}
#| eval: false
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          fill.legend = tm_legend(title = "Dependency ratio",
                                  position = tm_pos_in("left", "bottom")),
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```
:::

`set_zoom_limits` stops users from zooming out until they are lost over the ocean. It keeps them inside the study area.

`tmap_mode("plot")` switches back to static maps for the rest of the page.

```{r}
tmap_mode("plot")
```

# Beyond the Exercise

The two chapters end at the choropleth. After that, three things bothered me, so I looked into them.

I wanted to know whether any of this is actually spatial and whether the pattern holds when the units get bigger. I also wanted to know whether the map itself misleads me about who lives where.

## Shuffling the Values to Test Whether the Map Is Spatial

Anselin's lecture gives a test for this. If moving the values around changes nothing, the analysis was not spatial. If it does change something, it was.

This is easy to run on my own data. `sample()` only reshuffles which subzone holds which number, so every value is kept. `set.seed()` makes the shuffle repeatable.

```{r}
set.seed(1234)
mpsz_shuffled <- mpsz_pop2024
mpsz_shuffled$DEPENDENCY <- sample(mpsz_pop2024$DEPENDENCY)
```

If the values are the same, any summary that ignores location should be identical. A small `tibble()` puts the two side by side.

```{r}
tibble(
  version  = c("Real Singapore", "Same numbers, shuffled"),
  subzones = c(sum(!is.na(mpsz_pop2024$DEPENDENCY)),
               sum(!is.na(mpsz_shuffled$DEPENDENCY))),
  mean     = c(mean(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               mean(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  sd       = c(sd(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               sd(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  median   = c(median(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),
               median(mpsz_shuffled$DEPENDENCY, na.rm = TRUE))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

I do the same for the histogram, drawn with `geom_histogram()` and `facet_wrap()`.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5.5
bind_rows(
  tibble(version = "Real Singapore",         value = mpsz_pop2024$DEPENDENCY),
  tibble(version = "Same numbers, shuffled", value = mpsz_shuffled$DEPENDENCY)
) %>%
  filter(!is.na(value), value < 2) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white", linewidth = 0.2) +
  facet_wrap(~version) +
  labs(title = "The same histogram twice",
       subtitle = "Values under 2 only, so the single subzone at 10 does not flatten the axis",
       x = "Dependency ratio", y = "Subzones") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        panel.spacing = grid::unit(2, "lines"))
```

## The Code

```{r}
#| eval: false
bind_rows(
  tibble(version = "Real Singapore",         value = mpsz_pop2024$DEPENDENCY),
  tibble(version = "Same numbers, shuffled", value = mpsz_shuffled$DEPENDENCY)
) %>%
  filter(!is.na(value), value < 2) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white", linewidth = 0.2) +
  facet_wrap(~version) +
  labs(title = "The same histogram twice",
       subtitle = "Values under 2 only, so the single subzone at 10 does not flatten the axis",
       x = "Dependency ratio", y = "Subzones") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        panel.spacing = grid::unit(2, "lines"))
```
:::

A small function builds the same quantile map for either version, and `tmap_arrange()` stacks them. The two legends are identical, because the values and so the classes are the same.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 9
shuffle_map <- function(dat, ttl, show_legend = FALSE) {
  tm_shape(dat) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(title = "Dependency ratio",
                                        show = show_legend,
                                        position = tm_pos_out("right", "center"))) +
    tm_borders(col_alpha = 0.4) +
    tm_title(ttl, size = 1.3, fontface = "bold")
}

tmap_arrange(shuffle_map(mpsz_pop2024,  "Real Singapore",
                         show_legend = TRUE),
             shuffle_map(mpsz_shuffled, "Same numbers, shuffled",
                         show_legend = TRUE),
             ncol = 1)
```

## The Code

```{r}
#| eval: false
shuffle_map <- function(dat, ttl, show_legend = FALSE) {
  tm_shape(dat) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(title = "Dependency ratio",
                                        show = show_legend,
                                        position = tm_pos_out("right", "center"))) +
    tm_borders(col_alpha = 0.4) +
    tm_title(ttl, size = 1.3, fontface = "bold")
}

tmap_arrange(shuffle_map(mpsz_pop2024,  "Real Singapore",
                         show_legend = TRUE),
             shuffle_map(mpsz_shuffled, "Same numbers, shuffled",
                         show_legend = TRUE),
             ncol = 1)
```
:::

::: {.callout-tip}
## Alternative explored
Shuffling the values across subzones is my own addition. The mean, the spread, the median and the histogram are all the same. But the two maps look very different.

So none of the statistics I know can see the difference between the two maps. The rest of this course covers that gap. Lesson 4 defines what neighbouring means and Lesson 5 turns this difference into a single number I can test.

It also answers my early question of why not just use pandas. A non-spatial toolkit would say these two datasets are identical, because they are the same on everything it can measure.
:::

## Testing the Pattern at a Coarser Unit

This is the Modifiable Areal Unit Problem. Results change when the size or shape of the units changes.

Singapore's subzones sit inside planning areas, so I can test the size half of it directly. `group_by()` and `summarise()` add the bands up to planning area level, and the ratio is recomputed from the sums.

```{r}
mpsz_pa <- mpsz_pop2024 %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL)) %>%
  group_by(PLN_AREA_N) %>%
  summarise(YOUNG  = sum(YOUNG),
            ACTIVE = sum(`ECONOMY ACTIVE`),
            AGED   = sum(AGED),
            n_subzones = n(),
            .groups = "drop") %>%
  mutate(DEPENDENCY_PA = (YOUNG + AGED) / ACTIVE) %>%
  filter(is.finite(DEPENDENCY_PA))
```

A `tibble()` compares the spread at the two levels.

```{r}
tibble(
  unit   = c("Subzone", "Planning area"),
  n      = c(sum(!is.na(mpsz_pop2024$DEPENDENCY)), nrow(mpsz_pa)),
  min    = c(min(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),    min(mpsz_pa$DEPENDENCY_PA)),
  median = c(median(mpsz_pop2024$DEPENDENCY, na.rm = TRUE), median(mpsz_pa$DEPENDENCY_PA)),
  max    = c(max(mpsz_pop2024$DEPENDENCY, na.rm = TRUE),    max(mpsz_pa$DEPENDENCY_PA))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))
```

`geom_density()` overlays the two distributions.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5.5
bind_rows(
  tibble(unit = "Subzone",
         value = mpsz_pop2024$DEPENDENCY[!is.na(mpsz_pop2024$DEPENDENCY)]),
  tibble(unit = "Planning area", value = mpsz_pa$DEPENDENCY_PA)
) %>%
  ggplot(aes(x = value, fill = unit)) +
  geom_density(alpha = 0.45) +
  scale_fill_manual(values = c("Planning area" = "#D55E00",
                               "Subzone" = "#0072B2")) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Aggregating flattens the spread",
       subtitle = "Zoomed to 0 to 1.5. The density still uses every value, including the outlier",
       x = "Dependency ratio", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        legend.position = "top")
```

## The Code

```{r}
#| eval: false
bind_rows(
  tibble(unit = "Subzone",
         value = mpsz_pop2024$DEPENDENCY[!is.na(mpsz_pop2024$DEPENDENCY)]),
  tibble(unit = "Planning area", value = mpsz_pa$DEPENDENCY_PA)
) %>%
  ggplot(aes(x = value, fill = unit)) +
  geom_density(alpha = 0.45) +
  scale_fill_manual(values = c("Planning area" = "#D55E00",
                               "Subzone" = "#0072B2")) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Aggregating flattens the spread",
       subtitle = "Zoomed to 0 to 1.5. The density still uses every value, including the outlier",
       x = "Dependency ratio", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 13),
        legend.position = "top")
```
:::

::: {.callout-tip}
## Alternative explored
Re-aggregating to planning areas is not in the book. The planning area curve is taller and narrower. The people and the formula are the same, but the extremes are gone.

A map at planning area level would make Singapore look far more even than it is. Neither unit is the right one. I can only state which one I used and note that the answer depends on it.

I could only test the size half of MAUP. Results also change when boundaries are redrawn at the same size, but I have no other zoning to test that.
:::

## Measuring How Much the Map Weights Land over People

A choropleth colours areas, and people read a larger area as a larger amount. So a big empty subzone stands out more than a small crowded one.

This can be measured. `st_area()` gives each subzone's land area, and `summarise()` compares the share of map area with the share of residents in the least populated half of subzones.

```{r}
mpsz_pop2024 %>%
  filter(!is.na(TOTAL)) %>%
  mutate(area_km2 = as.numeric(st_area(.)) / 1e6) %>%
  st_drop_geometry() %>%
  summarise(
    `Map area held by the least populated half of subzones (%)` =
      round(sum(area_km2[TOTAL <= median(TOTAL)]) / sum(area_km2) * 100, 1),
    `Population those same subzones hold (%)` =
      round(sum(TOTAL[TOTAL <= median(TOTAL)]) / sum(TOTAL) * 100, 1)
  ) %>%
  pivot_longer(everything(), names_to = "measure", values_to = "value")
```

::: {.callout-tip}
## Alternative explored
Comparing the map's area with its population is my own check. The two numbers show that a reader of this map is weighting land, not people.

This is like the young and aged count maps above, from a different angle. There I mapped counts when I wanted rates. Here I map rates, but the map itself is weighted by area.
:::

# Limitations

These are the limits of this exercise.

- The sf half uses the 2014 Master Plan boundaries and the tmap half uses 2019, as the chapters specify. Subzone definitions changed between them, so the pre-school counts and dependency ratios are not on quite the same geography.
- 94 subzones have no dependency ratio. Either nobody lives there, or the table has no working-age residents for them after the `ECONOMY ACTIVE > 0` filter. They are grey on every map. They are dropped, not shown as zero.
- The dependency ratio treats 25 to 64 as working age. This is only a convention, and it is changing over time.
- Point in polygon counting ignores catchments. A pre-school just over a subzone boundary still serves children in the next subzone.
- Only the scale half of MAUP is tested, as noted above.

::: {.callout-warning}
## Issue I hit
`na.rm = TRUE` does not remove `Inf`. I assumed it did.

If a subzone has residents but nobody of working age, the dependency ratio divides by zero and gives `Inf`. Then `mean(x, na.rm = TRUE)` returns `Inf` for the whole column. `is.finite()` is the filter that catches it.

My data avoids this because I filtered on `ECONOMY ACTIVE > 0` before the join. But that was luck, not a plan.
:::

# Reflections

These are four things I learned from this exercise.

**Most of my mistakes were small file and naming issues.** The cycling path zip unpacked into a subfolder, and the 2019 file had GEOJSON in the middle of its name. `st_intersects` and `st_intersection` are one letter apart but do different jobs. These cost time, and all came from not checking what was on disk before writing code.

**The projection and age band problems gave no error.** The code ran and the maps drew, but the numbers would have been wrong. I caught both by checking output I could have skipped, the bounding box in one case and `names()` in the other.

**I misread the young and aged count maps.** I expected them to show different parts of the island, but they look almost the same. Both are counts, which mostly follow how many people live in a place. I had to redraw them as rates to answer the question I thought I was already asking.

**Quantile and equal interval gave very different maps.** On the same column, they say opposite things about whether dependency varies across Singapore. One subzone at 10 flattens the equal interval version completely. So choosing a classification is part of the analysis.
