---
title: "Hands-on Exercise 1"
subtitle: "Geospatial Data Wrangling and Thematic Mapping"
author: "ZeXin Chen"
date: "2026-08-26"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Getting Started

## Installing and Loading R Packages

Three packages are used in this exercise. **sf** handles the geospatial data, **tmap** draws the thematic maps, and **tidyverse** covers importing and wrangling.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

A quick version check, since the tmap syntax used later only exists in version 4.

```{r}
packageVersion("tmap")
```

# Geospatial Data Wrangling with sf

## Importing Geospatial Data

### Polygon feature data

data.gov.sg has retired the shapefile and KML versions of this layer, so the geojson is what is available. For geojson `st_read()` takes a single path rather than the `dsn` and `layer` pair the shapefile version needs.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

The output tells us this is a multipolygon layer with 323 features and 13 fields. Two things differ from what the reference book reports. The layer arrives in **WGS84**, not SVY21, and the field names for area and length are `SHAPE_1.AREA` and `SHAPE_1.LEN`. Both matter later.

### Polyline feature data

```{r}
cyclingpath <- st_read(dsn = "data/geospatial",
                       layer = "CyclingPathGazette")
```

LTA ships this one as a shapefile rather than geojson, so `st_read()` takes the `dsn` and `layer` pair instead of a single path. Note the CRS it reports, LTA data usually arrives already in SVY21 unlike the data.gov.sg layers above.

### Point feature data

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

2290 point features, in WGS84.

```{r}
names(preschool)
```

Worth pausing on this. The preschool layer has no real attribute columns. Everything the file knows about each centre is packed into a single `Description` string as an HTML table, a quirk of how data.gov.sg exports these layers. Oddly the two Master Plan boundary layers do not have this problem, they come through with proper columns, so it is not consistent across data.gov.sg exports. Here it does not matter either way, because every use of `preschool` below relies only on the geometry.

## Checking the Content of a Simple Feature Data Frame

Before wrangling anything it is worth looking at what we actually have.

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

This prints only the geometry column, so we see the feature type, extent and coordinate system without the attribute table getting in the way.

### Working with glimpse()

```{r}
glimpse(mpsz)
```

`glimpse()` shows the data type of every field. Note the geometry column and the two shape measurement fields, which in this geojson are named `SHAPE_1.AREA` and `SHAPE_1.LEN` rather than the `SHAPE_Area` and `SHAPE_Leng` the book refers to. The `X_ADDR` and `Y_ADDR` centroid fields present in the old shapefile are absent here.

### Working with head()

```{r}
head(mpsz, n = 5)
```

The `n` argument controls how many records are shown, which keeps the output readable.

## Plotting the Geospatial Data

The default plot of an sf object draws every attribute as its own small map.

```{r}
plot(mpsz)
```

Passing only the geometry gives the outline on its own.

```{r}
plot(st_geometry(mpsz))
```

Or a single named attribute.

```{r}
plot(mpsz["PLN_AREA_N"])
```

`plot()` is fine for a quick look. Anything meant for presentation should go through tmap instead, which is the whole of the second half of this exercise.

## Working with Projection

Two layers can only be used together if they share a coordinate system, so this step is not optional.

### Checking what we actually have

```{r}
st_crs(mpsz)
```

This is the point where my file and the reference book diverge, and the difference is worth being careful about.

The book imports a shapefile that really is in SVY21 but carries a wrong EPSG label of 9001, so it fixes the label with `st_set_crs(3414)`. That is the right tool for a mislabelling, because the coordinates themselves are already correct and only the metadata is wrong.

My geojson is a different situation. It reports WGS84, and it means it. The bounding box runs 103.6 to 104.0 on x and 1.15 to 1.47 on y, which are decimal degrees, not the metre values SVY21 would give. So the label is accurate and the coordinates are the thing that needs changing.

Using `st_set_crs(3414)` here would be a real error rather than a cosmetic one. It would attach an SVY21 label to degree values without touching them, and every area, distance and buffer computed afterwards would be silently wrong. `st_transform()` is what actually reprojects the coordinates.

```{r}
mpsz3414 <- st_transform(mpsz, crs = 3414)
```

```{r}
st_crs(mpsz3414)$epsg
```

```{r}
st_bbox(mpsz3414)
```

The bounding box is now in metres and in the tens of thousands, which is the quick check that the transform did something real rather than just relabelling.

### Transforming preschool to svy21

Same situation, same tool.

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

```{r}
st_geometry(preschool3414)
```

Values well outside the 0 to 360 range of decimal degrees, so this one worked too.

### Overlaying the two layers

Before the transform, the preschool points would not have landed on the subzone map at all. Now they do.

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz3414))
plot(st_geometry(preschool3414), add = TRUE)
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz3414))
plot(st_geometry(preschool3414), add = TRUE)
```
:::

## Importing and Converting Aspatial Data

The Airbnb listings file is not geospatial data, but it carries latitude and longitude columns, which is enough to turn it into one.

### Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
glimpse(listings)
```

The two fields we need are `latitude` and `longitude`, both in decimal degrees. The safe assumption is WGS84.

### Creating a simple feature data frame

```{r}
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)
```

Three things worth noting about the arguments. `coords` wants x first then y, so longitude comes before latitude. `crs = 4326` is WGS84, the system the raw coordinates are in. The piped `st_transform()` then moves it to 3414 to match everything else.

```{r}
glimpse(listings_sf)
```

A `geometry` column has appeared and the separate longitude and latitude columns are gone.

### Plotting the listings over the subzones

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz3414))
plot(st_geometry(listings_sf), add = TRUE, col = "red", pch = 20, cex = 0.3)
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz3414))
plot(st_geometry(listings_sf), add = TRUE, col = "red", pch = 20, cex = 0.3)
```
:::

Most listings sit in the central and southern subzones. They thin out quickly towards the west and the north coast, with a few clusters in the east near the airport. Big parts of the island have almost none.

## Geoprocessing with sf

### Use case 1, land acquisition analysis

The scenario is that the authority wants to widen the existing cycling paths and needs 5 metres of reserved land on either side. The task is to work out how much land that comes to.

`dist = 5` in `st_buffer()` means five units of whatever coordinate system the layer is in. On a WGS84 layer that would be five degrees, a buffer some 550 km wide, and it would run without error and give a meaningless answer.

The cycling path is the one layer here that already arrived in SVY21, so this transform changes nothing. I have kept it anyway rather than relying on LTA continuing to ship it that way. It costs nothing and it makes the assumption explicit.

```{r}
cyclingpath3414 <- st_transform(cyclingpath, crs = 3414)
st_crs(cyclingpath3414)$epsg
```

Now `st_buffer()` builds buffers that really are 5 metres.

```{r}
buffer_cycling <- st_buffer(cyclingpath3414, dist = 5, nQuadSegs = 30)
```

Then the area of each buffer.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

And the total.

```{r}
sum(buffer_cycling$AREA)
```

To see what this looks like on the ground, we can clip the buffers to a single subzone. Tampines West here.

```{r}
mpsz_selected <- mpsz3414 %>%
  filter(SUBZONE_N == "TAMPINES WEST")

buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected)
```

::: panel-tabset
## The Plot

```{r}
#| echo: false
plot(st_geometry(mpsz_selected))
plot(st_geometry(buffer_cycling_selected), add = TRUE, col = "red")
```

## The Code

```{r}
#| eval: false
plot(st_geometry(mpsz_selected))
plot(st_geometry(buffer_cycling_selected), add = TRUE, col = "red")
```
:::

### Use case 2, counting pre-schools by planning subzone

The scenario is that the authority needs a pre-school count for every planning subzone to feed into forward planning.

`st_intersects()` returns, for each subzone, the list of preschools falling inside it. Wrapping that in `lengths()` turns the lists into counts in one step.

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

Note that this is `st_intersects()`, not `st_intersection()`. The two names are close but they do very different things.

```{r}
summary(mpsz3414$`PreSch Count`)
```

The median subzone has only a handful of pre-schools while the maximum is far higher, so the distribution is heavily skewed.

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

Counts on their own are misleading because subzones differ a lot in size, so we convert to a density per square kilometre.

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
```

### Exploratory data analysis

The base R histogram is quick but hard to customise.

```{r}
hist(mpsz3414$`PreSch Density`)
```

ggplot2 gives a far more presentable version, with room for a proper title and axis labels.

::: panel-tabset
## The Plot

```{r}
#| echo: false
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "Most subzones have very few pre-schools, while a small number have a great many",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```
:::

Same story as the summary. Over a hundred subzones are in the lowest bin and the count drops off fast after that, with a few stragglers past 30 per square kilometre.

A scatterplot of count against density shows how the two relate.

::: panel-tabset
## The Plot

```{r}
#| echo: false
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```

## The Code

```{r}
#| eval: false
ggplot(data = mpsz3414,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```
:::

Count and density do rise together, but the spread is wide. Two subzones with the same density can be twenty pre-schools apart. The points on the far right, around 35 per square kilometre, only have a few pre-schools each because those subzones are small.

One thing to watch. `xlim` and `ylim` cut the axes at 40, so any point past that just disappears instead of showing up as an outlier.

# Thematic Mapping and Geovisualisation with tmap

## Importing the Data

This half uses two different datasets, the Master Plan 2019 subzone boundaries and the 2024 resident population figures.

### Importing the geospatial data

```{r}
mpsz19 <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

### Checking the attributes

```{r}
names(mpsz19)
```

The reference book at this point runs a chunk of `rvest` code to parse attributes out of an HTML table buried in a `Description` field. That was necessary for the KML version of this layer, which data.gov.sg has since unpublished. The geojson that replaced it carries `SUBZONE_N`, `PLN_AREA_N` and `REGION_N` as proper columns, so there is nothing to parse and the join below can use them directly.

Worth noting the geojson is also in WGS84, which is why the `st_transform()` is piped onto the import above.

```{r}
mpsz19
```

332 features and 13 fields, with the three name columns the join needs.

### Importing the attribute data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

## Data Preparation

The target is one row per subzone with six derived fields. `YOUNG` covers ages 0 to 24, `ECONOMY ACTIVE` covers 25 to 64, `AGED` covers 65 and over, and `DEPENDENCY` is the ratio of the first and third against the second.

First the aggregate and pivot.

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP)
```

Now a step the reference book does not take, and which turns out to matter a great deal.

```{r}
names(popdata2024)
```

The book computes the three age bands using hardcoded column positions, `rowSums(.[3:6]) + rowSums(.[12])` for `YOUNG` and so on. That only works if the age groups land in a particular order. They do not. `pivot_wider()` names the new columns after the values it found, and those get sorted as text, so `5_to_9` sorts after `55_to_59` and ends up at position 14 rather than 12.

Run the book's indices against this and `YOUNG` picks up the 50 to 54 band instead of 5 to 9, while `ECONOMY ACTIVE` quietly absorbs primary school children. Nothing errors. The dependency ratio just comes out wrong, and every map built on it after that is wrong in a way no amount of staring at the map would reveal.

Selecting by name instead removes the dependency on ordering entirely.

```{r}
young_ag <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19", "20_to_24")

active_ag <- c("25_to_29", "30_to_34", "35_to_39", "40_to_44",
               "45_to_49", "50_to_54", "55_to_59", "60_to_64")

aged_ag <- c("65_to_69", "70_to_74", "75_to_79",
             "80_to_84", "85_to_89", "90_and_over")
```

A guard against the same class of problem returning. If a future release renames or adds a band, this stops the render rather than letting a wrong number through.

```{r}
stopifnot(all(c(young_ag, active_ag, aged_ag) %in% names(popdata2024)))
```

```{r}
popdata2024 <- popdata2024 %>%
  mutate(
    YOUNG = rowSums(across(all_of(young_ag))),
    `ECONOMY ACTIVE` = rowSums(across(all_of(active_ag))),
    AGED = rowSums(across(all_of(aged_ag))),
    TOTAL = rowSums(across(all_of(c(young_ag, active_ag, aged_ag)))),
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

A sanity check on the result. The three bands should sum to the total, and the dependency ratio for Singapore as a whole should land somewhere near 0.7.

```{r}
popdata2024 %>%
  summarise(
    young = sum(YOUNG),
    active = sum(`ECONOMY ACTIVE`),
    aged = sum(AGED),
    total = sum(TOTAL),
    bands_match_total = sum(YOUNG) + sum(`ECONOMY ACTIVE`) + sum(AGED) == sum(TOTAL),
    overall_dependency = (sum(YOUNG) + sum(AGED)) / sum(`ECONOMY ACTIVE`)
  )
```

### Joining the attribute and geospatial data

The `PA` and `SZ` values are mixed case while `PLN_AREA_N` and `SUBZONE_N` are uppercase, so they will not match until we fix that.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
mpsz_pop2024 <- left_join(mpsz19, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

The sf object goes on the left of the join deliberately. That is what keeps the result an sf data frame rather than a plain tibble.

```{r}
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## Choropleth Mapping with tmap

### The quick way, qtm()

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024, fill = "DEPENDENCY")
```

`qtm()` is short to write, but look at what the default binning did. The legend goes 0 to 10 in steps of two. Almost every subzone is below 1, so the whole island ends up in the lightest class with one dark polygon near Changi. Nothing is wrong with the map, it just does not show anything.

The cause is the single subzone with a ratio of 10. That is why the custom breaks section below matters, and why the rest of this section uses the longer syntax instead, since `qtm()` gives you no way to change the bins.

### Drawing a base map

`tm_shape()` sets the data, then layer functions draw on top of it.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

### Drawing a choropleth with tm_polygons()

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

The default binning is "pretty" and the default palette is blues3. Missing values come out grey.

### tm_fill() and tm_borders() separately

`tm_polygons()` is really a wrapper around these two.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```

Adding the borders back in.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

`tm_borders()` takes `col`, `lwd` and `lty` for colour, line width and line type.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60", lwd = 0.1, lty = "dashed")
```

## Data Classification Methods

tmap offers ten classification styles. The choice matters more than it looks, because the same data can tell quite different stories depending on where the breaks fall.

### Quantile

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

### Equal interval

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

Same column, same data, two maps that look nothing alike.

Quantile puts the same number of subzones in each class, so the breaks come out at 0.64, 0.70, 0.74 and 0.82. Those all sit inside the range where the data actually is, and the map separates the central and eastern subzones clearly.

Equal interval cuts 0 to 10 into five equal parts. Nearly everything falls in the first one, so the island goes pale with a single dark polygon. If you only saw that map you would think dependency hardly varies in Singapore, which is not true.

Neither map is wrong. One subzone at 10 is enough to make the default useless.

### Custom breaks

Always look at the numbers before choosing break points.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Most values sit between 0.6 and 1.0, so breaks at 0.60, 0.70, 0.80 and 0.90 will separate the bulk of the data. Remember tmap needs n+1 values including a minimum and maximum.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

## Colour Scheme

### ColorBrewer palettes

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

A minus prefix reverses the ramp.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### Cartographic furniture

A compass, scale bar, grid and credits turn a plot into something that reads as a map.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```
:::

## Map Layout

### Legend

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### Map style

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

Resetting to the default so it does not carry over into later chunks.

```{r}
tmap_style("white")
```

## Small Multiple Choropleth Maps

Facet maps put several maps side by side, which makes it easy to see how a spatial pattern shifts with another variable.

### By passing multiple variables to an aesthetic

::: panel-tabset
## The Plot

```{r}
#| echo: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```
:::

### By arranging separate maps in a grid

```{r}
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

The two maps look almost the same. I expected the aged map to light up the older estates and the young map to light up the newer towns, but the same subzones are dark on both.

That is because both are raw counts. A subzone with more people has more of everyone. So the maps are mostly showing where people live, not where the population is young or old. To answer that you need a rate, which is what `DEPENDENCY` does.

### By a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(style = "quantile",
                                          values = "brewer.blues")) +
  tm_facets(by = "REGION_N", nrow = 2, ncols = 3,
            free.coords = TRUE, drop.units = TRUE) +
  tm_borders(fill_alpha = 0.5)
```

## Mapping a Selection Criterion

Rather than facetting, we can just filter down to the region of interest.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## Complementing the Map with a Statistical Chart

A map shows where, a chart shows how much. `fill.chart` puts both in one figure.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

The boxplot shows something the map does not. All the outliers are below the box, near zero, not above it. Those are subzones with working age residents and hardly any children or elderly.

Once labelled, they turn out to be bunched in the downtown core. The labels overlap a lot at this size and would need fixing before I used this anywhere, but you can still see where they are.

::: panel-tabset
## The Plot

```{r}
#| echo: false
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## The Code

```{r}
#| eval: false
mpsz_selected_central <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected_central$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected_central[mpsz_selected_central$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected_central) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```
:::

## Creating an Interactive Map

Switching `tmap_mode()` to view turns the same code into a zoomable, clickable map.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

`set_zoom_limits` is doing real work here. Total freedom to zoom sounds generous but in practice users end up lost somewhere over the ocean, so bounding the range keeps them inside the study area.

```{r}
tmap_mode("plot")
```

# Reflections

Four things I am taking away from this.

First, copying a workflow step by step is not the same as understanding it. data.gov.sg has dropped the shapefile and KML versions of the Master Plan layer, and the geojson that replaced it comes in WGS84, not SVY21. The book uses `st_set_crs()` there to fix a wrong EPSG label. If I had copied that line, it would have put an SVY21 label on degree values and nothing would have errored. Every buffer, area and count after it would have been wrong. What caught it was looking at the bounding box and seeing the numbers were degrees.

Second, the same problem showed up again in the population wrangling. The book picks age bands by column position, which stops working the moment `pivot_wider()` sorts `5_to_9` after `55_to_59`. Run as written it would have counted primary school children as economically active and put 50 year olds in the young band, and it would have rendered fine. Checking that the bands added up to the published resident population, and that the overall ratio was around 0.73, took a minute and was the only thing that would have caught it.

Third, I got one of my own readings wrong. I assumed the young and aged maps would show different parts of the island and they do not, because both are counts and counts mostly follow population size. I only noticed after looking at the two maps side by side. A map can look useful while answering a different question from the one you had in mind.

Fourth, the classification method does a lot of the work. Quantile and equal interval on the same column give maps that say opposite things about whether dependency varies across Singapore. One subzone at 10 is enough to flatten the equal interval map. Picking a classification is part of the analysis, not formatting, and the default is not a safe choice.
