---
title: "Hands-on Exercise 3"
subtitle: "Spatio-Temporal Point Patterns Analysis"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "September 10, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

::: {.pagefacts}
[Two datasets, both global, both filtered to one province]{.fact}
[NASA FIRMS MODIS 2023, GADM 4.1 level 2]{.fact .data}
[`sf` `spatstat` `sparr` `stpp` `tmap` `tidyverse`]{.fact .pkg}
:::

# Overview

Exercises 1 and 2 treated every point as if it happened at the same time. This one adds time. Two fires a kilometre apart on 3 August and 4 August are a different pattern from the same two fires six months apart. The methods here can tell them apart.

The chapter asks two questions. Are the fires in Kepulauan Bangka Belitung independent in space and time? If not, where and when do they cluster?

# Loading the Packages

Two of the six packages are new. **sf** reads and transforms the boundary and the fire points. **spatstat** supplies the `ppp` and `owin` objects and the L function envelope. **sparr** computes the spatio-temporal kernel density and picks its bandwidths. **stpp** computes the spatio-temporal K function. **tmap** draws the point maps. **tidyverse** does the reading, filtering and counting.

```{r}
pacman::p_load(sf, spatstat, sparr, stpp, tmap, tidyverse)
```

::: {.callout-warning}
## Issue I hit
`sparr` would not load on my Mac. It depends on a package called `rpanel`, which needs X11, and Apple stopped shipping X11 years ago.

I fixed it by installing XQuartz, then logging out and back in so the display variable gets set. After that it loads with a warning about display `:0`, which I can ignore. X11 is only there so `rpanel` can draw sliders, and this exercise never uses them.
:::

# Preparing the Data

## Importing and Preparing Study Area

The book uses a kelurahan shapefile from an Indonesian portal, but the download ran at 44 kilobytes a second, three hours for 553 MB. I used GADM level 2 instead, which has regency boundaries for the whole country in 4.6 MB.

::: panel-tabset
## Importing the study area

`st_read()` of **sf** reads the national GADM file. The rest of the chain keeps the province, merges it into one outline, drops the Z dimension and projects it to UTM.

```{r}
kbb_sf <- st_read("data/rawdata/gadm/gadm41_IDN_2.json") %>%
  filter(NAME_1 == "BangkaBelitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

::: {.callout-note}
## Learning diary
The province name in GADM is `BangkaBelitung`, with no space. I printed the unique values instead of guessing, which avoided a silent empty filter.

`st_union()` merges the seven level-2 units (six regencies and the city of Pangkal Pinang) into one outline. Only the outer boundary is needed, so kelurahan, kecamatan or kabupaten would all merge to the same shape.

The CRS is 32748, UTM zone 48 south, not 3414. Bangka Belitung sits at 106 east, which is zone 48, and at 2 south, below the equator. Singapore's SVY21 would put the island in the wrong place.
:::

## Converting to owin

`as.owin()` of **spatstat** turns the merged outline into an observation window. Every point pattern below is clipped to it.

```{r}
kbb_owin <- as.owin(kbb_sf)
class(kbb_owin)
```
:::

A quick `plot()` of the window checks that the union and the projection worked before I add any points.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 8
#| fig-height: 6
#| echo: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
plot(kbb_owin, main = "Kepulauan Bangka Belitung, the observation window",
     col = "#EFE9DB", border = "grey40")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
plot(kbb_owin, main = "Kepulauan Bangka Belitung, the observation window",
     col = "#EFE9DB", border = "grey40")
```
:::

## Importing and Preparing Forest Fire Data

FIRMS publishes MODIS detections as one file per country per year. The all-countries bundle arrived cut off at 43 of 82 MB, with Indonesia in the missing part, so I downloaded the Indonesia file directly.

::: panel-tabset
## Importing the Indonesia file

`read_csv()` of **readr** (part of **tidyverse**) reads the file, and `nrow()` counts the rows.

```{r}
fire_all <- read_csv("data/rawdata/modis/modis_2023_Indonesia.csv")
nrow(fire_all)
```

## Checking the type column

This is every detection in Indonesia for 2023. Before clipping to the study area, I check one column the book never mentions. `table()` counts the rows of each `type`.

```{r}
table(fire_all$type)
```
:::

::: {.callout-caution}
## Where I differ from the book
The book keeps every row and calls them all forest fires. The `type` column shows that some are not.

Type 0 is a presumed vegetation fire. Type 1 is an active volcano. Type 2 is another static land source, such as gas flares or industrial heat. Type 3 is offshore. Indonesia has many volcanoes and flares, and 4,250 detections are one or the other (1,495 volcano, 2,755 static source), and another 23 are offshore.

A volcano that shows up every day at the same spot would look like strong spatio-temporal clustering. But it is not a fire. I keep type 0 only.
:::

::: panel-tabset
## Filtering and clipping

`filter()` keeps type 0, `st_as_sf()` turns the longitude and latitude into points, `st_transform()` projects them to the same UTM zone as the boundary, and `st_filter()` of **sf** keeps only the points inside Bangka Belitung.

```{r}
fire_sf <- fire_all %>%
  filter(type == 0) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32748) %>%
  st_filter(st_as_sf(kbb_sf))

nrow(fire_sf)
```

::: {.callout-note}
## Learning diary
The book's csv was already cut to Bangka Belitung, so it has no clipping step. Mine covers all of Indonesia, so `st_filter()` keeps only the points that intersect the boundary polygon.

My count differs from the book's 741, as expected. The boundary source is different, and the type filter removed one static-source row that the book would have kept.
:::

## Deriving the time fields

`as.ppp()` only accepts numbers or text as marks, so the date has to become a number. `yday()` and `month()` of **lubridate** (part of **tidyverse**) give three versions for three uses later on.

```{r}
fire_sf <- fire_sf %>%
  mutate(acq_date = as.Date(acq_date),
         DayofYear = yday(acq_date),
         Month_num = month(acq_date),
         Month_fac = month(acq_date, label = TRUE, abbr = FALSE))
```
:::

# Visualising the Fire Points

## Overall Plot

`tm_shape()` and `tm_dots()` of **tmap** put every 2023 detection on the merged outline.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 8
#| fig-height: 6
#| echo: false
tm_shape(kbb_sf) +
  tm_polygons(fill = "#F7F4EC", col = "grey60") +
  tm_shape(fire_sf) +
  tm_dots(fill = "#D55E00", fill_alpha = 0.5, size = 0.15) +
  tm_title("MODIS fire detections, Bangka Belitung, 2023", size = 1.5, fontface = "bold") +
  tm_credits("Source: NASA FIRMS, GADM", position = c("left", "bottom"))
```

## The Code

```{r}
#| eval: false
tm_shape(kbb_sf) +
  tm_polygons(fill = "#F7F4EC", col = "grey60") +
  tm_shape(fire_sf) +
  tm_dots(fill = "#D55E00", fill_alpha = 0.5, size = 0.15) +
  tm_title("MODIS fire detections, Bangka Belitung, 2023", size = 1.5, fontface = "bold") +
  tm_credits("Source: NASA FIRMS, GADM", position = c("left", "bottom"))
```
:::

## Visualising Geographic Distribution of Forest Fires by Month

Next I split the map by month, which is where time first shows up. `tm_facets()` of **tmap** splits the same map into one panel per value of `Month_fac`.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 7
#| echo: false
tm_shape(kbb_sf) +
  tm_polygons(fill = "#F7F4EC", col = "grey70") +
  tm_shape(fire_sf) +
  tm_dots(fill = "#D55E00", fill_alpha = 0.6, size = 0.2) +
  tm_facets(by = "Month_fac", free.coords = FALSE, drop.units = TRUE) +
  tm_title("Fires by month, the dry season lights up from July", size = 1.3, fontface = "bold") +
  tm_layout(panel.label.size = 1.2)
```

## The Code

```{r}
#| eval: false
tm_shape(kbb_sf) +
  tm_polygons(fill = "#F7F4EC", col = "grey70") +
  tm_shape(fire_sf) +
  tm_dots(fill = "#D55E00", fill_alpha = 0.6, size = 0.2) +
  tm_facets(by = "Month_fac", free.coords = FALSE, drop.units = TRUE) +
  tm_title("Fires by month, the dry season lights up from July", size = 1.3, fontface = "bold") +
  tm_layout(panel.label.size = 1.2)
```
:::

To get numbers for the panels, `st_drop_geometry()` turns the points back into a table and `count()` counts the detections per month.

```{r}
fire_sf %>%
  st_drop_geometry() %>%
  count(Month_fac, name = "fires")
```

The first half of the year is nearly empty, with 101 of the 897 detections. August to October have most of the fires, 589 of them. This is the dry season, and it suggests time matters at least as much as location.

# Computing STKDE by Month

Exercise 2's KDE gave one density surface for the whole year. Here there is one surface for every point in time. So the estimate at a location depends on how many fires happened nearby *and* recently.

## Creating ppp and Including Owin Object

::: panel-tabset
## Creating the ppp inside the owin

`as.ppp()` of **spatstat** turns the points into a point pattern with `Month_num` as the mark. Indexing it by `kbb_owin` clips it to the observation window.

```{r}
fire_month_ppp <- fire_sf %>%
  select(Month_num) %>%
  as.ppp()

fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

## Checking for duplicates

`duplicated()` flags points with the same location and mark. Kernel estimators do not handle these well.

```{r}
any(duplicated(fire_month_owin))
```
:::

## Computing Spatio-Temporal KDE

`spattemp.density()` of **sparr** estimates the density over space and time with its default bandwidths, and `summary()` prints them.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

::: {.callout-note}
## Learning diary
This gives two bandwidths. `h` is the spatial one in metres, around 19 km. `lambda` is the temporal one in months. It is 0.027 of a month, which is less than a day.

Less than a day of smoothing on data recorded in whole months is odd. It means month 8 takes almost nothing from months 7 or 9.

I did not understand why until the day-of-year section below, where the same function picks a very different lambda. The default rule depends on how the time values are spaced, and whole months are very coarse.
:::

## Plotting the Spatio-Temporal KDE Object

**sparr**'s `plot()` method draws one time slice per call. It also adds `axis(1)`, `axis(2)` and an L-shaped box to every panel. So my first version labelled each small map in projected metres like 9.7e+06, which mean nothing to a reader and got cut off at the edges. The function below draws one slice with **spatstat**'s `plot()` for pixel images and adds the window outline, without the axes. It takes the colour range as an argument so panels can share one scale.

```{r}
plot_slice <- function(kde, t, zlim, main) {
  tt <- as.numeric(names(kde$z))
  slice <- kde$z[[which.min(abs(tt - t))]]
  plot(slice, main = main, zlim = zlim, box = FALSE,
       ribargs = list(box = TRUE))
  plot(as.polygonal(Window(kde$pp)), add = TRUE)
}
```

::: {.callout-note}
## Learning diary
The obvious fix is `par(xaxt = "n", yaxt = "n")` before the loop. But that would also remove the numbers on the colour ribbon. spatstat draws those with `axis()` on side 4, and a global `yaxt = "n"` turns that off too. The source of `plot.stden()` is a short loop around a `plot.im()` call, so I copied the two lines I needed.
:::

The panels show July to December. The colour key is the estimated joint density, the share of the year's detections per square metre per month, so the numbers are tiny (the peak is about 6.4e-10). `zlim_month` is the range over all twelve monthly surfaces, so the six panels share one colour scale.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 5.5
#| echo: false
zlim_month <- range(sapply(st_kde$z, range))
tims <- c(7, 8, 9, 10, 11, 12)
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (i in tims) {
  plot_slice(st_kde, i, zlim = zlim_month,
             main = paste(month.name[i], "2023"))
}
mtext("Monthly fire intensity, July to December 2023", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
zlim_month <- range(sapply(st_kde$z, range))
tims <- c(7, 8, 9, 10, 11, 12)
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (i in tims) {
  plot_slice(st_kde, i, zlim = zlim_month,
             main = paste(month.name[i], "2023"))
}
mtext("Monthly fire intensity, July to December 2023", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

The shared `zlim` works like sparr's `fix.range = TRUE`, which uses the same range for every time slice. Without it each month gets its own scale, and a quiet December looks as busy as a peak September.

# Computing STKDE by Day of Year

## Computing Spatio-Temporal KDE

::: panel-tabset
## Creating the ppp inside the owin

These are the same steps as the monthly version, but with `DayofYear` as the mark.

```{r}
fire_yday_owin <- fire_sf %>%
  select(DayofYear) %>%
  as.ppp() %>%
  .[kbb_owin]

summary(fire_yday_owin)
```

## Computing the KDE

I run `spattemp.density()` again with its default bandwidths, now with time measured in days.

```{r}
kde_yday <- spattemp.density(fire_yday_owin)
summary(kde_yday)
```
:::

::: {.callout-warning}
## Issue I hit
The book's next line is `plot(kde_yday)`. It produces 343 separate plots, one for every day in the temporal bound. I fetched the rendered chapter page and counted 343 image tags in a row.

`plot()` on a sparr object takes a `tselect` argument for which times to draw, and `plot_slice()` above takes one time in the same way. Six days from the dry season show the same pattern as all 343.
:::

## Plotting the Spatio-Temporal KDE Object

I plot six days at fifteen-day steps, from 1 August to 15 October. The colour key is now the share of the year's detections per square metre per day (the peak is about 1.9e-12). `zlim_yday` gives all six panels one scale.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 5.5
#| echo: false
zlim_yday <- range(sapply(kde_yday$z, range))
days <- c(213, 228, 243, 258, 273, 288)
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days) {
  plot_slice(kde_yday, d, zlim = zlim_yday,
             main = paste0("Day ", d, " (", format(as.Date("2023-01-01") + d - 1, "%d %b"), ")"))
}
mtext("Daily fire intensity, default bandwidths", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
zlim_yday <- range(sapply(kde_yday$z, range))
days <- c(213, 228, 243, 258, 273, 288)
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days) {
  plot_slice(kde_yday, d, zlim = zlim_yday,
             main = paste0("Day ", d, " (", format(as.Date("2023-01-01") + d - 1, "%d %b"), ")"))
}
mtext("Daily fire intensity, default bandwidths", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

# Computing STKDE by Day of Year: Improved Method

## Computing Spatio-Temporal KDE

The two default bandwidths above come from a rule of thumb. `BOOT.spattemp()` of **sparr** picks them by bootstrap instead. It minimises the estimated error of the whole surface.

::: panel-tabset
## Choosing the bandwidths by bootstrap

`set.seed()` fixes the resampling so the result can be repeated. Then `BOOT.spattemp()` searches over both bandwidths at once.

```{r}
set.seed(1234)
bw_boot <- BOOT.spattemp(fire_yday_owin)
bw_boot
```

## Refitting with the rounded bandwidths

I run `spattemp.density()` again and pass in `h` and `lambda` by hand.

```{r}
kde_yday_boot <- spattemp.density(fire_yday_owin,
                                  h = 45000,
                                  lambda = 17)
summary(kde_yday_boot)
```
:::

::: {.callout-note}
## Learning diary
The bootstrap is slow. It prints thirty or so iterations before it stops, and each one refits the surface. This chunk is why I set `freeze: true` at the top of the file, so every render does not rerun it.

The book's bootstrap gives h of 9000 and lambda of 19, and the chapter rounds them before use. Mine gives about 45.6 km and 17 days. The samples are different, so I round my own numbers to 45000 and 17 instead of using the book's.

Rounding makes sense in both cases. Six decimal places from a bootstrap suggest more precision than the method has.
:::

## Plotting the Output Spatio-Temporal KDE

These are the same six days from the bootstrap surface, in the same unit as above. But this figure has its own `zlim_boot`, so I compare patterns between the two figures, not colours.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 5.5
#| echo: false
zlim_boot <- range(sapply(kde_yday_boot$z, range))
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days) {
  plot_slice(kde_yday_boot, d, zlim = zlim_boot,
             main = paste0("Day ", d, " (", format(as.Date("2023-01-01") + d - 1, "%d %b"), ")"))
}
mtext("Daily fire intensity, bootstrap bandwidths", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
zlim_boot <- range(sapply(kde_yday_boot$z, range))
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days) {
  plot_slice(kde_yday_boot, d, zlim = zlim_boot,
             main = paste0("Day ", d, " (", format(as.Date("2023-01-01") + d - 1, "%d %b"), ")"))
}
mtext("Daily fire intensity, bootstrap bandwidths", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

::: {.callout-note}
## Learning diary
The default gave h of about 19 km and lambda of about 6 days. The bootstrap gave 45 km and 17 days. So the bootstrap smooths more on both axes, more than twice in space and nearly three times in time.

Both sets of maps show the same pattern. A hotspot starts in west Bangka in early August and moves to the south of the island by late September. Belitung stays quiet the whole time.

The default keeps the hotspot tight and Belitung dark. The bootstrap spreads the same hotspot across half of Bangka and halves its peak (the summaries give a maximum of 1.9e-12 for the default and 1.0e-12 for the bootstrap). It also makes Belitung light up, because a 45 km kernel reaches across the strait.

Which one to use depends on the question. For where fires happen, the default's sharper space is more useful. For when they happen, the bootstrap's smoother time is better. The bootstrap minimises one overall error and does not know which question I am asking.

The book got 9 km and 19 days on its 741 points, which is sharper in space than its own default. With the same method on a different sample, the spatial bandwidth moved the other way. So the bandwidth depends on the data in hand.
:::

# Spatio-Temporal Point Patterns Analysis: stpp Methods

The KDE is a first-order method. It shows where and when fires are dense. The K function is second order. It tests whether fires happen near other fires in space and time together, more than they would if location and timing were unrelated.

## Preparing Spatio-Temporal Point Process Object of stpp

stpp needs a plain three-column matrix with x, y and an integer time. `st_coordinates()` of **sf** pulls out the projected x and y, and `as.3dpoints()` of **stpp** wraps the three columns as an `stpp` object.

```{r}
coords <- st_coordinates(fire_sf)

fire_df <- data.frame(x = coords[, 1],
                      y = coords[, 2],
                      t = fire_sf$DayofYear)

fire_stpp <- as.3dpoints(fire_df)
class(fire_stpp)
```

stpp's own `plot()` for this object gives two panels titled "xy-locations" and "cumulative number", with axes labelled x, y and t and a stretched map. I draw the same two panels with base `plot()` from `fire_df`, which has the same three columns. The map uses `asp = 1` so a kilometre east is as long as a kilometre north, and the count is a step line.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 5
#| echo: false
par(mfrow = c(1, 2), cex.main = 1.3, cex.lab = 1.1)
plot(fire_df$x, fire_df$y, asp = 1, pch = 20, cex = 0.5, col = "#D55E00",
     main = "Fire locations", xlab = "Easting (m)", ylab = "Northing (m)")
plot(sort(fire_df$t), seq_len(nrow(fire_df)), type = "s",
     main = "Cumulative fires through 2023", xlab = "Day of year", ylab = "Fires")
par(mfrow = c(1, 1))
```

## The Code

```{r}
#| eval: false
par(mfrow = c(1, 2), cex.main = 1.3, cex.lab = 1.1)
plot(fire_df$x, fire_df$y, asp = 1, pch = 20, cex = 0.5, col = "#D55E00",
     main = "Fire locations", xlab = "Easting (m)", ylab = "Northing (m)")
plot(sort(fire_df$t), seq_len(nrow(fire_df)), type = "s",
     main = "Cumulative fires through 2023", xlab = "Day of year", ylab = "Fires")
par(mfrow = c(1, 1))
```
:::

The left panel shows every detection in UTM metres, with Bangka on the left and Belitung on the right. The right panel counts the fires cumulatively through the year. The dry season is the steep climb from about day 200.

## Computing Spatio-Temporal K-Function

`STIKhat()` of **stpp** estimates the space-time K function on a default grid of spatial and temporal distances.

```{r}
kbb_stik <- STIKhat(fire_stpp)
```

::: {.callout-warning}
## Issue I hit
The book's `plotK(kbb_stik)` rendered nothing. The chunk ran, but the page had an empty tab where the surface should be.

The function source explains why. `plotK()` closes the current graphics device and opens a new one for its own layout. Knitr was recording into the device that got closed, so the plot went to a window nobody saw.

My fix was to skip `plotK()` and draw the surface myself. The object holds `Khat`, the estimate, and `Ktheo`, the value expected if space and time were independent. Both are over `dist` and `times`. I plot the difference, which the reading below uses, and label the axes properly.
:::

`filled.contour()` from base **graphics** draws the difference as a coloured surface, and `contour()` adds labelled lines on top. `STIKhat()` returns distances in metres, so I divide `u` by 1000 to put the x axis in kilometres. The colour key, titled "K excess", is `Khat` minus `Ktheo`. It is in the K function's own units, square metres times days.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 8
#| fig-height: 6
#| echo: false
u <- kbb_stik$dist / 1000
v <- kbb_stik$times
k_excess <- kbb_stik$Khat - kbb_stik$Ktheo

par(cex.main = 1.3)
filled.contour(u, v, k_excess,
               color.palette = function(n) hcl.colors(n, "YlOrRd", rev = TRUE),
               xlab = "u, spatial distance (km)",
               ylab = "v, temporal distance (days)",
               main = "Space-time K minus the independence expectation",
               key.title = title(main = "K excess", cex.main = 0.9),
               plot.axes = {
                 axis(1); axis(2)
                 contour(u, v, k_excess, add = TRUE, col = "grey10", lwd = 0.6,
                         labcex = 0.9)
               })
```

## The Code

```{r}
#| eval: false
u <- kbb_stik$dist / 1000
v <- kbb_stik$times
k_excess <- kbb_stik$Khat - kbb_stik$Ktheo

par(cex.main = 1.3)
filled.contour(u, v, k_excess,
               color.palette = function(n) hcl.colors(n, "YlOrRd", rev = TRUE),
               xlab = "u, spatial distance (km)",
               ylab = "v, temporal distance (days)",
               main = "Space-time K minus the independence expectation",
               key.title = title(main = "K excess", cex.main = 0.9),
               plot.axes = {
                 axis(1); axis(2)
                 contour(u, v, k_excess, add = TRUE, col = "grey10", lwd = 0.6,
                         labcex = 0.9)
               })
```
:::

## Guide to Interpreting the Plot

::: {.callout-note}
## Learning diary
This plot took me a while to read.

The plot is a surface over two axes. `u` is distance in space, which I plotted in kilometres, and `v` is distance in days. The height is how many more fires lie within `u` kilometres and `v` days of a typical fire than expected if space and time were independent. This is the `Khat` minus `Ktheo` I drew above.

If the surface rises steeply near the origin, at small `u` and small `v`, fires happen close to other fires and soon after them. That is space-time clustering. A flat surface means no such interaction.

The name says inhomogeneous, but the call above is not. `STIKhat()` only corrects for varying intensity if you pass it a `lambda`, the intensity at every point. The book does not pass one and neither did I. So it uses a constant, the point count divided by the space-time volume, and is really a homogeneous K.

The surface shows this. The excess is positive everywhere and keeps rising with both `u` and `v`, out to 45 km and 80 days. Clustering that keeps growing with scale comes from the dry season and the fire belt of Bangka. The KDE already found both. It does not show fires near fires setting off more fires.

A proper inhomogeneous version would pass the spatio-temporal KDE in as `lambda`. I want to use that version for the take-home.
:::

# Beyond the Exercise

I checked three things that the chapter does not.

## Does the Type Filter Change the Answer

I dropped volcanoes and static sources while preparing the data. Here I check what keeping them would have changed. The same `st_as_sf()`, `st_transform()` and `st_filter()` chain runs on the unfiltered table, and a `tibble()` puts the two counts side by side.

```{r}
fire_all_kbb <- fire_all %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32748) %>%
  st_filter(st_as_sf(kbb_sf))

tibble(
  kept = c("all types", "type 0 only"),
  detections = c(nrow(fire_all_kbb), nrow(fire_sf)),
  dropped = detections[1] - detections
)
```

`count()` on the unfiltered points shows which types were there to drop.

```{r}
fire_all_kbb %>%
  st_drop_geometry() %>%
  count(type, name = "detections")
```

::: {.callout-tip}
## Alternative explored
For Bangka Belitung the filter removes exactly one row. It is a single type 2 static-source detection, and the count goes from 898 to 897. The maps above would look the same either way. There are no volcanoes on the island.

This is still a useful result. The filter was cheap and the data dictionary supports it. For a Kalimantan regency in the take-home it will matter much more, because Kalimantan has oil and gas flares and other industrial heat sources that FIRMS tags as type 2. I would rather build the habit now.
:::

## Does Detection Confidence Change the Answer

MODIS gives every detection a confidence score from 0 to 100. The book ignores it. `quantile()` shows how the scores are spread.

```{r}
quantile(fire_sf$confidence, c(0, 0.25, 0.5, 0.75, 1))
```

`filter()` keeps the detections at 80 or above, and the `tibble()` compares the counts.

```{r}
fire_hi <- fire_sf %>% filter(confidence >= 80)

tibble(
  threshold = c("all", "confidence >= 80"),
  detections = c(nrow(fire_sf), nrow(fire_hi))
)
```

The subset goes through the same `as.ppp()` and `spattemp.density()` steps with the bootstrap bandwidths, so only the data change.

```{r}
fire_hi_owin <- fire_hi %>% select(DayofYear) %>% as.ppp() %>% .[kbb_owin]
kde_hi <- spattemp.density(fire_hi_owin, h = 45000, lambda = 17)
```

I plot three of the six days, with every detection on the top row and the high confidence subset below. `zlim_conf` is the range over both surfaces, so all six panels share one colour scale in the same unit as before.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 10
#| fig-height: 5.5
#| echo: false
zlim_conf <- range(c(sapply(kde_yday_boot$z, range), sapply(kde_hi$z, range)))
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days[c(2, 4, 6)]) {
  plot_slice(kde_yday_boot, d, zlim = zlim_conf,
             main = paste("All, day", d))
  plot_slice(kde_hi, d, zlim = zlim_conf,
             main = paste("Conf >= 80, day", d))
}
mtext("All fires against confidence 80 and above, one scale", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
zlim_conf <- range(c(sapply(kde_yday_boot$z, range), sapply(kde_hi$z, range)))
par(mfcol = c(2, 3), mar = c(0.5, 0.5, 2, 4), oma = c(0, 0, 3, 0), cex.main = 1.8)
for (d in days[c(2, 4, 6)]) {
  plot_slice(kde_yday_boot, d, zlim = zlim_conf,
             main = paste("All, day", d))
  plot_slice(kde_hi, d, zlim = zlim_conf,
             main = paste("Conf >= 80, day", d))
}
mtext("All fires against confidence 80 and above, one scale", outer = TRUE, cex = 2.1, font = 2)
par(mfcol = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

::: {.callout-tip}
## Alternative explored
Eighty is FIRMS's own cut-off for high confidence, and it is strict. Half the detections are below 67 and only 164 of 897 reach 80. So the bottom row uses under a fifth of the data.

Even with four fifths of the points gone, the hotspots are in the same places and move the same way through the season. Both surfaces are densities that integrate to one over space and time, so the smaller subset does not just come out dimmer. The thing to look for is the bright patches moving, and they do not.

This is a useful sanity check. A very different pattern in the high confidence subset would have meant the full-data maps were driven by noise. The pattern held, so they are not. The chapter skips the choice of threshold, which is a data quality decision. For the take-home I have to state and justify it.
:::

## Is It Space, Time, or Both

The K surface above tests for space-time interaction. It does not show whether there would be spatial clustering anyway with time ignored, which Exercise 2's tools can check. `unmark()` drops the day marks. `envelope()` of **spatstat** simulates 39 random patterns in the same window around `Lest`, with `global = TRUE` for a simultaneous envelope.

```{r}
fire_sp <- unmark(fire_yday_owin)
set.seed(1234)
L_sp <- envelope(fire_sp, Lest, nsim = 39, rank = 1,
                 global = TRUE, correction = "Ripley", verbose = FALSE)
```

The plot shows L(d) minus d against distance in metres. The solid line is the observed pattern and the dashed line at zero is complete spatial randomness. The grey band is the global envelope.

::: panel-tabset
## The Plot

```{r}
#| fig-width: 8
#| fig-height: 6
#| echo: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
plot(L_sp, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r",
     main = "Spatial L function, time ignored, global envelope")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1)
plot(L_sp, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r",
     main = "Spatial L function, time ignored, global envelope")
```
:::

::: {.callout-tip}
## Alternative explored
The two results need to be read together. The L function shows the fires cluster in space when the whole year is put on one map. The STIK surface shows whether there is *additional* clustering once time is included.

If both show clustering, fires are drawn to the same places and also happen in bursts. If only the spatial one does, the bursts are just the dry season, and nearby fires are not setting each other off.

Here both show clustering. But the K surface used a constant intensity, so the dry season is still inside it. So I can only say fires cluster in space and in time. Whether the two interact is untested until I pass in `lambda`.

The chapter runs the second test without the first, so it cannot make that distinction. I need both to interpret the space-time result.
:::

# Limitations

- GADM's boundary and the book's kelurahan shapefile come from different sources, so my observation window and point count differ from the book's.
- MODIS detects heat, not fire. A detection is a hot pixel about a kilometre across, so one fire may show as several points and several fires as one.
- Cloud cover hides fires. Missing days in the dry season may be weather, not a lack of fire.
- Two satellites, Terra and Aqua, pass at different times of day, so the daily record is two snapshots, not a continuous one.
- The bootstrap bandwidths optimise one overall error. They do not know whether the analyst cares more about where or about when.
- The STIK surface has no envelope in this exercise, so a high contour is suggestive but not tested. It also used a constant intensity, so it is not yet the inhomogeneous version its name says.

# Reflections

These are five things I learned from this exercise.

**The chapter never mentions the `type` column.** It was the most important column, because FIRMS puts volcanoes, gas flares and offshore pixels in the same table as vegetation fires and labels them. Calling them all forest fires would have been wrong here, and much worse for Kalimantan in the take-home. I want to keep the habit of reading the data dictionary before the data.

**Two downloads failed before one worked.** First a 553 MB shapefile at 44 kilobytes a second, then a bundle that arrived cut off with Indonesia in the missing part. The per-country FIRMS file and GADM level 2 gave me the same analysis from 9 MB in total. The book's data source is not the only option.

**The default and bootstrap bandwidths disagreed on both space and time.** Bandwidth is now two numbers. On my data the bootstrap smoothed more on both axes, but on the book's data it went sharper in space. Neither is wrong, and the right choice depends on whether the question is where or when, which no selector can know.

**Two plot calls in the book did not work as written.** `plot(kde_yday)` with no time selected gives 343 images. `plotK()` gives none, because it swaps graphics devices under knitr. I caught the first from the book's own rendered page, and the second only because I checked my own render before pushing it.

**The space-time K function needs the spatial L function and an intensity next to it.** By itself it only shows there is clustering. Next to the L function from Exercise 2, and with a `lambda` so it is really inhomogeneous, it can show whether the clustering is more than the dry season. The chapter does neither, and I think that is a gap.
