---
title: "Hands-on Exercise 2"
subtitle: "Spatial Point Patterns Analysis"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "September 5, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

::: {.pagefacts}
[Two datasets, both from data.gov.sg]{.fact}
[Child Care Services, Master Plan 2019 subzones]{.fact .data}
[`sf` `spatstat` `terra` `tmap` `tidyverse`]{.fact .pkg}
:::

# Overview

Both chapters ask one question. Are childcare centres spread evenly across Singapore, and if not, where do they bunch up?

Chapter 4 looks at how many centres sit in each part of the map. Chapter 5 looks at how far each centre is from its neighbours. The readings also discuss whether these two views can be separated at all.

# Loading the Packages

`p_load()` of **pacman** loads five packages and installs any that are missing. **sf** handles the vector data and **spatstat** the point pattern analysis. **terra** converts spatstat's raster output into a form tmap can draw. **tmap** draws the maps and **tidyverse** does the wrangling.

```{r}
pacman::p_load(sf, spatstat, terra, tmap, tidyverse)
```

::: {.callout-warning}
## Issue I hit
Loading terra after spatstat prints a masking warning. terra masks eight names, including `rescale` and `area`, which spatstat also uses.

This exercise calls `rescale()` on point patterns a lot. The book writes `rescale.ppp()` in full to avoid the clash, which I only understood after reading the warning properly.
:::

# Preparing the Data

data.gov.sg offers both KML and geojson. I used the geojson, which removes a chunk of the book's code. Both layers are reprojected to SVY21 (EPSG 3414) when they are read, so all distances later are in metres.

::: panel-tabset
## Importing the childcare centres

`st_read()` of **sf** reads the geojson and `st_transform()` reprojects it to EPSG 3414.

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson") %>%
  st_transform(crs = 3414)
```

`names()` lists the attribute columns, so I can compare the geojson with the book's KML.

```{r}
names(childcare_sf)
```

::: {.callout-note}
## Learning diary
As in Exercise 1, the book reads the KML, which hides every attribute in an HTML table inside a `Description` field. It then writes an `extract_kml_field()` helper with **rvest** to pull them out. The geojson already has the columns.

So I skip the helper and drop rvest from the package list. The KML is also XYZ, so the book calls `st_zm(drop = TRUE)` to strip the elevation. My geojson is already XY, so I skip that too. The `DESCRIPTION` column here just holds the text "Child Care Services", unlike the KML's `Description` blob.
:::

## Importing the subzone boundary

The Master Plan 2019 subzones come in the same way, again reprojected to EPSG 3414.

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

## Dropping the offshore islands

First I remove the offshore subzones. They sit far from the mainland and would stretch the window over empty sea, pulling every density estimate down. `filter()` of **dplyr** drops them, and the two `nrow()` calls show the count going from 332 to 327.

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

nrow(mpsz_sf)
nrow(mpsz_cl)
```
:::

## Mapping the geospatial data sets

`tm_shape()`, `tm_polygons()` and `tm_dots()` of **tmap** draw the centres over the cleaned subzones.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(mpsz_cl) +
  tm_polygons(col = "grey85", fill = "#F7F4EC") +
  tm_shape(childcare_sf) +
  tm_dots(size = 0.08, fill = "#0072B2", fill_alpha = 0.7) +
  tm_title("Childcare centres over the 2019 subzone boundaries", size = 1.5, fontface = "bold") +
  tm_credits("Source: data.gov.sg", position = c("left", "bottom"))
```

## The Code

```{r}
#| eval: false
tm_shape(mpsz_cl) +
  tm_polygons(col = "grey85", fill = "#F7F4EC") +
  tm_shape(childcare_sf) +
  tm_dots(size = 0.08, fill = "#0072B2", fill_alpha = 0.7) +
  tm_title("Childcare centres over the 2019 subzone boundaries", size = 1.5, fontface = "bold") +
  tm_credits("Source: data.gov.sg", position = c("left", "bottom"))
```
:::

Both layers line up on the same extent, which confirms they share a coordinate system.

## Geospatial data wrangling

spatstat does not work with sf objects. It needs a `ppp` for the points and an `owin` for the region they can fall in.

::: panel-tabset
## Converting sf data frames to ppp class

`as.ppp()` of **spatstat** turns the sf points into a `ppp` object, and `class()` confirms it.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
class(childcare_ppp)
```

`summary()` shows 1925 points, with every attribute carried along as a mark.

```{r}
summary(childcare_ppp)
```

## Creating the owin and combining

`as.owin()` turns the cleaned subzone polygons into the observation window. Indexing the `ppp` by that window keeps only the centres inside it.

```{r}
sg_owin <- as.owin(mpsz_cl)
childcareSG_ppp <- childcare_ppp[sg_owin]
childcareSG_ppp
```

## Checking for duplicates

`duplicated()` flags points that share an exact location, and `any()` reduces that to one answer.

```{r}
any(duplicated(childcareSG_ppp))
```

Duplicated points break several of the distance methods, because the theory does not allow a nearest neighbour at distance zero. Here the answer is FALSE, so there is nothing to jitter.
:::

::: {.callout-note}
## Learning diary
The `owin` matters more than I first thought. Every later function measures distances from each point to its neighbours. Near the edge, some neighbours fall outside and are never counted, which biases the answer downward.

Floch, Marcon and Puech call this the edge effect and spend a section on it. This is why `Gest()`, `Kest()` and the others all take a `correction` argument. Before the reading, I would have ignored those arguments.
:::

# Clark-Evans Test for Nearest Neighbour Analysis

The Clark-Evans index R compares the average nearest neighbour distance with its expected value under complete spatial randomness. Below 1 means clustered, above 1 means spread out, and around 1 means random.

- H0: childcare centres are randomly distributed.
- H1: childcare centres are not randomly distributed.

::: panel-tabset
## Without CSR simulation

`clarkevans.test()` of **spatstat** runs the test, here with the p-value from a normal approximation and a one-sided alternative.

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                alternative = "clustered")
```

## With CSR simulation

This time the p-value comes from simulation, not a normal approximation. `method = "MonteCarlo"` with `nsim = 99` compares R against 99 random patterns.

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                alternative = "clustered",
                method = "MonteCarlo",
                nsim = 99)
```
:::

::: {.callout-note}
## Learning diary
The two p-values differ, as they must. The Monte Carlo version uses 99 simulated patterns, so its smallest possible p-value is 0.01. This floor is set by the number of simulations, so the result is not more or less significant than the first.

The required reading gives the rule. For a two-sided test, alpha is 2 divided by the number of simulations plus 1, so 39 simulations gives 5 percent. Every simulation count is therefore a choice about alpha, and the book changes it between sections without comment.
:::

::: {.callout-caution}
## Where I differ from the book
R comes out well below 1 and the p-value is tiny, so the test says clustered. I am not sure that tells me much.

The null here is that a childcare centre is equally likely to be anywhere in Singapore, including reservoirs, the airport, industrial land and military areas. But childcare centres are built where children live.

Floch et al. make the same point about French pharmacies, whose locations are regulated by population. They conclude the CSR reference "does not appear to be the most relevant in this case", and I think the same applies here. I return to this in Beyond the Exercise with a more realistic null.
:::

# Kernel Density Estimation Method

Clark-Evans gives one number for the whole island. KDE gives a surface, so it can also show *where* the centres concentrate.

## Working with automatic bandwidth selection method

`density()` of **spatstat** computes the kernel density surface. `sigma = bw.diggle` lets the data pick the bandwidth, `edge = TRUE` corrects for the window edge, and the kernel is Gaussian.

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
```

`plot()` draws the resulting `im` object, with the colour ribbon on the right giving centres per square metre.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(0.5, 0.5, 2, 3))
plot(kde_childcareSG_bw,
     main = "Childcare centres per sq m, bw.diggle bandwidth")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(0.5, 0.5, 2, 3))
plot(kde_childcareSG_bw,
     main = "Childcare centres per sq m, bw.diggle bandwidth")
```
:::

Calling `bw.diggle()` directly prints the bandwidth it chose, in metres.

```{r}
bw.diggle(childcareSG_ppp)
```

## Rescaling KDE values

The density values are in points per square metre, so the legend shows numbers like 3e-05. Rescaling to kilometres makes them easier to read without changing the pattern. `rescale.ppp()` converts the coordinates to km, and `density()` runs again.

```{r}
childcareSG_ppp.km <- rescale.ppp(childcareSG_ppp, 1000, "km")

kde_childcareSG.km <- density(childcareSG_ppp.km,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
```

This is the same map, with the ribbon now in centres per square kilometre.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(0.5, 0.5, 2, 3))
plot(kde_childcareSG.km,
     main = "Same surface, now in centres per square kilometre")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(0.5, 0.5, 2, 3))
plot(kde_childcareSG.km,
     main = "Same surface, now in centres per square kilometre")
```
:::

## Working with different automatic bandwidth methods

`bw.CvL()`, `bw.scott()`, `bw.ppl()` and `bw.diggle()` are four rules for choosing sigma. Each prints its choice in km.

```{r}
bw.CvL(childcareSG_ppp.km)
bw.scott(childcareSG_ppp.km)
bw.ppl(childcareSG_ppp.km)
bw.diggle(childcareSG_ppp.km)
```

Here are the diggle and ppl surfaces side by side, on one shared colour scale.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.8
kde_ppl <- density(childcareSG_ppp.km, sigma = bw.ppl,
                   edge = TRUE, kernel = "gaussian")
zl <- range(c(kde_childcareSG.km$v, kde_ppl$v), na.rm = TRUE)
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.3)
plot(kde_childcareSG.km, main = "bw.diggle, tighter (per km sq)", zlim = zl)
plot(kde_ppl,            main = "bw.ppl, smoother (per km sq)",   zlim = zl)
mtext("Two automatic bandwidths, one colour scale", outer = TRUE, cex = 1.5, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
kde_ppl <- density(childcareSG_ppp.km, sigma = bw.ppl,
                   edge = TRUE, kernel = "gaussian")
zl <- range(c(kde_childcareSG.km$v, kde_ppl$v), na.rm = TRUE)
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.3)
plot(kde_childcareSG.km, main = "bw.diggle, tighter (per km sq)", zlim = zl)
plot(kde_ppl,            main = "bw.ppl, smoother (per km sq)",   zlim = zl)
mtext("Two automatic bandwidths, one colour scale", outer = TRUE, cex = 1.5, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

::: {.callout-note}
## Learning diary
The four rules differ by a factor of about fifteen, from 4.36 with CvL (the widest) to 0.30 with diggle (the narrowest). The maps they produce look quite different.

Baddeley recommends ppl for patterns of mostly tight clusters, and diggle for finding one cluster in noise. Either way, I choose the bandwidth and the data does not. The reading is more direct. The kernel barely matters, but the bandwidth matters a lot.

As with the classification method in Exercise 1, it looks like a default setting but changes what the map shows.
:::

## Working with different kernel methods

`density()` again with sigma held at 0.6 km, swapping the `kernel` argument through the four shapes it offers. All four panels share one colour scale.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
kde_gau <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "gaussian")
kde_epa <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "epanechnikov")
kde_qua <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "quartic")
kde_dsc <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "disc")
zl <- range(c(kde_gau$v, kde_epa$v, kde_qua$v, kde_dsc$v), na.rm = TRUE)

par(mfrow = c(2, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(kde_gau, main = "Gaussian (per km sq)",     zlim = zl)
plot(kde_epa, main = "Epanechnikov (per km sq)", zlim = zl)
plot(kde_qua, main = "Quartic (per km sq)",      zlim = zl)
plot(kde_dsc, main = "Disc (per km sq)",         zlim = zl)
mtext("Four kernels at a 600 m bandwidth, one colour scale", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
kde_gau <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "gaussian")
kde_epa <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "epanechnikov")
kde_qua <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "quartic")
kde_dsc <- density(childcareSG_ppp.km, sigma = 0.6, edge = TRUE,
                   kernel = "disc")
zl <- range(c(kde_gau$v, kde_epa$v, kde_qua$v, kde_dsc$v), na.rm = TRUE)

par(mfrow = c(2, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(kde_gau, main = "Gaussian (per km sq)",     zlim = zl)
plot(kde_epa, main = "Epanechnikov (per km sq)", zlim = zl)
plot(kde_qua, main = "Quartic (per km sq)",      zlim = zl)
plot(kde_dsc, main = "Disc (per km sq)",         zlim = zl)
mtext("Four kernels at a 600 m bandwidth, one colour scale", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

With the bandwidth at 0.6, the four kernels give nearly identical maps. This matches Floch et al., who said the kernel choice has limited impact.

# Fixed and Adaptive KDE

A single bandwidth assumes the pattern is equally dense everywhere, which is not true here. An adaptive bandwidth narrows the window where points are dense and widens it where they are sparse.

`density()` with `sigma = 0.6` gives the fixed version and `adaptive.density()` of **spatstat** the adaptive one. Both panels share one colour scale so their peaks can be compared.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.8
kde_fb <- density(childcareSG_ppp.km, sigma = 0.6,
                  edge = TRUE, kernel = "gaussian")
kde_ab <- adaptive.density(childcareSG_ppp.km, method = "kernel")

zl <- range(c(kde_fb$v, kde_ab$v), na.rm = TRUE)
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.3)
plot(kde_fb, main = "Fixed bandwidth, 600 m (per km sq)", zlim = zl)
plot(kde_ab, main = "Adaptive bandwidth (per km sq)",     zlim = zl)
mtext("Fixed against adaptive bandwidth, one colour scale", outer = TRUE, cex = 1.5, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
kde_fb <- density(childcareSG_ppp.km, sigma = 0.6,
                  edge = TRUE, kernel = "gaussian")
kde_ab <- adaptive.density(childcareSG_ppp.km, method = "kernel")

zl <- range(c(kde_fb$v, kde_ab$v), na.rm = TRUE)
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.3)
plot(kde_fb, main = "Fixed bandwidth, 600 m (per km sq)", zlim = zl)
plot(kde_ab, main = "Adaptive bandwidth (per km sq)",     zlim = zl)
mtext("Fixed against adaptive bandwidth, one colour scale", outer = TRUE, cex = 1.5, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

# Plotting Cartographic Quality KDE Map

## Converting gridded output into raster

tmap cannot draw spatstat's `im` object, so `rast()` of **terra** converts it. The surface was computed in km, so `ext()` multiplies the extent by 1000 to put the coordinates back in metres. Then `crs()` assigns EPSG 3414. The cell values stay in centres per km sq.

```{r}
kde_terra <- rast(kde_childcareSG.km)
ext(kde_terra) <- as.vector(ext(kde_terra)) * 1000
crs(kde_terra) <- "EPSG:3414"
kde_terra
```

::: {.callout-caution}
## Where I differ from the book
My first version followed the book and assigned EPSG 3414 directly to the raster built from the km surface. EPSG 3414 is in metres, so that raster was a thousand times too small. It was about 53 by 29 metres, at the grid's zero point south-west of Singapore. The graticules showed it, placing the whole island between 103.5818 and 103.5822 E.

Scaling the extent by 1000 fixes the coordinates and leaves the values unchanged, so the legend still reads in centres per km sq.
:::

## Plotting KDE map with tmap

`tm_raster()` of **tmap** draws the raster, with the legend outside the map on the right and the compass in the empty top left corner.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
tm_shape(kde_terra) +
  tm_raster(col.scale = tm_scale_continuous(values = "-magma"),
            col.legend = tm_legend(title = "Centres per km sq",
                                   position = tm_pos_out("right", "center"))) +
  tm_graticules(labels.size = 0.7) +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_title("Where childcare provision concentrates", size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tm_shape(kde_terra) +
  tm_raster(col.scale = tm_scale_continuous(values = "-magma"),
            col.legend = tm_legend(title = "Centres per km sq",
                                   position = tm_pos_out("right", "center"))) +
  tm_graticules(labels.size = 0.7) +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_title("Where childcare provision concentrates", size = 1.5, fontface = "bold")
```
:::

# First Order SPPA at the Planning Area Level

The island-wide view hides detail, so I zoom in on four planning areas chosen by the book, where individual centres are visible. The book calls this the planning subzone level, but all four units are planning areas, so I use that name.

## Geospatial data wrangling

::: panel-tabset
## Extracting study area

`filter()` pulls each planning area out of the cleaned subzones.

```{r}
pg <- mpsz_cl %>% filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>% filter(PLN_AREA_N == "JURONG WEST")
```

## Creating owin objects and combining

`as.owin()` turns each area into an `owin`, and the centres are clipped to it. `rescale.ppp()` then converts each pattern to km for the density maps.

```{r}
childcare_pg_ppp <- childcare_ppp[as.owin(pg)]
childcare_tm_ppp <- childcare_ppp[as.owin(tm)]
childcare_ck_ppp <- childcare_ppp[as.owin(ck)]
childcare_jw_ppp <- childcare_ppp[as.owin(jw)]

childcare_pg_ppp.km <- rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km <- rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km <- rescale.ppp(childcare_jw_ppp, 1000, "km")
```

## Counting centres per area

`npoints()` counts the centres that landed inside each window, collected into one `tibble()`.

```{r}
tibble(
  area   = c("Punggol", "Tampines", "Choa Chu Kang", "Jurong West"),
  centres = c(npoints(childcare_pg_ppp), npoints(childcare_tm_ppp),
              npoints(childcare_ck_ppp), npoints(childcare_jw_ppp))
)
```
:::

`unmark()` drops the attribute columns, which would otherwise set the symbols, so `plot()` shows just the centres.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(unmark(childcare_pg_ppp.km), main = "Punggol",       pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_tm_ppp.km), main = "Tampines",      pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_ck_ppp.km), main = "Choa Chu Kang", pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_jw_ppp.km), main = "Jurong West",   pch = 20, cex = 0.7, cols = "#0072B2")
mtext("Childcare centres in the four planning areas", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(unmark(childcare_pg_ppp.km), main = "Punggol",       pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_tm_ppp.km), main = "Tampines",      pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_ck_ppp.km), main = "Choa Chu Kang", pch = 20, cex = 0.7, cols = "#0072B2")
plot(unmark(childcare_jw_ppp.km), main = "Jurong West",   pch = 20, cex = 0.7, cols = "#0072B2")
mtext("Childcare centres in the four planning areas", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

## Clark and Evans Test

I run `clarkevans.test()` again, now two-sided, for the two areas the book picks.

::: panel-tabset
## Choa Chu Kang planning area

Choa Chu Kang first, with `alternative = "two.sided"`.

```{r}
clarkevans.test(childcare_ck_ppp, correction = "none",
                alternative = "two.sided")
```

## Tampines planning area

Tampines with the same settings.

```{r}
clarkevans.test(childcare_tm_ppp, correction = "none",
                alternative = "two.sided")
```
:::

::: {.callout-caution}
## Where I differ from the book
The book passes `nsim = 999` to both of these calls. `clarkevans.test()` only uses `nsim` when `method = "MonteCarlo"`. So those 999 simulations never run, and the header of each result still says Z-test. Nothing I read from these results depends on simulation, so I dropped the unused argument.
:::

::: {.callout-note}
## Learning diary
The alternative changed from `"clustered"` to `"two.sided"` between the island test and these two. A one-sided test can only find clustering, not a pattern that is unusually spread out. A two-sided test can find either, but needs a stronger signal for the same p-value.

Neither the book nor I pointed this out at first. The direction is part of the test, so it should be set before looking at the data.
:::

## Computing KDE surfaces by planning area

`density()` with `bw.diggle` runs for each area, drawn as a 2 by 2 grid. Each panel has its own colour scale, because Tampines peaks far higher and one shared scale would wash out the other three. So compare shapes within each panel, not colours across panels.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
par(mfrow = c(2, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(density(childcare_pg_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Punggol (per km sq)")
plot(density(childcare_tm_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Tampines (per km sq)")
plot(density(childcare_ck_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Choa Chu Kang (per km sq)")
plot(density(childcare_jw_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Jurong West (per km sq)")
mtext("Kernel density in the four planning areas, each on its own scale", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```

## The Code

```{r}
#| eval: false
par(mfrow = c(2, 2), mar = c(0.5, 0.5, 2, 3), oma = c(0, 0, 3, 0), cex.main = 1.5)
plot(density(childcare_pg_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Punggol (per km sq)")
plot(density(childcare_tm_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Tampines (per km sq)")
plot(density(childcare_ck_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Choa Chu Kang (per km sq)")
plot(density(childcare_jw_ppp.km, sigma = bw.diggle, edge = TRUE,
             kernel = "gaussian"), main = "Jurong West (per km sq)")
mtext("Kernel density in the four planning areas, each on its own scale", outer = TRUE, cex = 1.8, font = 2)
par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
```
:::

# Second-order Spatial Point Patterns Analysis

First order analysis asked how many points are in each place. Second order asks whether points sit closer to each other than chance would put them. I use four distance functions, each compared against simulated randomness.

`set.seed()` fixes the random number stream, so the simulated envelopes below come out the same on every render.

```{r}
set.seed(1234)
```

# Analysing Spatial Point Process Using G-Function

G takes the distance from each point to its nearest neighbour and plots their cumulative distribution. A curve above the theoretical line means neighbours are closer than random, which points to clustering.

## Choa Chu Kang planning area

`Gest()` of **spatstat** computes G with the border correction. The black line is the estimate and the orange dashed line is G under CSR.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_CK <- Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim = c(0, 500), xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "G function, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_CK <- Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim = c(0, 500), xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "G function, Choa Chu Kang")
```
:::

- H0: childcare centres in Choa Chu Kang are randomly distributed.
- H1: they are not.

::: {.callout-caution}
## Where I differ from the book
The book says to reject if p is below 0.001. But an envelope never prints a p-value, and 999 simulations cannot reach 0.001 anyway. By the rule in the Clark-Evans diary, a pointwise envelope from 999 simulations at the default rank of 1 is a two-sided test at alpha = 2/1000 = 0.002 at each distance.

So I reject at a distance if the black curve leaves the grey band there, at 0.002 rather than 0.001.
:::

`envelope()` of **spatstat** simulates 999 CSR patterns in the same window and computes G for each. Black is the observed G, orange dashed is G under CSR and grey is the range of the simulations.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999, verbose = FALSE)
plot(G_CK.csr, xlab = "d (m)", col = c("black", "#D55E00", "grey", "grey"),
     main = "G against 999 simulations of CSR, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999, verbose = FALSE)
plot(G_CK.csr, xlab = "d (m)", col = c("black", "#D55E00", "grey", "grey"),
     main = "G against 999 simulations of CSR, Choa Chu Kang")
```
:::

The black curve leaves the band only at the shortest distances. It jumps to about 0.19 within the first metre or two. So roughly a fifth of the centres have another centre almost on top of them, most likely two operators in the same block. The band catches up at about 45 m, and after that the curve stays inside it.

::: {.callout-caution}
## Where I differ from the book
The grey band looks like a confidence interval and is often read as one, but it is not.

Floch et al. say this directly. "The envelope should not be interpreted as a confidence interval around the indicator being studied: it indicates the critical values of the test." The band is the range the simulated patterns covered, not a range of plausible values for my curve.

So my plot titles name the simulations instead of calling it a confidence band, and I read every grey band on this page as critical values.
:::

## Tampines planning area

`Gest()` with `correction = "best"` lets spatstat choose the edge correction, which for G is Kaplan-Meier.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_tm <- Gest(childcare_tm_ppp, correction = "best")
plot(G_tm, xlab = "d (m)", col = c("black", "#D55E00"),
     main = "G function, Tampines")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_tm <- Gest(childcare_tm_ppp, correction = "best")
plot(G_tm, xlab = "d (m)", col = c("black", "#D55E00"),
     main = "G function, Tampines")
```
:::

# Analysing Spatial Point Process Using F-Function

F works the other way round. It places a grid of arbitrary locations and measures from each one to the nearest point. So it describes empty space rather than occupied space.

## Choa Chu Kang planning area

`Fest()` of **spatstat** computes F with its three default edge corrections, drawn together with F under CSR.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_CK <- Fest(childcare_ck_ppp)
plot(F_CK, xlab = "d (m)",
     col = c("black", "#D55E00", "#0072B2", "#CC79A7"),
     main = "F function, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_CK <- Fest(childcare_ck_ppp)
plot(F_CK, xlab = "d (m)",
     col = c("black", "#D55E00", "#0072B2", "#CC79A7"),
     main = "F function, Choa Chu Kang")
```
:::

## Performing Complete Spatial Randomness Test

`envelope()` with `Fest` runs the same 999-simulation test for F. Black is observed, orange dashed is CSR and grey is the range of the simulations.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999, verbose = FALSE)
plot(F_CK.csr, xlab = "d (m)", col = c("black", "#D55E00", "grey", "grey"),
     main = "F against 999 simulations of CSR, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999, verbose = FALSE)
plot(F_CK.csr, xlab = "d (m)", col = c("black", "#D55E00", "grey", "grey"),
     main = "F against 999 simulations of CSR, Choa Chu Kang")
```
:::

Observed F runs below the CSR line, which is the direction clustering predicts. But it stays inside the grey band at every distance, so F by itself finds no departure from CSR.

::: {.callout-note}
## Learning diary
G and F point opposite ways for the same pattern, which confused me until I drew it out.

Clustered points have close neighbours, so G climbs fast. But clustering also leaves big empty gaps, so a random location is far from any point and F climbs slowly. So clustering puts G above the line and F below it. Once I understood that, the plots stopped looking contradictory.
:::

## Tampines planning area

`Fest()` with `correction = "best"` for Tampines.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_tm <- Fest(childcare_tm_ppp, correction = "best")
plot(F_tm, xlab = "d (m)", col = c("black", "#D55E00"),
     main = "F function, Tampines")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
F_tm <- Fest(childcare_tm_ppp, correction = "best")
plot(F_tm, xlab = "d (m)", col = c("black", "#D55E00"),
     main = "F function, Tampines")
```
:::

# Analysing Spatial Point Process Using K-Function

G and F only look at the nearest neighbour. K counts every neighbour within distance r, for every r, so it works at all scales at once.

Under CSR, K(d) equals pi d^2. So I plot K(d) minus pi d^2, divided by a million so it reads in km sq. Zero means random. Above zero means more neighbours than chance would give, and below zero means fewer.

::: {.callout-caution}
## Where I differ from the book
The book plots K with the formula `. - r ~ r` and labels it K(d) - r. But K is an area and r is a length. Subtracting r only moves the curve down by a few hundred square metres, against values in the millions. The result is still just K, rising like a parabola, not centred on zero.

Subtracting r makes sense for L, which is already a distance. For K the matching reference is its CSR value, pi d^2, so I subtract that with the formula `(. - pi * r^2) / 1e6 ~ r`.
:::

## Choa Chu Kang planning area

`Kest()` of **spatstat** with Ripley's isotropic correction. The orange dashed line is CSR, flat at zero.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, (. - pi * r^2) / 1e6 ~ r,
     ylab = "K(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "K function, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, (. - pi * r^2) / 1e6 ~ r,
     ylab = "K(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "K function, Choa Chu Kang")
```
:::

`envelope()` with `Kest`, 99 simulations and `global = TRUE` builds one simultaneous band for all distances. Black is the observed K, orange dashed at zero is CSR and grey is the global envelope. The legend is off so it cannot cover the band.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1,
                     global = TRUE, verbose = FALSE)
plot(K_ck.csr, (. - pi * r^2) / 1e6 ~ r, xlab = "d (m)",
     ylab = "K(d) - pi d^2 (km sq)", legend = FALSE,
     col = c("black", "#D55E00", "grey", "grey"),
     main = "K against CSR, global envelope, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1,
                     global = TRUE, verbose = FALSE)
plot(K_ck.csr, (. - pi * r^2) / 1e6 ~ r, xlab = "d (m)",
     ylab = "K(d) - pi d^2 (km sq)", legend = FALSE,
     col = c("black", "#D55E00", "grey", "grey"),
     main = "K against CSR, global envelope, Choa Chu Kang")
```
:::

The observed curve rises above zero, to about 0.3 km sq by 800 m. But the global band is about 0.58 km sq wide on each side at every distance, so the curve never comes close to leaving it. Tested at all distances together, K does not reject CSR for Choa Chu Kang.

::: {.callout-caution}
## Where I differ from the book
The book writes `glocal = TRUE` here and for L, a typo for `global = TRUE`. `envelope()` has no `glocal` argument, so R passes it through `...` to `Kest()`, which silently ignores it. So the book's K and L envelopes are pointwise, even though the code seems to ask for global ones. I copied it at first without noticing, which shows it is easy to miss. I use `global = TRUE`.

Global matters because the test is repeated at dozens of r values, and reporting the ones that cross pushes the false positive rate well above the stated alpha. A global envelope corrects for that and a pointwise one does not. With `rank = 1` and 99 simulations, the global band is a single test at alpha = 1/100 across all distances at once.

A global band contains the pointwise band from the same simulations, so it is harder for the curve to leave. The G and F results above use pointwise bands, so they are weaker evidence than the K and L results here. The book never says so. I redo G with a global band in Beyond the Exercise.
:::

## Tampines planning area

`Kest()` for Tampines, cut at 1000 m.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, (. - pi * r^2) / 1e6 ~ r,
     ylab = "K(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     xlim = c(0, 1000), main = "K function, Tampines")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, (. - pi * r^2) / 1e6 ~ r,
     ylab = "K(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     xlim = c(0, 1000), main = "K function, Tampines")
```
:::

# Analysing Spatial Point Process Using L-Function

L is a transformed K. Under randomness K grows as pi r squared, a curve that is hard to judge by eye. L is the square root of K over pi, so under randomness it is a straight line and `L(r) - r` is flat at zero.

## Choa Chu Kang planning area

`Lest()` of **spatstat** computes L directly, plotted as L(d) - d with CSR as the orange dashed line at zero.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "L function, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "L function, Choa Chu Kang")
```
:::

::: {.callout-note}
## Learning diary
Floch et al. explain `L(r) - r` in a way that finally made sense to me. If the curve reaches 2 at a radius of 1, each point has as many neighbours within 1 unit as a random pattern would have within 3 units.

So the height of the curve is how much further you would need to walk, under randomness, to meet the same number of neighbours.
:::

`envelope()` with `Lest`, again 99 simulations with `global = TRUE`. Black is the observed L(d) - d, orange dashed at zero is CSR and grey is the global envelope. The legend is off so it cannot hide the lower half of the band.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1,
                     global = TRUE, verbose = FALSE)
plot(L_ck.csr, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r",
     legend = FALSE, col = c("black", "#D55E00", "grey", "grey"),
     main = "L against CSR, global envelope, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1,
                     global = TRUE, verbose = FALSE)
plot(L_ck.csr, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r",
     legend = FALSE, col = c("black", "#D55E00", "grey", "grey"),
     main = "L against CSR, global envelope, Choa Chu Kang")
```
:::

L gives the same answer. The biggest excursion is the spike at the very start, about 70 m against a band edge near 75 m, from the same near-duplicate centres that G picked up. The curve never leaves the band.

## Tampines planning area

`Lest()` for Tampines, cut at 1000 m.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)",
     col = c("black", "#D55E00"),
     xlim = c(0, 1000), main = "L function, Tampines")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)",
     col = c("black", "#D55E00"),
     xlim = c(0, 1000), main = "L function, Tampines")
```
:::

# Beyond the Exercise

By the end of the chapter sections, four things bothered me. All four came from the required reading, not from me.

## Testing Against a Null That Is Not a Strawman

Everything so far compares childcare centres to a pattern where any point in Singapore is equally likely, including reservoirs, the airport, industrial estates and military land. At least Jurong Island was dropped with the offshore subzones at the start.

Floch et al. raise this for French pharmacies, where regulation ties locations to population. They say CSR "does not appear to be the most relevant" reference there. Childcare centres similarly follow housing. So the clustering result may only tell me that Singapore's residents are clustered, which I already knew.

The reading suggests a fix. `Kinhom()` drops the assumption of constant intensity and compares the pattern against a null that uses the observed intensity. If the centres still clump beyond that, the clustering is real. If not, it was caused by population.

`density()` with `bw.diggle` estimates the intensity surface for Choa Chu Kang, and `range()` checks its values.

```{r}
lambda_ck <- density(childcare_ck_ppp, sigma = bw.diggle, edge = TRUE)
range(lambda_ck$v, na.rm = TRUE)
```

::: {.callout-note}
## Learning diary
My surface is entirely positive, but I clamp it anyway.

The book's own output in section 4.7.1 has a density range starting at -6.58e-21. That is floating point noise from the edge correction, not a real negative density, and it does not show on the plot.

It only matters when the surface goes to `rpoispp()` for the simulation below, which refuses a negative intensity with a "negative probability" error that does not say where the negative came from. Whether you hit it depends on your data.

Adding one line now is easier than debugging that later.
:::

The clamp sets any negative cell to zero, then counts how many there were.

```{r}
lambda_pos <- lambda_ck
lambda_pos$v[lambda_pos$v < 0] <- 0

# confirm whether the clamp actually had anything to do
sum(lambda_ck$v < 0, na.rm = TRUE)
```

`Kinhom()` of **spatstat** computes the inhomogeneous K with that surface as lambda, plotted against pi d^2 the same way as K above.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
Kinhom_ck <- Kinhom(childcare_ck_ppp, lambda = lambda_pos,
                    correction = "Ripley")
plot(Kinhom_ck, (. - pi * r^2) / 1e6 ~ r,
     ylab = "Kinhom(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "Inhomogeneous K, intensity taken from the data")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
Kinhom_ck <- Kinhom(childcare_ck_ppp, lambda = lambda_pos,
                    correction = "Ripley")
plot(Kinhom_ck, (. - pi * r^2) / 1e6 ~ r,
     ylab = "Kinhom(d) - pi d^2 (km sq)", xlab = "d (m)",
     col = c("black", "#D55E00"),
     main = "Inhomogeneous K, intensity taken from the data")
```
:::

With the intensity taken from the data, the curve falls below zero and keeps falling, to about -0.7 km sq by 800 m. Taken literally, this means the centres have fewer neighbours than the local intensity predicts.

`envelope()` with `Kinhom` and `simulate = expression(rpoispp(lambda_pos))` draws 39 patterns from the estimated intensity instead of from CSR. Black is observed, orange dashed is the mean of the simulated patterns and grey is their pointwise range. The legend is off so it cannot cover the band.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
Kinhom_ck.csr <- envelope(childcare_ck_ppp, Kinhom,
                          simulate = expression(rpoispp(lambda_pos)),
                          nsim = 39, verbose = FALSE)
plot(Kinhom_ck.csr, (. - pi * r^2) / 1e6 ~ r, xlab = "d (m)",
     ylab = "Kinhom(d) - pi d^2 (km sq)", legend = FALSE,
     col = c("black", "#D55E00", "grey", "grey"),
     main = "Inhomogeneous K against a population-shaped null")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
Kinhom_ck.csr <- envelope(childcare_ck_ppp, Kinhom,
                          simulate = expression(rpoispp(lambda_pos)),
                          nsim = 39, verbose = FALSE)
plot(Kinhom_ck.csr, (. - pi * r^2) / 1e6 ~ r, xlab = "d (m)",
     ylab = "Kinhom(d) - pi d^2 (km sq)", legend = FALSE,
     col = c("black", "#D55E00", "grey", "grey"),
     main = "Inhomogeneous K against a population-shaped null")
```
:::

Here the observed curve stays close to zero and inside the band at every distance. Once the local intensity is accounted for, Choa Chu Kang shows no extra clustering.

::: {.callout-important}
## Check that caught something
The black curve here ends near -0.05 km sq, but the previous plot of the same data ended near -0.7. They are different estimates. In the earlier plot I passed `lambda_pos`, built from these same points with a narrow bw.diggle bandwidth. So each centre's intensity includes its own kernel and comes out too high, which pulls K down. Inside `envelope()` I did not pass lambda, so `Kinhom()` estimated it itself and left each point out of its own intensity. That leave-one-out estimate is the one to trust, so the deep dip earlier is mostly an artefact.
:::

::: {.callout-tip}
## Alternative explored
This uses the same K function for a different question. `Kest` asks whether centres are more clustered than a uniform scatter. `Kinhom` asks whether they are more clustered than the local intensity already implies.

Compare this plot with the `Kest` one further up. Any gap left here is clustering that population density does not explain.

One caveat is that the null comes from the same data it is tested against, so it is not independent evidence. Floch et al. warn that bias appears when the smoothing window is close in size to the r values being tested.
:::

## Cumulative Against Density

K counts every neighbour out to r, so it is cumulative. A bump at 400 metres could come from tight clustering at 50 metres still carried in the total.

The pair correlation function counts in a ring at distance r instead of a disc out to r. So it shows the distance where clustering happens, not where it has built up.

`pcf()` of **spatstat** estimates it. It shoots up near zero distance, so I cap the y axis at 4 to keep the rest readable.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
pcf_ck <- pcf(childcare_ck_ppp)
plot(pcf_ck, main = "Pair correlation, Choa Chu Kang",
     xlab = "d (m)", ylim = c(0, 4), legendpos = "topright",
     col = c("black", "#0072B2", "#D55E00"))
abline(h = 1, lty = 2, col = "#D55E00")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
pcf_ck <- pcf(childcare_ck_ppp)
plot(pcf_ck, main = "Pair correlation, Choa Chu Kang",
     xlab = "d (m)", ylim = c(0, 4), legendpos = "topright",
     col = c("black", "#0072B2", "#D55E00"))
abline(h = 1, lty = 2, col = "#D55E00")
```
:::

::: {.callout-tip}
## Alternative explored
Under randomness the pair correlation is 1, shown by the dashed line. Above 1 means more neighbours at that exact distance than chance would give.

Its advantage over K is that it pins down the distance of the clustering. Its disadvantage is that only a cumulative function can detect clusters of clusters. They answer different questions, so one does not replace the other.
:::

## Redoing G With a Global Envelope

The earlier G test used a pointwise envelope, which inflates the false positive rate across the r values tested. Here I rerun it with `envelope()` set to `global = TRUE` and `rank = 1` over 39 simulations.

::: panel-tabset
## The Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_ck.global <- envelope(childcare_ck_ppp, Gest, nsim = 39, rank = 1,
                        global = TRUE, verbose = FALSE)
plot(G_ck.global, xlab = "d (m)", ylim = c(0, 1),
     col = c("black", "#D55E00", "grey", "grey"),
     main = "G with a global envelope, Choa Chu Kang")
```

## The Code

```{r}
#| eval: false
par(cex.main = 1.3, cex.sub = 1.1, cex.lab = 1.1, mar = c(5, 5, 4, 2))
G_ck.global <- envelope(childcare_ck_ppp, Gest, nsim = 39, rank = 1,
                        global = TRUE, verbose = FALSE)
plot(G_ck.global, xlab = "d (m)", ylim = c(0, 1),
     col = c("black", "#D55E00", "grey", "grey"),
     main = "G with a global envelope, Choa Chu Kang")
```
:::

The curve stays inside the global band at every distance, including the short-distance jump that left the pointwise band. The band's upper edge at zero distance is already about 0.27, above the 0.19 jump.

::: {.callout-tip}
## Alternative explored
The correction makes a global band wider than a pointwise one, so it is harder to reject the null, as intended.

If the curve still leaves the band, the result survives the multiple testing problem. If not, the original finding was partly an artefact of testing at every r and reporting the crossings.
:::

## A Second Opinion on Randomness

All the tests so far are distance based. A quadrat test, from the other family, divides the region into cells and compares the counts against a Poisson expectation. `quadrat.test()` of **spatstat** runs it on a 4 by 4 grid.

```{r}
quadrat_ck <- quadrat.test(childcare_ck_ppp, nx = 4, ny = 4)
quadrat_ck
```

::: {.callout-warning}
## Issue I hit
I first stored this as `qt`, which is base R's t-distribution quantile function. R refused with "cannot change value of locked binding".

It is the same mistake as naming something `c` or `df`, so I should check a name is free first.
:::

::: {.callout-important}
## Check that caught something
The quadrat test does not agree with Clark-Evans.

Clark-Evans gave R of 0.84 with p of 0.0089, rejecting randomness for Choa Chu Kang. The quadrat test on the same points gives p of 0.084, so it does not. On one dataset, the two tests reach opposite conclusions at the five percent level.

Neither is wrong, because they measure different things. Clark-Evans looks at nearest neighbour distances, so it notices points in tight pairs. The quadrat test only counts points per cell, so a pattern can be clumpy inside cells and still even across them.

The tile count also changed. I asked for a four by four grid, which is sixteen cells, but the output says thirteen tiles and twelve degrees of freedom. Three cells fell outside the planning area boundary and were dropped, and twelve is thirteen minus one. So the tested grid is not the one I specified.
:::

::: {.callout-tip}
## Alternative explored
The quadrat result depends entirely on the grid I picked. A different `nx` and `ny` would change it.

This is the Modifiable Areal Unit Problem again, which I met in Exercise 1 when aggregating subzones to planning areas. The reading gives it as one of the main reasons to prefer distance-based methods.
:::

# Limitations

- Every test in the chapter sections uses CSR as the null, a weak comparison for a facility that follows housing. Beyond the Exercise addresses this for one planning area only.
- The `Kinhom` null is estimated from the same data it tests. It asks a better question than CSR, but it is not independent.
- Simulation counts vary through the exercise, 99 here and 999 there, and each choice sets a different alpha.
- The four planning areas come from the book and are not a sample, so nothing here generalises to other planning areas.
- First and second order effects cannot be separated from a single observed pattern. I cannot fix this here, and it weakens the whole document, so it is also the second point in the reflections.

# Reflections

Five things I learned from this exercise.

**CSR was a weak null for childcare centres.** Every result in the chapter sections compares childcare centres against a pattern that could put them in a reservoir. The tests came back clustered with tiny p-values, but that may only reflect where people live. Floch et al. on French pharmacies made me see this, because the situation is the same.

**First and second order effects cannot be separated here.** Chapter 4 is first order (intensity) and Chapter 5 is second order (interaction). Floch et al. state that the two cannot be told apart from one observed pattern, citing work where natural advantage and genuine attraction give indistinguishable results. So I can say the centres are clustered, but not whether they are drawn to the same places or to each other.

**Some arguments I copied change the results.** `correction` exists because of edge effects, `global` because testing at every r inflates the error rate, and `nsim` sets alpha. The book's `glocal` typo showed me that R will quietly accept a misspelt argument that does nothing. I had been pasting all three as boilerplate.

**Clark-Evans and the quadrat test disagreed on Choa Chu Kang.** Clark-Evans rejected randomness and the quadrat test did not, on the same points at the same alpha. One reads distances between points and the other only counts them in boxes. Working this out taught me more about what each test can see than either result alone.

**Small coding issues took more time than the concepts.** A masking warning I ignored and a variable name that collided with a base R function both caused problems. Neither was hard, but neither showed up until something broke.
