---
title: "In-class Exercise 5"
subtitle: "Global and Local Measures of Spatial Autocorrelation"
author: "ZeXin Chen"
format:
  html:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 6
    fig-align: center
date: "September 26, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

# Overview

This in-class exercise uses the 88 counties of Hunan to measure spatial autocorrelation of GDP per capita with the **sfdep** package, first globally for the whole province and then locally for each county.

# Loading the Packages

```{r}
#| output: false
pacman::p_load(sf, sfdep, tmap, tidyverse, knitr, GWmodel, ggstatsplot)
```

# Preparing the Data

::: panel-tabset
## Importing Hunan shapefile

`st_read()` of **sf** reads the county boundaries into an sf polygon object.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

## Importing Hunan_2012 table

`read_csv()` of **readr** reads the 2012 indicators into a tibble.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```
:::

# Combining Both Data Frames by Using Left Join

`left_join()` of **dplyr** attaches the 2012 indicators to the county polygons by the shared `County` column. `select()` keeps columns 1 to 4, 7, 15 and 33.

```{r}
hunan_GDPPC <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15, 33)
```

::: {.callout-important}
## Learning diary
`hunan` has to be on the left of the join. The left data frame decides the type of the result, so putting the sf object first keeps the geometry and the output stays an sf data frame.
:::

# Global Measures of Spatial Association

## Step 1: Deriving Queen's Contiguity Weights with sfdep

`st_contiguity()` of **sfdep** builds the Queen neighbour list, and `st_weights()` turns it into weights. Both go into `hunan_GDPPC` as new columns, and `.before = 1` puts them at the front.

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

::: {.callout-note}
## Learning diary
`st_weights()` takes three inputs. `nb` is the neighbour list from `st_contiguity()`. `style` sets the weighting, and the default `"W"` row-standardises so each county's weights sum to 1. The other options are `"B"` binary, `"C"` globally standardised, `"U"` which is C divided by the number of neighbours, `"minmax"`, and `"S"` the variance-stabilising scheme of Tiefelsdorf et al. (1999). `allow_zero = TRUE` gives a lagged value of zero to a county with no neighbours.
:::

## Step 2: Computing Global Moran's I

`global_moran()` of **sfdep** computes Moran's I from the GDPPC values, the neighbour list and the weights. Unlike **spdep**, it returns its result as an object I can inspect with `glimpse()`, rather than a printed test report.

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)

glimpse(moranI)
```

::: {.callout-note}
## Learning diary
`I` is Moran's I, 0.301, the same value `moran.test()` gave in Hands-on Exercise 5, so sfdep and spdep agree. `K` is the sample kurtosis of GDPPC, 7.64. That is well above 3, the value for a normal distribution, so GDPPC has heavy tails, mostly from Changsha. This function only computes the statistic. It does not test it, so there is no p-value here yet.
:::

## Step 3: Performing Global Moran's I Permutation Test

In practice, Moran's I should be tested with a Monte Carlo simulation. `global_moran_perm()` of **sfdep** shuffles GDPPC across the counties `nsim` times and compares the real Moran's I with the shuffled ones. `set.seed()` fixes the shuffles so the result is the same on every run.

::: panel-tabset
## The Code

```{r}
#| eval: false
set.seed(1234)

global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

## The Report

```{r}
#| echo: false
set.seed(1234)

global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```
:::

::: {.callout-tip}
## Reminder
It is good practice to call `set.seed()` before any simulation, so the result can be reproduced. The number of simulations is always `nsim` + 1, so `nsim = 99` gives 100.
:::

# Local Measures of Spatial Association

## Computing Local Moran's I

`local_moran()` of **sfdep** computes local Moran's I of GDPPC for every county and tests each one with 99 simulations. It returns a data frame for each county, and `unnest()` of **tidyr** spreads it into ordinary columns.

::: panel-tabset
## The Code

```{r}
#| eval: false
set.seed(1234)

lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

## The Output

```{r}
#| echo: false
set.seed(1234)

lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)

glimpse(lisa)
```
:::

::: {.callout-note}
## Learning diary
The main columns are `ii` (local Moran's I), `eii` (its expected value), `var_ii` (its variance), `z_ii` (the standard deviate), `p_ii` (the p-value from the normal approximation), `p_ii_sim` (the p-value from the simulations) and `mean`, `median` and `pysal` (the quadrant, High-High, Low-Low, Low-High or High-Low). Unlike the global test, there is one value per county, so this shows where the clustering is, not just whether it exists.
:::

## Plotting LISA Map

The `lisa` sf data frame has three fields with the LISA categories, `mean`, `median` and `pysal`. In general `mean` is used. `filter()` keeps only the counties whose simulated p-value is below 0.05, and those are drawn on top of a grey map of all 88 counties.

::: panel-tabset
## The Plot

```{r}
#| echo: false
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)

tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons(fill = "grey90", col = "grey60") +
tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("#0072B2",    # Low-Low
                           "#CC79A7",    # High-Low
                           "#56B4E9",    # Low-High
                           "#D55E00")),  # High-High
              fill.legend = tm_legend(
                title = "LISA cluster",
                position = tm_pos_out("right", "center"),
                text.size = 1.15,
                width = 6),
              col = "grey60") +
  tm_title("Significant local clusters of GDPPC (p < 0.05)",
           size = 1.5, fontface = "bold") +
  tm_layout(legend.title.size = 1.25)
```

## The Code

```{r}
#| eval: false
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)

tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons(fill = "grey90", col = "grey60") +
tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("#0072B2",    # Low-Low
                           "#CC79A7",    # High-Low
                           "#56B4E9",    # Low-High
                           "#D55E00")),  # High-High
              fill.legend = tm_legend(
                title = "LISA cluster",
                position = tm_pos_out("right", "center"),
                text.size = 1.15,
                width = 6),
              col = "grey60") +
  tm_title("Significant local clusters of GDPPC (p < 0.05)",
           size = 1.5, fontface = "bold") +
  tm_layout(legend.title.size = 1.25)
```
:::

::: {.callout-caution}
## Where I differ from the prof's code
I made three small changes. The default palette uses green for High-High, and red with green is the pair colour-blind readers struggle with most, so I set blue, sky blue, pink and orange in the same class order. I moved the legend outside the map so it cannot cover a county, and added a title. I also dropped `fill_alpha` from `tm_borders()`, since borders have no fill, and set the border colour inside `tm_polygons()` instead.
:::

# Hot Spot and Cold Spot Area Analysis (HCSA)

## Computing Local Gi* Statistics

`include_self()` adds each county to its own neighbour list, which Gi* needs. `st_inverse_distance()` of **sfdep** then weights each neighbour by one over its distance, so closer counties count more.

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wts = st_inverse_distance(nb,
                                   geometry,
                                   scale = 1,
                                   alpha = 1),
         .before = 1)
```

`local_gstar_perm()` computes Gi* for every county and tests each one with 99 simulations.

```{r}
set.seed(1234)

HCSA <- wm_idw %>%
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wts, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
```

## Visualising p-value of Local Gi*

`tm_polygons()` maps `p_sim`, the simulated p-value of Gi*, with fixed breaks at the usual significance levels. Counties above 0.05 are grey.

::: panel-tabset
## The Plot

```{r}
#| echo: false
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 1.0),
      labels = c("p < 0.001", "p < 0.01", "p < 0.05", "Not Significant"),
      values = c("#de2d26", "#fc9272", "#fee0d2", "#e0e0e0")),
    fill.legend = tm_legend(
      title = "p-values",
      position = tm_pos_out("right", "center")),
    col = "grey60") +
  tm_title("p-values of Gi* for GDP per capita",
           size = 1.5, fontface = "bold")
```

## The Code

```{r}
#| eval: false
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 1.0),
      labels = c("p < 0.001", "p < 0.01", "p < 0.05", "Not Significant"),
      values = c("#de2d26", "#fc9272", "#fee0d2", "#e0e0e0")),
    fill.legend = tm_legend(
      title = "p-values",
      position = tm_pos_out("right", "center")),
    col = "grey60") +
  tm_title("p-values of Gi* for GDP per capita",
           size = 1.5, fontface = "bold")
```
:::

