---
title: "Where and When Pulang Pisau Burned"
subtitle: "Where fires burned, when, and whether one fire leads to the next"
author: "ZeXin Chen"
date: "2026-09-27"
date-format: "D MMMM YYYY"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    embed-resources: true
    transition: fade
    show-notes: false
    width: 1600
    height: 900
    footer: "ISSS626 Take-home Exercise 1 | ZeXin Chen | [Technical Report](Take-home_Ex01.html) | [Back to site](https://isss-626-gaa-six.vercel.app)"
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.googleapis.com">
        <link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
        <link href="https://fonts.googleapis.com/css2?family=Inter:wght@400;500;600;700&family=Newsreader:opsz,wght@6..72,400;6..72,600&display=swap" rel="stylesheet">
execute:
  echo: false
  eval: true
  warning: false
  message: false
  freeze: true
---

```{r}
#| label: setup
pacman::p_load(sf, spatstat, sparr, terra, tmap, tidyverse, knitr)
knitr::opts_chunk$set(out.width = "100%", dpi = 150, dev.args = list(bg = "transparent"))

prep   <- read_rds("data/rds/prepared.rds")
so     <- read_rds("data/rds/results_second_order.rds")
bonus  <- read_rds("data/rds/results_bonus.rds")
kal_rate <- read_rds("data/rds/kal_rate.rds")

pp_sf <- prep$pp_sf
fires <- prep$fires
ppp   <- prep$ppp

col_fire   <- "#D55E00"
col_accent <- "#0072B2"
col_band   <- "#E69F00"
col_land   <- "#F7F4EC"

to_rast <- function(im) {
  r <- rast(im)
  ext(r) <- as.vector(ext(r)) * 1000
  crs(r) <- "EPSG:32749"
  r
}

theme_slide <- theme_minimal(base_size = 18) +
  theme(plot.title = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", size = 15),
        plot.background = element_rect(fill = "#FDFAF3", colour = NA),
        panel.background = element_rect(fill = "#FDFAF3", colour = NA))
```

## Contents

::::: columns
::: {.column width="50%"}
**Part 1, The Problem and the Data**

- Why Pulang Pisau
- Cleaning the raw satellite data
- Six questions, each checked against chance

**Part 2, First-Order Analysis**

- Insight 1, when the fires came
- Insight 2, where they burned
:::

::: {.column width="50%"}
**Part 3, Second-Order Analysis**

- Insight 3, how clustered fires are
- Insight 4, nearby fires follow each other

**Part 4, Bonus**

- Insight 5, a daily watch zone for patrols

**Part 5, So What**

- Key findings and their limits
- Public-safety implications
:::
:::::

::: notes
This deck summarises the technical report. It moves from the problem and the data, to where and when fires burned, to how fires relate to each other, and ends with what the results mean for fire management in Pulang Pisau.
:::

## Pulang Pisau had the densest fire detections of any Kalimantan regency

<div class="part">Part 1 · The problem and the data</div>

::::: columns
::: {.column width="58%"}
```{r}
#| fig-width: 9
#| fig-height: 7.4
tm_shape(kal_rate) +
  tm_polygons(fill = "per100",
              fill.scale = tm_scale_intervals(style = "fixed",
                breaks = c(0, 2, 5, 10, 20, 35, 80), values = "brewer.oranges",
                labels = c("0 to 2", "2 to 5", "5 to 10", "10 to 20", "20 to 35", "35 to 80")),
              fill.legend = tm_legend(title = "Detections\nper 100 km²",
                position = tm_pos_out("right", "center")),
              col = "grey55", lwd = 0.4) +
  tm_shape(filter(kal_rate, NAME_2 == "PalangkaRaya")) +
  tm_borders(col = col_accent, lwd = 2.5) +
  tm_shape(filter(kal_rate, NAME_2 == "PulangPisau")) +
  tm_borders(col = "black", lwd = 3) +
  tm_add_legend(type = "lines", labels = c("Pulang Pisau", "Palangka Raya (city)"),
                col = c("black", col_accent), lwd = c(3, 2.5),
                position = tm_pos_out("right", "center")) +
  tm_layout(bg.color = "#FDFAF3", outer.bg.color = "#FDFAF3", frame = FALSE,
            legend.text.size = 1, legend.title.size = 1.2) +
  tm_credits("NASA FIRMS MODIS C6.1, 1 Jan to 22 Sep 2026", position = c("left", "bottom"), size = 0.9)
```
:::

::: {.column width="42%"}
<div class="kicker">The 2026 season</div>

<div class="bignum">3,693.7 ha</div>
<div class="bigcap">burned in 8 subdistricts by 17 Sep</div>

<div class="bignum">595 µg/m³</div>
<div class="bigcap">PM2.5 on 17 Sep, about 40 times the WHO 24-hour guideline (Berita Sampit, 2026; WHO, 2021)</div>

<div class="small">The regent raised the fire status to emergency response. Most of the regency is deep peat inside the failed Mega Rice Project, drained by canals in the 1990s.

**Research question.** Where and when did fires burn in 2026, and were fires closer together in space and time than uneven fuel alone would explain?</div>
:::
:::::

::: notes
Pulang Pisau is on the south coast of Central Kalimantan. It had the most fire detections per square kilometre of any regency in Kalimantan this year. Only Palangka Raya city next door was higher. Local reports put burned area at 3,694 hectares by 17 September and PM2.5 at a hazardous 595. The question is not just whether fires cluster, because fire maps always look clustered. It is whether they cluster beyond what the uneven spread of land that burns easily would give.
:::

## Three fixes turned 50,877 raw detections into 3,065 fire-days

<div class="part">Part 1 · The problem and the data</div>

::::: columns
::: {.column width="52%"}
```{r}
#| fig-width: 7.6
#| fig-height: 5.6
funnel <- prep$prep_log %>%
  mutate(step = c("Detections over Kalimantan", "Inside Pulang Pisau", "Confidence 30 or above",
                  "One per 1 km square per day", "1 Jul to 22 Sep 2026"),
         step = factor(step, levels = rev(step)),
         last = row_number() == n())

ggplot(funnel, aes(rows, step, fill = last)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = format(rows, big.mark = ",")), hjust = -0.12, size = 6.2, fontface = "bold") +
  scale_fill_manual(values = c("grey75", col_fire), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(x = NULL, y = NULL) +
  theme_slide +
  theme(panel.grid = element_blank(), axis.text.x = element_blank(),
        axis.text.y = element_text(size = 17, colour = "grey20"))
```
:::

::: {.column width="48%"}
<div class="fix"><b>Wrong day.</b> Satellite times are in UTC. 6,026 night detections were moved to the right local day.</div>

<div class="fix"><b>Counted up to four times.</b> Four passes a day can see the same fire, so repeats were merged to one event per 1 km square per day.</div>

<div class="fix"><b>False alarms.</b> Detections with confidence under 30 were dropped.</div>

<div class="small">A <b>fire-day</b> is one 1 km square burning on one local day. MODIS is the main sensor, and VIIRS is rerun as a check.</div>
:::
:::::

::: notes
The raw file had no missing or impossible coordinates, so the real work was elsewhere. Satellite times are in UTC, so night passes landed on the wrong day. The same fire is seen up to four times a day, so repeats were merged. Low confidence detections include false alarms, so they were dropped. That leaves 3,065 fire-days between 1 July and 22 September.
:::

## Six questions, each checked against what chance would give

<div class="part">Part 1 · The problem and the data</div>

| | Question | Method | Compared with |
|---|---|---|---|
| **Q1** | When and where did fires burn? | Kernel density, in space and in space-time | Nothing, it maps where fires were dense |
| **Q2** | Are fires closer together than random? | L function | Fires placed at random |
| **Q3** | Still closer once busy areas are allowed for? | Inhomogeneous L function | Random, but following the density map |
| **Q4** | Do nearby fires also burn close in time? | Knox test | Same places, dates shuffled within 10 km blocks |
| **Q5** | Do roads and canals explain it? *(bonus)* | Point process model | Random, but more fires near roads and canals |
| **Q6** | Do recent fires point to new ones? *(bonus)* | Daily back-test | A hotspot map of the same size |

: {tbl-colwidths="[6,34,26,34]"}

::: notes
A fire map always looks clustered, so every question is judged against what chance would give. Q2 compares with fires placed at random. Q3 allows for the fact that some areas burn more. Q4 keeps every location and shuffles only the dates, so it asks about timing alone. Q5 and Q6 are the bonus analysis.
:::

## Insight 1. Fire came late and fast, 82% of fire-days fell in September

<div class="part">Part 2 · First-order analysis</div>

```{r}
#| fig-width: 13
#| fig-height: 5.4
daily <- fires %>% st_drop_geometry() %>% count(date_local, name = "events") %>%
  complete(date_local = seq(as_date("2026-07-01"), as_date("2026-09-22"), by = "day"),
           fill = list(events = 0)) %>%
  mutate(roll7 = as.numeric(stats::filter(events, rep(1 / 7, 7), sides = 1)))
peak <- slice_max(daily, events, n = 1)

ggplot(daily, aes(date_local, events)) +
  annotate("rect", xmin = as_date("2026-08-31") + 0.5, xmax = as_date("2026-09-22") + 0.5,
           ymin = -Inf, ymax = Inf, fill = "#F3E6D3") +
  annotate("text", x = as_date("2026-09-11"), y = 322, label = "September, 82% of fire-days",
           size = 5.5, fontface = "bold", colour = "#8E3B00") +
  geom_col(fill = "grey70", width = 1) +
  geom_line(aes(y = roll7), colour = col_fire, linewidth = 1.6, na.rm = TRUE) +
  annotate("text", x = peak$date_local - 1, y = peak$events - 2, hjust = 1,
           label = paste0(format(peak$date_local, "%d %b"), ", busiest day, ", peak$events), size = 5) +
  annotate("text", x = as_date("2026-08-19"), y = 60, hjust = 1, size = 5,
           label = "Rise starts\naround 20 Aug") +
  annotate("text", x = as_date("2026-07-20"), y = 40, size = 5, colour = col_fire, fontface = "bold",
           label = "Orange line, 7-day average") +
  annotate("text", x = as_date("2026-07-20"), y = 22, size = 5, colour = "grey40",
           label = "Grey bars, fire-days per day") +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b", expand = expansion(mult = 0.01)) +
  scale_y_continuous(limits = c(0, 335), breaks = c(0, 100, 200, 300)) +
  labs(x = NULL, y = "Fire-days per day") +
  theme_slide + theme(panel.grid.major.x = element_blank())

reburn_share <- fires %>% st_drop_geometry() %>% arrange(cell, date_local) %>%
  group_by(cell) %>% mutate(gap = as.numeric(date_local - lag(date_local))) %>%
  ungroup() %>% summarise(mean(!is.na(gap) & gap <= 7)) %>% pull()
```

<div class="small">The one-day dips, such as 16 and 17 Sep, are most likely cloud or smoke hiding fires. `r round(100 * reburn_share)`% of fire-days were a 1 km square burning again within a week, most often the next day, so crews should go back to fires that look out.</div>

::: notes
Through July and early August there was very little fire. From about 20 August the 7-day average climbs, and it peaks in the final week of the record. September holds 82 percent of all fire-days. The sharp one-day drops are more likely cloud or smoke than fires stopping. Almost half of all fire-days were the same square kilometre burning again within a week.
:::

## Insight 2. Fire spread from the neck to the southern peat

<div class="part">Part 2 · First-order analysis</div>

::::: columns
::: {.column width="66%"}
```{r}
#| fig-width: 10
#| fig-height: 6.6
bw_st <- OS.spattemp(ppp)
st <- spattemp.density(ppp, h = bw_st[1], lambda = bw_st[2])
days <- c(213, 238, 261)
labs_d <- c("1 Aug", "26 Aug", "18 Sep")
st_df <- map2_dfr(days, labs_d, function(d, l) {
  as.data.frame(to_rast(st$z[[as.character(d)]] * npoints(ppp) * 100), xy = TRUE) %>%
    setNames(c("x", "y", "v")) %>% mutate(day = l)
}) %>%
  mutate(day = factor(day, levels = labs_d),
         cls = cut(v, c(-Inf, 0.5, 1, 2.5, 5, 10, Inf),
                   labels = c("under 0.5", "0.5 to 1", "1 to 2.5", "2.5 to 5", "5 to 10", "10 or more")))
tags <- tibble(day = factor(c("1 Aug", "26 Aug", "18 Sep", "18 Sep"), levels = labs_d),
               x = c(851, 826, 835.5, 870) * 1000, y = c(9745, 9800, 9640, 9716) * 1000,
               lab = c("Neck", "Northern\nstrip", "South-central\npeat", "East\nedge"),
               hj = c(0, 1, 0.5, 1))

ggplot(st_df) +
  geom_raster(aes(x, y, fill = cls)) +
  geom_sf(data = pp_sf, fill = NA, colour = "grey30", linewidth = 0.4) +
  geom_label(data = tags, aes(x, y, label = lab, hjust = hj), size = 4.6, fontface = "bold",
             lineheight = 0.85, label.size = 0, fill = alpha("white", 0.8)) +
  scale_fill_manual(values = c("#FEEDDE", "#FDD0A2", "#FDAE6B", "#FD8D3C", "#E6550D", "#A63603"),
                    drop = FALSE, name = "Fire-days per\n100 km² per day") +
  facet_wrap(~ day, nrow = 1) +
  coord_sf(crs = 32749, datum = NA) +
  labs(x = NULL, y = NULL) +
  theme_void(base_size = 17) +
  theme(strip.text = element_text(face = "bold", size = 19, margin = margin(b = 6)),
        legend.position = "bottom", legend.title = element_text(size = 14),
        legend.text = element_text(size = 13),
        plot.background = element_rect(fill = "#FDFAF3", colour = NA)) +
  guides(fill = guide_legend(nrow = 1, title.position = "left"))
```
:::

::: {.column width="34%"}
<div class="fix"><b>July.</b> Almost all burning was just below the narrow neck.</div>

<div class="fix"><b>Late August.</b> New fires started in the north, on the east edge and on the south coast.</div>

<div class="fix"><b>September.</b> The south-central peat burned hardest, with a second hot spot on the east edge.</div>

<div class="small">Spatio-temporal kernel density. The same peak appears with 2 km, 7.5 km and adaptive bandwidths in the report.</div>
:::
:::::

::: notes
This is spatio-temporal kernel density on three dates. In July the only warm spot is just below the narrow neck of the regency. By late August the northern strip and the south warm up. By mid September the south-central peat has by far the highest density, with a second hot spot on the east edge. The report checks this with three different bandwidths and the peak stays in the same place, so this is where crews and water points belong.
:::

## Insight 3. Fires cluster far more than chance, but a map cannot say why

<div class="part">Part 3 · Second-order analysis</div>

::::: columns
::: {.column width="58%"}
```{r}
#| fig-width: 8.4
#| fig-height: 5.6
lc <- bonus$l_compare %>%
  setNames(c("model", "obs", "hi")) %>%
  mutate(model = factor(c("Fires placed at random", "Random, more fires near\nroads and canals",
                          "Random, following the\n7.5 km density map"),
                        levels = rev(c("Fires placed at random", "Random, more fires near\nroads and canals",
                                       "Random, following the\n7.5 km density map"))))

ggplot(lc, aes(y = model)) +
  geom_segment(aes(x = 0, xend = hi, yend = model), linewidth = 9, colour = "grey80") +
  geom_segment(aes(x = hi, xend = obs, yend = model), linewidth = 0.8, colour = "grey45",
               linetype = "dotted") +
  geom_point(aes(x = obs), size = 7, colour = col_fire) +
  geom_text(aes(x = obs, label = sprintf("%.1f", obs)), vjust = -1.3, size = 6, fontface = "bold",
            colour = col_fire) +
  annotate("text", x = 0.02, y = 0.35, hjust = 0, size = 5, colour = "grey35",
           label = "Grey bar = range from chance") +
  annotate("text", x = 6.5, y = 0.35, hjust = 1, size = 5, colour = col_fire, fontface = "bold",
           label = "Dot = the real fires") +
  scale_x_continuous(limits = c(0, 6.6), expand = expansion(mult = c(0, 0.02))) +
  scale_y_discrete(expand = expansion(add = c(1, 0.6))) +
  labs(x = "Extra closeness at 5 km, L(5) minus 5 (km)", y = NULL) +
  theme_slide +
  theme(panel.grid.major.y = element_blank(), axis.text.y = element_text(size = 16, colour = "grey15"))
```
:::

::: {.column width="42%"}
<div class="fix"><b>Far from random.</b> The nearest fire is about half as far away as chance would put it (Clark-Evans ratio 0.48).</div>

<div class="fix"><b>Roads and canals barely help.</b> Allowing for access closes little of the gap.</div>

<div class="fix"><b>The density map decides.</b> With a 5 km map instead of 7.5 km, the gap disappears. One map cannot tell fires affecting each other from land that simply burns easily, so time is needed.</div>
:::
:::::

::: notes
The L function asks whether fires have more neighbours than chance would give. Against fires placed at random the answer is a clear yes. Allowing for roads and canals barely changes that. Allowing for the density map shrinks the gap, and with a 5 kilometre map it disappears. So space alone cannot say whether fires affect each other or whether some land simply burns easily. The next slide adds time.
:::

## Insight 4. Fires under 3 km apart follow each other within 1 to 3 days

<div class="part">Part 3 · Second-order analysis</div>

::::: columns
::: {.column width="58%"}
```{r}
#| fig-width: 7.6
#| fig-height: 5.9
kx2 <- so$kx %>% filter(str_detect(test, "Step 2"))
ggplot(kx2, aes(lag, dist, fill = pmin(pmax(log2(ratio), -2), 2))) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.1f", ratio)), size = 4.6) +
  annotate("rect", xmin = 1.5, xmax = 3.5, ymin = 0.5, ymax = 3.5,
           fill = NA, colour = "black", linewidth = 1.4) +
  scale_fill_gradient2(low = col_accent, mid = "white", high = col_fire, midpoint = 0,
                       limits = c(-2, 2), guide = "none") +
  labs(x = "Days apart", y = "Distance apart (km)",
       caption = "Black box, the main test. Each value is observed pairs ÷ pairs expected when\ndates are shuffled within 10 km blocks. Orange, more than chance. Blue, fewer") +
  coord_cartesian(clip = "off") +
  theme_slide + theme(panel.grid = element_blank(),
                      plot.caption = element_text(size = 12.5, colour = "grey35", hjust = 0),
                      axis.text.x = element_text(angle = 30, hjust = 1))
```
:::

::: {.column width="42%"}
<div class="bignum">1.33×</div>
<div class="bigcap">as many close pairs as chance, p = 0.005, even with each 10 km block keeping its own fire season</div>

```{r}
#| fig-width: 5.6
#| fig-height: 3
chk <- so$sens_tab %>%
  filter(!is.na(`Q4 ratio, under 3 km, 1 to 3 days`)) %>%
  transmute(check = factor(c("MODIS, main", "VIIRS satellite", "New fires only", "Pairs 2 to 3 km only"),
                           levels = rev(c("MODIS, main", "VIIRS satellite", "New fires only", "Pairs 2 to 3 km only"))),
            ratio = `Q4 ratio, under 3 km, 1 to 3 days`)
ggplot(chk, aes(ratio, check)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_segment(aes(x = 1, xend = ratio, yend = check), colour = "grey70", linewidth = 1.2) +
  geom_point(size = 4.5, colour = col_fire) +
  geom_text(aes(label = sprintf("%.2f", ratio)), hjust = -0.45, size = 5) +
  annotate("text", x = 1.01, y = 0.35, hjust = 0, size = 4.2, colour = "grey35", label = "chance = 1") +
  scale_y_discrete(expand = expansion(add = c(0.9, 0.5))) +
  scale_x_continuous(limits = c(0.97, 1.47)) +
  labs(x = NULL, y = NULL, title = "Holds in every check") +
  coord_cartesian(clip = "off") +
  theme_slide + theme(panel.grid = element_blank(), axis.text.x = element_blank(),
                      plot.title = element_text(size = 16, face = "bold", colour = "grey20"),
                      axis.text.y = element_text(size = 14.5, colour = "grey15"))
```
:::
:::::

::: notes
This is the strongest finding. Pairs of fires under 3 kilometres and 1 to 3 days apart are about a third more common than shuffled dates give, even when each 10 kilometre block keeps its own fire season. It holds for a second satellite, for newly burning cells only, and for pairs 2 to 3 kilometres apart. The test cannot tell spread from one person lighting several plots, but both call for fast follow-up nearby.
:::

## Insight 5. Watching near recent fires beat a hotspot map

<div class="part">Part 4 · Bonus</div>

::::: columns
::: {.column width="30%"}
```{r}
#| fig-width: 4.4
#| fig-height: 7.6
tm_shape(pp_sf) +
  tm_polygons(fill = col_land, col = "grey50") +
  tm_shape(bonus$ex_zone) +
  tm_polygons(fill = col_band, fill_alpha = 0.45, col = col_band) +
  tm_shape(bonus$ex_new) +
  tm_dots(fill = "where", size = 0.4,
          fill.scale = tm_scale_categorical(values = c("inside the zone" = col_fire,
                                                       "outside the zone" = col_accent)),
          fill.legend = tm_legend(title = "New fires, 18 Sep",
                                  position = tm_pos_out("center", "bottom"))) +
  tm_layout(bg.color = "#FDFAF3", outer.bg.color = "#FDFAF3", frame = FALSE,
            legend.text.size = 1.2, legend.title.size = 1.3)
```
:::

::: {.column width="70%"}
<div class="statrow">
<div><div class="bignum">`r round(100 * bonus$watch_sum$watch_zone)`%</div><div class="bigcap">of new fires caught by the watch zone</div></div>
<div><div class="bignum" style="color:#6B6B6B">`r round(100 * max(bonus$watch_sum$hotspot_1.5km, bonus$watch_sum$hotspot_3km, bonus$watch_sum$hotspot_6km))`%</div><div class="bigcap">by the best hotspot map of the same size</div></div>
<div><div class="bignum" style="color:#0072B2">`r round(100 * bonus$watch_sum$land_covered)`%</div><div class="bigcap">of the regency covered, on average</div></div>
</div>

```{r}
#| fig-width: 10
#| fig-height: 4.1
ww <- bonus$watch_week %>% filter(week >= as_date("2026-08-01"))
watch_lab <- ww %>% filter(!is.na(share)) %>% group_by(measure) %>%
  slice_max(week, n = 1) %>% ungroup() %>%
  mutate(label = c("Watch zone", "Hotspot map", "Land covered")[as.integer(measure)])

ggplot(ww, aes(week, share, colour = measure, linetype = measure)) +
  geom_line(linewidth = 1.4, na.rm = TRUE) + geom_point(size = 3.2, na.rm = TRUE) +
  geom_text(data = watch_lab, aes(label = label), hjust = 0, nudge_x = 2, size = 5.3, fontface = "bold") +
  scale_colour_manual(values = c(col_fire, "grey40", col_accent), guide = "none") +
  scale_linetype_manual(values = c("solid", "solid", "dashed"), guide = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  scale_x_date(date_labels = "%d %b", expand = expansion(mult = c(0.02, 0.22))) +
  labs(x = "Week starting", y = "New fires caught") +
  theme_slide
```

<div class="small">Watch zone, the land within 3 km of the last three days' fires. Hotspot map, the same amount of land where the most fires had burned so far.</div>
:::
:::::

::: notes
This turns the space-time result into a daily rule. Each morning, draw 3 kilometres around every fire of the past three days, and send patrols there first. Over the season that zone caught 76 percent of newly burning cells while covering about 15 percent of the regency. A hotspot map of the same size, built from where the most fires had burned so far, caught at most 64 percent. It is one season, so it needs testing on another year.
:::

## Key findings and their limits {.smaller}

<div class="part">Part 5 · So what</div>

| | Finding | Evidence | Main limit |
|---|---|---|---|
| **1** | The season came late and fast | 82% of fire-days fell in September | Cloud and smoke hide fires at the peak, and data after 30 June is provisional |
| **2** | The south-central peat burned most | Same peak at 2 km, 7.5 km and adaptive bandwidths | Counts fire-days, not burned area or harm |
| **3** | Fires cluster far beyond random | L function far above the envelope, Clark-Evans ratio 0.48 | After allowing for density, the result changes with bandwidth, so Q3 stays open |
| **4** | Fires close in space are close in time | Pairs under 3 km and 1 to 3 days apart `r so$sens_tab[["Q4 ratio, under 3 km, 1 to 3 days"]][2]` times as common as chance, also for VIIRS | Cannot tell spread from one person lighting several plots |
| **5** | Recent fires find new fires better than a hotspot map | Watch zone `r round(100 * bonus$watch_sum$watch_zone)`% against at most `r round(100 * max(bonus$watch_sum$hotspot_1.5km, bonus$watch_sum$hotspot_3km, bonus$watch_sum$hotspot_6km))`% on the same area | One season, and the 3 km and 3 days came from the same data |
| **6** | Roads and canals explain little of the clustering | Only 4 of 3,065 fire-days over 14 km from a road, but the access model barely lowers L | OSM misses many canals |

: {tbl-colwidths="[4,28,34,34]"}

::: {.callout-note appearance="minimal"}
**Core finding.** Density maps show where to base crews. The space-time result, and the watch zone built on it, show where to look each day.
:::

::: notes
This slide pulls the findings together, each with its evidence and its main limit. The season was late and fast. The south-central peat burned most at every bandwidth. Fires cluster far beyond random, but after allowing for density the answer depends on the bandwidth, so that question stays open. The space-time test is the strongest result, and the watch zone shows it is useful in practice, though only on one season so far. Roads and canals explain little of the clustering.
:::

## What this means for fire management in Pulang Pisau

<div class="part">Part 5 · So what</div>

::::: columns
::: {.column width="34%"}
```{r}
#| fig-width: 4.6
#| fig-height: 7.6
kd <- density(unmark(ppp), sigma = 2)
kd_df <- as.data.frame(to_rast(kd), xy = TRUE) %>% setNames(c("x", "y", "v")) %>% filter(v >= 2)
plan_tags <- tibble(x = c(835.5, 868, 849, 822) * 1000, y = c(9627, 9716, 9745, 9800) * 1000,
                    lab = c("South-central\npeat", "East\nedge", "Neck", "Northern\nstrip"),
                    hj = c(0.5, 1, 0, 1))
ggplot() +
  geom_sf(data = pp_sf, fill = col_land, colour = "grey45") +
  geom_sf(data = fires, colour = "grey75", size = 0.25) +
  geom_raster(data = kd_df, aes(x, y), fill = col_fire, alpha = 0.85) +
  geom_label(data = plan_tags, aes(x, y, label = lab, hjust = hj), size = 4.4, fontface = "bold",
             lineheight = 0.85, label.size = 0, fill = alpha("white", 0.85)) +
  coord_sf(crs = 32749, datum = NA) +
  labs(x = NULL, y = NULL, caption = "Orange, the densest fire-days\n(2 km KDE, 2 or more per km²)") +
  theme_void(base_size = 14) +
  theme(plot.caption = element_text(hjust = 0.5, size = 13, colour = "grey30"))
```
:::

::: {.column width="66%"}
<div class="act"><b>Base crews and water in the south-central peat and east edge</b><br><span class="tag">Because</span> both had the densest burning at every bandwidth, and water shortages already slow firefighting<br><span class="tag">Check first</span> burned area and peat depth</div>

<div class="act"><b>Draw a 3 km watch zone each morning and patrol it first</b><br><span class="tag">Because</span> it caught `r round(100 * bonus$watch_sum$watch_zone)`% of new fires, against `r round(100 * max(bonus$watch_sum$hotspot_1.5km, bonus$watch_sum$hotspot_3km, bonus$watch_sum$hotspot_6km))`% for a hotspot map of the same size<br><span class="tag">Check first</span> a test on another fire season</div>

<div class="act"><b>Go back to put-out fires the next day</b><br><span class="tag">Because</span> `r round(100 * reburn_share)`% of fire-days were a square burning again within a week<br><span class="tag">Check first</span> field records of fires that restarted</div>

<div class="act"><b>Watch the neck early in the season</b><br><span class="tag">Because</span> it burned from July, weeks before the rest of the regency<br><span class="tag">Check first</span> earlier years and rainfall data</div>

<div class="small">Each action rests on evidence from this study but needs the check first. Nothing here shows cause. The watch zone says where the next fire is likely, not whether it spread or was lit.</div>
:::
:::::

::: notes
Four actions come out of the analysis. Base crews and water where burning was densest. Draw a watch zone around recent fires each morning, because it found more new fires than a hotspot map. Go back to fires that look out, because almost half of fire-days were the same square burning again. And watch the neck early. Each one names what should be checked before acting, and none of it claims to show cause.
:::

## References {.smaller}

<div class="small">

- Baddeley, A., Rubak, E. and Turner, R. (2015). *Spatial Point Patterns: Methodology and Applications with R*. Chapman and Hall/CRC.
- Berita Sampit (2026, 18 September). Karhutla Pulpis capai 3.693 hektare, kualitas udara berada pada level berbahaya.
- Davies, T. M., Marshall, J. C. and Hazelton, M. L. (2018). Tutorial on kernel estimation of continuous spatial and spatiotemporal relative risk. *Statistics in Medicine*, 37(7), 1191 to 1221.
- Diggle, P. J. (2013). *Statistical Analysis of Spatial and Spatio-Temporal Point Patterns* (3rd ed.). CRC Press.
- Giglio, L., Schroeder, W. and Justice, C. O. (2016). The collection 6 MODIS active fire detection algorithm and fire products. *Remote Sensing of Environment*, 178, 31 to 41.
- Knox, E. G. (1964). The detection of space-time interactions. *Journal of the Royal Statistical Society, Series C*, 13(1), 25 to 30.
- Page, S. E. and Hooijer, A. (2016). In the line of fire, the peatlands of Southeast Asia. *Philosophical Transactions of the Royal Society B*, 371, 20150176.
- Townsley, M., Homel, R. and Chaseling, J. (2003). Infectious burglaries, a test of the near repeat hypothesis. *British Journal of Criminology*, 43(3), 615 to 633.
- WHO (2021). *WHO Global Air Quality Guidelines*. World Health Organization, Geneva.

Data from NASA FIRMS, GADM 4.1 and OpenStreetMap. Full list, code and workflow in the [Technical Report](Take-home_Ex01.html).

</div>
