---
title: "Where and When Pulang Pisau Burned"
subtitle: "Spatial and spatio-temporal point patterns of the 2026 fire season"
author: "ZeXin Chen"
date: "September 27, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    embed-resources: true
    transition: fade
    show-notes: false
    width: 1600
    height: 900
    footer: "ISSS626 Take-home Exercise 1 | ZeXin Chen | [Technical report](Take-home_Ex01.html) | [Back to site](https://isss-626-gaa-six.vercel.app)"
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.googleapis.com">
        <link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
        <link href="https://fonts.googleapis.com/css2?family=Inter:wght@400;500;600;700&family=Newsreader:opsz,wght@6..72,400;6..72,600&display=swap" rel="stylesheet">
execute:
  echo: false
  eval: true
  warning: false
  message: false
  freeze: true
---

```{r}
#| label: setup
pacman::p_load(sf, spatstat, sparr, terra, tmap, tidyverse, knitr)
knitr::opts_chunk$set(out.width = "100%", dpi = 150)

prep   <- read_rds("data/rds/prepared.rds")
so     <- read_rds("data/rds/results_second_order.rds")
bonus  <- read_rds("data/rds/results_bonus.rds")
kal_rate <- read_rds("data/rds/kal_rate.rds")

pp_sf <- prep$pp_sf
fires <- prep$fires
ppp   <- prep$ppp

col_fire   <- "#D55E00"
col_accent <- "#0072B2"
col_band   <- "#E69F00"
col_land   <- "#F7F4EC"

to_rast <- function(im) {
  r <- rast(im)
  ext(r) <- as.vector(ext(r)) * 1000
  crs(r) <- "EPSG:32749"
  r
}

theme_slide <- theme_minimal(base_size = 18) +
  theme(plot.title = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", size = 15),
        plot.background = element_rect(fill = "#FDFAF3", colour = NA),
        panel.background = element_rect(fill = "#FDFAF3", colour = NA))
```

## Contents

::::: columns
::: {.column width="50%"}
**Part 1, The problem and the data**

- Why Pulang Pisau
- Three fixes and two checks on raw satellite data
- Six questions and their null models

**Part 2, First-order findings**

- When the fires came
- Where they burned
:::

::: {.column width="50%"}
**Part 3, Second-order findings**

- What spatial clustering can and cannot show
- Near-repeat fires in space and time
- A watch zone against a hotspot map

**Part 4, So what**

- Limitations
- Public-safety implications
:::
:::::

::: notes
This deck summarises the technical report. It moves from the problem and the data, to where and when fires burned, to how fires relate to each other, and ends with what the results mean for fire management in Pulang Pisau.
:::

## Pulang Pisau had the densest fire detections of any Kalimantan regency

::::: columns
::: {.column width="58%"}
```{r}
#| fig-width: 9
#| fig-height: 7.4
tm_shape(kal_rate) +
  tm_polygons(fill = "per100",
              fill.scale = tm_scale_intervals(style = "fixed",
                breaks = c(0, 2, 5, 10, 20, 35, 80), values = "brewer.oranges",
                labels = c("0 to 2", "2 to 5", "5 to 10", "10 to 20", "20 to 35", "35 to 80")),
              fill.legend = tm_legend(title = "Detections\nper 100 km²",
                position = tm_pos_out("right", "center")),
              col = "grey55", lwd = 0.4) +
  tm_shape(filter(kal_rate, NAME_2 == "PalangkaRaya")) +
  tm_borders(col = col_accent, lwd = 2.5) +
  tm_shape(filter(kal_rate, NAME_2 == "PulangPisau")) +
  tm_borders(col = "black", lwd = 3) +
  tm_add_legend(type = "lines", labels = c("Pulang Pisau", "Palangka Raya (city)"),
                col = c("black", col_accent), lwd = c(3, 2.5),
                position = tm_pos_out("right", "center")) +
  tm_layout(bg.color = "#FDFAF3", outer.bg.color = "#FDFAF3", frame = FALSE,
            legend.text.size = 1, legend.title.size = 1.2) +
  tm_credits("NASA FIRMS MODIS C6.1, 1 Jan to 22 Sep 2026", position = c("left", "bottom"), size = 0.9)
```
:::

::: {.column width="42%"}
<div class="kicker">The 2026 season</div>

<div class="bignum">3,693.7 ha</div>
<div class="bigcap">burned in 8 subdistricts by 17 Sep</div>

<div class="bignum">595 µg/m³</div>
<div class="bigcap">PM2.5 on 17 Sep, about 40 times the WHO 24-hour guideline</div>

<div class="small">The regent raised the fire status to emergency response. Most of the regency is deep peat inside the failed Mega Rice Project, drained by canals in the 1990s.

**Research question.** Where and when did fires burn in 2026, and were fires closer together in space and time than uneven fuel alone would explain?</div>
:::
:::::

::: notes
Pulang Pisau is on the south coast of Central Kalimantan. It had the most fire detections per square kilometre of any regency in Kalimantan this year. Only Palangka Raya city next door was higher. Local reports put burned area at 3,694 hectares by 17 September and PM2.5 at a hazardous 595. The question is not just whether fires cluster, because fire maps always look clustered. It is whether they cluster beyond what the uneven spread of land that burns easily would give.
:::

## Raw satellite detections needed three fixes and two checks

::::: columns
::: {.column width="48%"}
```{r}
kable(prep$prep_log %>% mutate(rows = format(rows, big.mark = ",")),
      col.names = c("Step", "Rows"), align = c("l", "r"))
```

<div class="small">**Event** = one 1 km cell burning on one local day. **Window** = the GADM regency boundary, 1 Jul to 22 Sep 2026. **CRS** = WGS 84 UTM 49S.</div>
:::

::: {.column width="52%"}
| Problem | Fix |
|---|---|
| FIRMS dates are UTC. 6,026 Aqua night passes land a day early | Convert to local time, UTC+7 |
| One fire is seen by up to four passes a day | Merge to one event per 1 km cell per day |
| Low confidence detections include false alarms | Drop confidence under 30 |
| Data after 30 June is provisional NRT with no `type` flag | Checked, no cell burned on more than 14 of 84 days, so no sign of a static source |
| About 15% of pixels are wider than 2 km, so one fire can light two cells | Same-day pairs left out of the space-time test |

<div class="small">MODIS is the main sensor because VIIRS has gaps from late April to early June and in mid July. VIIRS is rerun as a check.</div>
:::
:::::

::: notes
The raw file had no missing or impossible coordinates. The real problems were less obvious. FIRMS uses UTC dates, so every Aqua night pass sat a day early. The same fire is seen up to four times a day. Low confidence points include false alarms. And data after 30 June is provisional. After fixing these, 3,065 fire events remain in the 84-day window.
:::

## Six questions, each tested against a stated null model

| # | Question | Method | Null model |
|--|------------|------------|------------|
| Q1 | Where and when was intensity high? | KDE, adaptive KDE, spatio-temporal KDE | none, descriptive |
| Q2 | More clustered than random? | L function, 99-sim global envelope | CSR |
| Q3 | Clustered beyond uneven intensity? | Inhomogeneous L function at 5, 7.5 and 10 km | Inhomogeneous Poisson, KDE intensity |
| Q4 | Close in space also close in time? | Knox style test, 199 shuffles | Days shuffled, then shuffled within 10 km blocks |
| Q5 | Does access explain where fires burn? | `rhohat()`, Poisson point process model | Inhomogeneous Poisson, road, canal and river intensity |
| Q6 | Do recent fires beat a hotspot map? | Daily back-test, 4 Jul to 22 Sep | Hotspot zone of the same size from all earlier fires |

<div class="small">Q5 and Q6 are the bonus analysis. Q6 tests whether the Q4 result is useful beyond a plain hotspot map.</div>

::: notes
Every second-order claim in this study is made against a named null model. Q2 uses complete spatial randomness. Q3 uses an inhomogeneous Poisson process, so clustering that comes only from uneven intensity is removed. Q4 shuffles the days, first across the regency and then only within 10 km blocks, so regional differences in timing cannot fake a result. Q5 and Q6 are the bonus parts.
:::

## Fire came late and fast, 82% of fire-days fell in September

```{r}
#| fig-width: 12
#| fig-height: 5
daily <- fires %>% st_drop_geometry() %>% count(date_local, name = "events") %>%
  complete(date_local = seq(as_date("2026-07-01"), as_date("2026-09-22"), by = "day"),
           fill = list(events = 0)) %>%
  mutate(roll7 = as.numeric(stats::filter(events, rep(1 / 7, 7), sides = 1)))

ggplot(daily, aes(date_local, events)) +
  geom_col(fill = "grey78", width = 1) +
  geom_line(aes(y = roll7), colour = col_fire, linewidth = 1.4, na.rm = TRUE) +
  annotate("text", x = as_date("2026-07-20"), y = 200, hjust = 0, size = 6,
           label = "Grey bars are fire-days per day.\nThe line is the 7-day average.") +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = NULL, y = "Fire-days per day") +
  theme_slide + theme(panel.grid.major.x = element_blank())

reburn_share <- fires %>% st_drop_geometry() %>% arrange(cell, date_local) %>%
  group_by(cell) %>% mutate(gap = as.numeric(date_local - lag(date_local))) %>%
  ungroup() %>% summarise(mean(!is.na(gap) & gap <= 7)) %>% pull()
```

<div class="small">The 7-day average starts rising around 20 August and peaks in the last week of the record. The one-day drops are most likely cloud or smoke hiding fires. `r round(100 * reburn_share)`% of fire-days were a 1 km cell burning again within a week, most often the next day.</div>

::: notes
Through July and early August there was very little fire. From about 20 August the 7-day average climbs, and it peaks in the final week of the record. September holds 82 percent of all fire-days. The sharp one-day drops are more likely cloud or smoke than fires stopping, because satellites cannot see through them. Almost half of all fire-days were the same square kilometre burning again within a week, which is why crews need to go back to fires that look out.
:::

## Burning moved from the narrow neck in July to the southern peat in September

```{r}
#| fig-width: 13
#| fig-height: 7.2
bw_st <- OS.spattemp(ppp)
st <- spattemp.density(ppp, h = bw_st[1], lambda = bw_st[2])
days <- c(196, 213, 227, 238, 254, 261)
st_stack <- do.call(c, lapply(days, function(d) to_rast(st$z[[as.character(d)]] * npoints(ppp))))
names(st_stack) <- format(as_date(days - 1, origin = "2026-01-01"), "%d %b")

tm_shape(st_stack) +
  tm_raster(col.scale = tm_scale_intervals(style = "fixed",
              breaks = c(0, 0.005, 0.01, 0.025, 0.05, 0.1, 0.2),
              labels = c("under 0.005", "0.005 to 0.01", "0.01 to 0.025",
                         "0.025 to 0.05", "0.05 to 0.1", "0.1 or more"),
              values = "brewer.oranges"),
            col.legend = tm_legend(title = "Fire-days per km²\nper day",
              position = tm_pos_out("right", "center")),
            col.free = FALSE) +
  tm_shape(pp_sf) + tm_borders(col = "grey30") +
  tm_facets(ncol = 6) +
  tm_layout(panel.label.size = 1.3, panel.label.bg.color = "white",
            bg.color = "#FDFAF3", outer.bg.color = "#FDFAF3",
            legend.text.size = 1.3, legend.title.size = 1.4,
            outer.margins = 0, frame = FALSE)
```

<div class="small">Oversmoothing bandwidths `r round(bw_st[1], 1)` km and `r round(bw_st[2], 1)` days. The report finds the same peak at 2 km, 7.5 km and adaptive bandwidths.</div>

::: notes
This is spatio-temporal kernel density at six dates. In July the main warm spot is the narrow neck of the regency. By late August the northern strip and the south warm up. By mid September the south-central peat has the highest density, with a second hot spot on the east edge. The report checks this with three different spatial bandwidths and the peak stays in the same place.
:::

## Fires cluster in space, but a map alone cannot say why

::::: columns
::: {.column width="62%"}
```{r}
#| fig-width: 7.5
#| fig-height: 5
ggplot(so$env_df, aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = col_band, alpha = 0.4) +
  geom_line(aes(y = ref), colour = col_accent, linetype = "dashed", linewidth = 0.9) +
  geom_line(aes(y = obs), colour = "black", linewidth = 1.2) +
  facet_wrap(~ test, scales = "free_y",
             labeller = labeller(test = function(x) str_wrap(x, 28))) +
  labs(x = "Distance r (km)", y = "L(r) minus r (km)") +
  theme_slide
```
:::

::: {.column width="38%"}
```{r}
bonus$l_compare %>%
  rename(`Null model` = 1, `Observed at 5 km` = 2, `Envelope top` = 3) %>%
  kable()
```

<div class="small">Black is the observed curve, orange the 95% global envelope. Fires are far more clustered than random, and roads, canals and rivers explain little of this.

After allowing for density, the answer depends on the bandwidth. The curve is inside the envelope at 5 km and above it at 7.5 and 10 km. One map cannot tell fires affecting each other apart from land that burns easily, so the space-time test gives the stronger evidence.</div>
:::
:::::

::: notes
Against complete spatial randomness the fires are hugely clustered, but that mostly repeats what the density map shows. Against a KDE intensity, the answer depends on the bandwidth. With 7.5 or 10 km there is extra clustering, with 5 km there is not. Roads, canals and rivers barely reduce the clustering. One map cannot tell fires affecting each other apart from land that burns easily, or from the same land burning again, so the space-time test gives the stronger evidence.
:::

## Fire pairs under 3 km and 1 to 3 days apart are `r so$sens_tab[["Q4 ratio, under 3 km, 1 to 3 days"]][2]` times as common as with random dates

::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 7.5
#| fig-height: 5.2
so$kx %>% filter(str_detect(test, "Step 2")) %>%
  ggplot(aes(lag, dist, fill = pmin(pmax(log2(ratio), -2), 2))) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = paste0(sprintf("%.1f", ratio), sig)), size = 4.5) +
  scale_fill_gradient2(low = col_accent, mid = "white", high = col_fire, midpoint = 0,
                       limits = c(-2, 2), breaks = -2:2,
                       labels = c("1/4", "1/2", "1", "2", "4"),
                       name = "Observed /\nexpected") +
  labs(x = "Days apart", y = "Distance apart (km)") +
  theme_slide + theme(panel.grid = element_blank(),
                      axis.text.x = element_text(angle = 35, hjust = 1))
```
:::

::: {.column width="40%"}
<div class="small">Dates are shuffled only within 10 km blocks, so every location and each block's fire season stay the same. * marks cells that beat 95% of 199 shuffles. Same-day pairs are left out of the pooled test, because one wide pixel can split a fire in two.</div>

```{r}
so$sens_tab %>%
  filter(!is.na(`Q4 ratio, under 3 km, 1 to 3 days`)) %>%
  transmute(Data = c("MODIS", "VIIRS", "Newly burning cells", "Pairs 2 to 3 km only"), Events = events,
            Ratio = `Q4 ratio, under 3 km, 1 to 3 days`, p = `Q4 p-value`) %>%
  kable()
```

<div class="small">The result holds for a second satellite, for newly burning cells, and for pairs 2 to 3 km apart. It could still be one fire front moving across part of a block.</div>
:::
:::::

::: notes
This is the strongest finding. Pairs of fires under 3 kilometres and 1 to 3 days apart are about 1.33 times as common as shuffled dates give, even when each 10 kilometre block keeps its own fire season. It holds for a second sensor, for newly burning cells and for pairs 2 to 3 kilometres apart. The test cannot tell spread from one person lighting several plots, or from a fire front crossing part of a block, but all of these call for fast follow-up nearby.
:::

## A 3-day watch zone beat a same-size hotspot map, `r round(100 * bonus$watch_sum$watch_zone)`% to `r round(100 * max(bonus$watch_sum$hotspot_1.5km, bonus$watch_sum$hotspot_3km, bonus$watch_sum$hotspot_6km))`%

::::: columns
::: {.column width="36%"}
```{r}
#| fig-width: 5.5
#| fig-height: 7.5
tm_shape(pp_sf) +
  tm_polygons(fill = col_land, col = "grey50") +
  tm_shape(bonus$ex_zone) +
  tm_polygons(fill = col_band, fill_alpha = 0.45, col = col_band) +
  tm_shape(bonus$ex_new) +
  tm_dots(fill = "where", size = 0.35,
          fill.scale = tm_scale_categorical(values = c("inside the zone" = col_fire,
                                                       "outside the zone" = col_accent)),
          fill.legend = tm_legend(title = "Newly burning, 18 Sep",
                                  position = tm_pos_out("center", "bottom"))) +
  tm_layout(bg.color = "#FDFAF3", outer.bg.color = "#FDFAF3", frame = FALSE,
            legend.text.size = 1.1, legend.title.size = 1.2)
```
:::

::: {.column width="64%"}
```{r}
#| fig-width: 8
#| fig-height: 5
watch_lab <- bonus$watch_week %>% filter(!is.na(share)) %>% group_by(measure) %>%
  slice_max(week, n = 1) %>% ungroup() %>%
  mutate(label = c("Watch zone", "Hotspot zone,\nsame area", "Share of land\nin the zone")[as.integer(measure)])

ggplot(bonus$watch_week, aes(week, share, colour = measure, linetype = measure)) +
  geom_line(linewidth = 1.3, na.rm = TRUE) + geom_point(size = 3, na.rm = TRUE) +
  geom_text(data = watch_lab, aes(label = label), hjust = 0, nudge_x = 2.5,
            size = 5, lineheight = 0.9, fontface = "bold") +
  scale_colour_manual(values = c(col_fire, "grey35", col_accent), guide = "none") +
  scale_linetype_manual(values = c("solid", "solid", "dashed"), guide = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_x_date(date_labels = "%d %b", breaks = as_date(c("2026-07-01", "2026-08-01", "2026-09-01")),
               expand = expansion(mult = c(0.03, 0.32))) +
  labs(x = "Week starting", y = "Newly burning cells caught") +
  theme_slide
```

<div class="small">Watch zone = the area within 3 km of the past three days' fires. Hotspot zone = the same amount of land on the densest part of a KDE of all earlier fires. Lines show the share of newly burning cells caught, for weeks with at least 15.</div>
:::
:::::

::: notes
This tests whether the space-time result is actually useful. Each morning, draw 3 kilometres around every fire of the past three days. A crew could instead watch the places that have burned most so far, using a hotspot map of the same size. Over the season the watch zone caught clearly more newly burning cells than the hotspot map, at every bandwidth tried. The left map shows 18 September. It is one season, so it needs testing on another year.
:::

## Limitations, and what each one does to the results

| Limitation | Likely effect |
|---|---|
| Satellites miss fires under cloud and smoke, worst at the peak | September counts are low, daily timing is noisy |
| Data after 30 June is provisional NRT | Counts may shift when standard processing replaces it |
| Spatial tests cannot tell fires affecting each other apart from land that burns easily | Q3 is left open, and the space-time test gives the main evidence |
| Wide off-track pixels and the fixed 1 km grid | One fire can become two events, same-day pairs are excluded for this reason |
| One season, one regency, watch zone tested on its own data | The 3 km and 3 day rule needs a second year before it is fixed |
| OSM misses many canals, especially in the north | The canal signal is weak and should not rank canal blocking |

::: notes
The main limits come from the satellites and from using one season. Cloud and smoke hide fires. Post-June data is provisional. The spatial clustering test cannot tell fires affecting each other apart from land that burns easily, which is why the space-time test matters. And the watch zone needs testing on another year.
:::

## What this means for fire management in Pulang Pisau {.smaller}

| | Evidence | Possible action | Needed first |
|---|---|---|---|
| **Where to base crews** | South-central peat held peak intensity at every bandwidth, the east edge was a second belt | Standing crews and water points there | Burned area, peat depth, water access |
| **When to scale up** | The neck burned from July, the rest from about 20 August | Check whether the neck burns first in other years | Earlier seasons, rainfall data |
| **Where to send patrols** | The 3-day watch zone beat a same-size hotspot map, on this season's data | Draw the zone each morning, send patrols there first | A second season, crew travel times |
| **Where patrols can reach** | Only 4 of 3,065 fire-days lay over 14 km from a road or track | Plan routes along roads and tracks | Concession and land-use maps |
| **Finishing the job** | Almost half of fire-days were a cell burning again within a week | Go back to put-out fires the next day | Ground records of rekindling |

<div class="small">Nothing here shows cause. The watch zone says where the next fire is likely, not whether it spread or was lit. Full workflow, code and references are in the [technical report](Take-home_Ex01.html).</div>

::: notes
The first-order maps say where to base crews. The space-time result says where to send them next, and it beats a plain hotspot map. Access corridors show where patrols can reach almost all fires. Each row separates the evidence from the recommendation and from what needs checking first. None of this shows cause.
:::
